### 2022-10-20 notebooks/expected_and_actual_counts_ratio.Rmd
Create notebook for observed vs expected analyses

---
title: "Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades"
output: html_document
date: "2022-10-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("ggplot2")
#install.packages("stringr")
library(ggplot2)
library(stringr)
```

## SARS-CoV-2 Mutation Data

First, we need to read in the data.

```{r read_data}
SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE)
```

Look at the data table.

```{r head_data}
head(SARS_mutation, n=5)
```

Describe the entire dataset. How many clades? How many subsets? How many nucleotide sites? How many mutations to exclude?

```{r data_breakdown}

```

After excluding and filtering by subset. Describe the data again? 
Distribution of each mutation (e.g., A->T) for each clade (clade along x-axis, count along y, stacked bar graph with 12 types)

```{r visualize_data}
#Include data from all countries AND exclude branches with very high mutation rates and questionable mutations.
filtered_SARS_mutation <- subset(SARS_mutation, subset=="all" & exclude=='False')

#Add column with the mutation type (clade founder.mutation).
filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1),".",str_sub(filtered_SARS_mutation$nt_mutation,-1,-1),sep="")

#For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
```

Insert information about the plot. Plot to visualize data.
```{r normalized_counts, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+ #stacked bar graph for the mutation counts normalized by the expected mutation rate
  geom_bar(position="stack", stat="identity")+
  theme_bw()+
  labs(x="Phylogenetic Clade",y="Normalized Mutation Count")
```

Insert information about the plot. Plot to visualize data.
```{r expected_counts, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(color=clade, y=actual_count/expected_count, x=nt_site))+ #stacked bar graph for the expected mutation counts
  geom_point()+
  theme_bw()+
  labs(x="",y="")
```
Insert information about the plot. Plot to visualize data.
```{r insert, echo=FALSE}
ggplot(filtered_SARS_mutation, aes(fill=log(actual_count/expected_count), y=nt_site, x=clade))+
  geom_tile()+
  theme_bw()+
  labs(x="",y="")
```

## Contingency Table Analyses

```{r contingency}

```

## False Discovery Rate Correction

```{r false_discovery}

```

## Volcano Plots

```{r pressure, echo=FALSE}

```

### 2022-10-21 notebooks/expected_and_actual_counts_ratio.Rmd
Add commands to compute p-values using Fisher's Exact Test.

@@
 #install.packages("stringr")
+#install.packages("gtools")
+
 library(ggplot2)
 library(stringr)
+library(gtools)
 ```
+
+Discuss purpose of notebook and steps to achieve that goal. Workflow to produce a volcano plot between any two clades.
+List steps to do analysis (e.g., how to structure the data).
+Do some sanity checks.
 
@@
 ```{r read_data}
-SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE)
+SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE) # Read data from 'kfeldmann' computational notebook on server.
+SARS_mutation <- read.csv("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv") # Read data from github.com.
 ```
@@
 #Add column with the mutation type (clade founder.mutation).
-filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1),".",str_sub(filtered_SARS_mutation$nt_mutation,-1,-1),sep="")
+filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 
 #For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
-filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
+#filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
+#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
+#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
+#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
+#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
 ```
@@
 ```{r contingency}
+# CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
+temp <- filtered_SARS_mutation
+filtered_SARS_mutation <- temp[which(temp$clade=="20A" | temp$clade=="20B"),]
 
+# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
+fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count)
+
+# Identify unique clades in the dataset.
+clades <- unique(filtered_SARS_mutation$clade)
+
+# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
+clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
+
+# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
+nt_mutation_list <- unique(paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, sep="_"))
+
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
+rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))
+
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
+rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
+
+# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
+comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
+
+# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
+filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
+
+# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
+p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))
+
+save(p_values,file="fisher_pvalues.Rda")
+load("fisher_pvalues.Rda")
+
+# Combine p-value list with the reference list of clade comparisons.
+
+# run fishers exact an save p-value to dataframe (clade comparison, site, mutation, p-value, fdr p-value, bonferroni p-value)
+
+# test run volcano plot for each clade comparison
+
+# sanity check analyses (distribution of synonymous p-values)
 ```

### 2022-10-22 notebooks/expected_and_actual_counts_ratio.Rmd
Calculate fold change and create volcano plot

@@
 
+#Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count+0.5)/(filtered_SARS_mutation$expected_count+0.5)
+
 #For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
@@
 ```{r contingency}
-# CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
-temp <- filtered_SARS_mutation
-filtered_SARS_mutation <- temp[which(temp$clade=="20A" | temp$clade=="20B"),]
+#CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
+all_data <- filtered_SARS_mutation
+filtered_SARS_mutation <- all_data[which(all_data$clade=="20A" | all_data$clade=="20B"),]
 
-# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
-fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count)
+#Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
+fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count, observed_expected=filtered_SARS_mutation$observed_expected)
 
-# Identify unique clades in the dataset.
+#Identify unique clades in the dataset.
 clades <- unique(filtered_SARS_mutation$clade)
 
-# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
+#Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
 clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
 
-# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
+#Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
 nt_mutation_list <- unique(paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, sep="_"))
 
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
+#In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
 rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))
 
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
+#In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
 rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
 
-# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
+#Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
 comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
 
-# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
+#Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
 filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
 
-# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
+#Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
 p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))
@@
 
-# Combine p-value list with the reference list of clade comparisons.
+#Combine p-value list with the clade comparison list.
+comparison_p_values <- cbind(filtered_comparisons, p_values)
 
-# run fishers exact an save p-value to dataframe (clade comparison, site, mutation, p-value, fdr p-value, bonferroni p-value)
+#False Discovery Rate and Bonferroni Corrections
 
-# test run volcano plot for each clade comparison
+#Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
+comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]),"observed_expected"],fisher_data[which(fisher_data$nt_mutation_clade==c[2]),"observed_expected"])})
+
+#Volcano Plot
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+  geom_point()
 

### 2022-10-24 notebooks/expected_and_actual_counts_ratio.Rmd
Break up code chunks and add descriptions

--- 2022-10-20 notebooks/expected_and_actual_counts_ratio.Rmd
@@
 
+#install.packages("data.table")
 #install.packages("ggplot2")
 #install.packages("stringr")
+#install.packages("gtools")
+
+library(data.table)
 library(ggplot2)
 library(stringr)
+library(gtools)
 ```
+
+Discuss purpose of notebook and steps to achieve that goal. Workflow to produce a volcano plot between any two clades.
+List steps to do analysis (e.g., how to structure the data).
+Do some sanity checks.
 
@@
 
-First, we need to read in the data.
+First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using either access to the server or from github.com.
 
 ```{r read_data}
-SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE)
+# Read data using access to the Fred Hutch server.
+SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
+
+# Read data from github.com.
+#SARS_mutation <- fread("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
 ```
 
-Look at the data table.
+Next, take a look at the first 5 rows of the data.
 
@@
 
-```{r data_breakdown}
+```{r describe_data}
 
@@
 
-After excluding and filtering by subset. Describe the data again? 
-Distribution of each mutation (e.g., A->T) for each clade (clade along x-axis, count along y, stacked bar graph with 12 types)
+## Filter and Prepare the Data
 
-```{r visualize_data}
-#Include data from all countries AND exclude branches with very high mutation rates and questionable mutations.
-filtered_SARS_mutation <- subset(SARS_mutation, subset=="all" & exclude=='False')
+For the analyses conducted in this notebook, subset the data by "all" to include data collected from all countries.
+Additionally, remove low-quality mutations or branches with abnormally high mutation rates.
+Finally, create a column with mutation type (there are 12 possible mutations).
 
-#Add column with the mutation type (clade founder.mutation).
-filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1),".",str_sub(filtered_SARS_mutation$nt_mutation,-1,-1),sep="")
+```{r filter_all_data}
+# Include data from all countries AND exclude branches with very high mutation rates and questionable mutations.
+filtered_SARS_mutation <- subset(SARS_mutation, subset=="all" & exclude=="FALSE")
 
-#For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
-filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
-filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
+# Add column with the mutation type (clade founder.mutation).
+filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 ```
 
+To compare mutations between phylogenetic clades, this notebook will plot volcano plots with p-value on the Y-axis and fold-change on the X-axis.
+Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated.
+
+```{r pseudocount}
+# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count+0.5)/(filtered_SARS_mutation$expected_count+0.5)
+```
+
+## Visualize Filtered Data
+
 Insert information about the plot. Plot to visualize data.
-```{r normalized_counts, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+ #stacked bar graph for the mutation counts normalized by the expected mutation rate
+
+```{r stacked_bar, echo=FALSE}
+# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
+ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
   geom_bar(position="stack", stat="identity")+
@@
 Insert information about the plot. Plot to visualize data.
-```{r expected_counts, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(color=clade, y=actual_count/expected_count, x=nt_site))+ #stacked bar graph for the expected mutation counts
+
+```{r nucleotide_site, echo=FALSE}
+ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=actual_count/expected_count, x=nt_site))+
   geom_point()+
@@
 ```
+
 Insert information about the plot. Plot to visualize data.
-```{r insert, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(fill=log(actual_count/expected_count), y=nt_site, x=clade))+
-  geom_tile()+
-  theme_bw()+
-  labs(x="",y="")
+
+```{r INSERT, echo=FALSE}
+#ggplot(filtered_SARS_mutation, aes(fill=INSERT, y=INSERT, x=INSERT))+
+  #geom_tile()+
+  #theme_bw()+
+  #labs(x="",y="")
 ```
@@
 
-```{r contingency}
+With such a large data set, it will take potentially hours to calculate the p-values for clade comparisons.
+To decrease the amount of time it takes to conduct analyses, create a subset of the data with only the clades to be compared.
+
+```{r subset_by_clade}
+# Create a subset of the data using only the clades that are going to be compared.
+clade_SARS_mutation <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade=="20A" | filtered_SARS_mutation$clade=="20B"),]
+
+# Identify unique clades in the dataset.
+clades <- unique(clade_SARS_mutation$clade)
+```
+
+Combine nucleotide site, mutation type and clade to create a reference column in the data set.
+This column will be referenced to obtain observed and expected counts for clade comparisons.
+Additionally, remove columns from the data set not necessary for contingency table analyses or creating volcano plots.
+
+```{r filter_contingency_data}
+# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
+fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$expected_count, observed_expected=clade_SARS_mutation$observed_expected)
+```
+
+Create a two-column matrix of all possible clade comparisons, each column listing one clade in the comparison. This matrix should have [n x (n-1)]/2 rows where n is the number of clades.
+
+```{r clade_comparisons}
+# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
+clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
+```
+
+For each mutation type (e.g., A -> T) at each nucleotide site, this notebook will generate contingency table analyses for each comparison in the set of clade comparisons.
+To expand the clade comparison matrix to include mutation types at nucleotide sites, generate a unique list of each mutation type (i.e., three mutations for one clade founder) at each nucleotide site.
+
+```{r unique_mutations}
+# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
+nt_mutation_list <- unique(paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, sep="_"))
+```
+
+Expand both the clade comparison matrix and the mutation list to create a combined matrix of all possible clade comparisons for each mutation type at each nucleotide site.
+To do this, repeat the clade comparison matrix by the number of unique mutations, and repeat each mutation in the mutation list by the number of clade comparisons.
+
+```{r combined_clade_comparisons}
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
+rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))
+
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
+rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
+
+# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
+comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
+```
+
+Not every mutation will exist in each clade. Therefore, remove comparisons where observed and expected count data does not exist for one or both clades in the comparison.
+
+```{r filter_for_comparisons}
+# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
+filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
+```
+
+For each comparison, calculate the probability of obtaining the observed and expected counts assuming there is no difference in selection on mutations between the clades being compared (i.e., p-value) using the Fisher's Exact test.
+
+```{r contingency_analysis}
+# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
+'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))'
+
+# Save and load variable because calculating p-values takes a long time.
+'save(p_values,file="fisher_pvalues.Rda")'
+load("fisher_pvalues.Rda")
+
+# Combine p-value list with the clade comparison list.
+comparison_p_values <- cbind(filtered_comparisons, p_values)
+```
+
+## P-Value Correction
+
+```{r p_value_correction}
+# False Discovery Rate and Bonferroni Corrections
 
@@
 
-## False Discovery Rate Correction
+## Volcano Plot
 
-```{r false_discovery}
+Use a function to calculate fold-change using the observed-expected ratios calculated with a pseudocount.
+This value will be the X-axis for the volcano plot.
+P-values clustering less than zero indicate a difference in degree of selection for one clade, whereas p-values clustering greater than zero indicate a similar difference in degree of selection but for the other clade.
+
+```{r fold-change}
+# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
+comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]),"observed_expected"],fisher_data[which(fisher_data$nt_mutation_clade==c[2]),"observed_expected"])})
+```
+
+Plot the volcano plot with -log10 p-values on the Y-axis and log fold-change on the X-axis.
+
+```{r volcano_plot}
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+  geom_point()
+```
+
+## Verify Analyses Using Analyses With Expected Outcomes
+
+```{r verify_analyses}
+# sanity check analyses (distribution of synonymous p-values)
 
 ```
-
-## Volcano Plots
-
-```{r pressure, echo=FALSE}
-
-```

### 2022-10-24 notebooks/expected_and_actual_counts_ratio.Rmd
Add code to describe data set

@@
 
-Discuss purpose of notebook and steps to achieve that goal. Workflow to produce a volcano plot between any two clades.
-List steps to do analysis (e.g., how to structure the data).
-Do some sanity checks.
+[INSERT] Discuss purpose of notebook and steps to achieve that goal.
 
@@
 
-Describe the entire dataset. How many clades? How many subsets? How many nucleotide sites? How many mutations to exclude?
+Describe the entire data set.
 
 ```{r describe_data}
+# How many rows? How many columns?
+dim(SARS_mutation)
 
+# How many phylogenetic clades?
+length(unique(SARS_mutation$clade))
+
+# How many rows for each subset? Data from all countries, only England or only the United States.
+data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA"),]))
+
+# How many nucleotide sites?
+length(unique(SARS_mutation$nt_site))
+
+# How many mutations to exclude for each subset?
+data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all" & SARS_mutation$exclude=="TRUE"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England" & SARS_mutation$exclude=="TRUE"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA" & SARS_mutation$exclude=="TRUE"),]))
 ```
@@
 # Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
-filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count+0.5)/(filtered_SARS_mutation$expected_count+0.5)
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + 0.5)/(filtered_SARS_mutation$expected_count + 0.5)
 ```
@@
 
-Insert information about the plot. Plot to visualize data.
+[INSERT] information about the plot. Plot to visualize data.
 
@@
   theme_bw()+
-  labs(x="Phylogenetic Clade",y="Normalized Mutation Count")
+  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
 ```
 
-Insert information about the plot. Plot to visualize data.
+[INSERT] information about the plot. Plot to visualize data.
 
@@
   theme_bw()+
-  labs(x="",y="")
+  labs(x="", y="")
 ```
 
-Insert information about the plot. Plot to visualize data.
+[INSERT] information about the plot. Plot to visualize data.
 
@@
   #theme_bw()+
-  #labs(x="",y="")
+  #labs(x="", y="")
 ```
@@
 # In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
-rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
+rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
 
@@
 # Save and load variable because calculating p-values takes a long time.
-'save(p_values,file="fisher_pvalues.Rda")'
+'save(p_values, file="fisher_pvalues.Rda")'
 load("fisher_pvalues.Rda")
@@
 # Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
-comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]),"observed_expected"],fisher_data[which(fisher_data$nt_mutation_clade==c[2]),"observed_expected"])})
+comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]), "observed_expected"], fisher_data[which(fisher_data$nt_mutation_clade==c[2]), "observed_expected"])})
 ```

### 2022-10-25 notebooks/expected_and_actual_counts_ratio.Rmd
Reorder code to characterize and filter data

@@
 output: html_document
-date: "2022-10-17"
+date: Sys.time()
 ---
@@
 
-## SARS-CoV-2 Mutation Data
-
 First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using either access to the server or from github.com.
@@
 
-Next, take a look at the first 5 rows of the data.
+## Define Filtering and Analysis Parameters
 
-```{r head_data}
-head(SARS_mutation, n=5)
+```{r parameters}
+# The subset column describes what countries the sequence data came from. Identify what the subset options are.
+unique(SARS_mutation$subset)
+
+# Define what subset of the data to analyze.
+country <- "all"
+
+# Print the clades that are in the data set (i.e., what clades are possible to include in comparisons).
+unique(SARS_mutation$clade)
+
+# Create a list of the clades to include in comparisons.
+clades <- c("20A","20B")
+
+# Define a minimum threshold for the number of expected counts. All mutations below this threshold will be removed.
+min_expected <- 10
+
+# To avoid infinities when calculating fold-change, add a pseudocount to the observed and expected counts.
+pseudo <- 0.5
 ```
 
-Describe the entire data set.
-
-```{r describe_data}
-# How many rows? How many columns?
-dim(SARS_mutation)
-
-# How many phylogenetic clades?
-length(unique(SARS_mutation$clade))
-
-# How many rows for each subset? Data from all countries, only England or only the United States.
-data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA"),]))
-
-# How many nucleotide sites?
-length(unique(SARS_mutation$nt_site))
-
-# How many mutations to exclude for each subset?
-data.frame(all=nrow(SARS_mutation[which(SARS_mutation$subset=="all" & SARS_mutation$exclude=="TRUE"),]), England=nrow(SARS_mutation[which(SARS_mutation$subset=="England" & SARS_mutation$exclude=="TRUE"),]), USA=nrow(SARS_mutation[which(SARS_mutation$subset=="USA" & SARS_mutation$exclude=="TRUE"),]))
-```
-
-## Filter and Prepare the Data
+## Filter and Prepare the SARS-CoV-2 Mutation Data
 
@@
 Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated.
+Mutations with a small count (i.e., 1) are going to be affected by the pseudocount more than mutations with a large count (i.e., 100).
 
@@
 # Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
-filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + 0.5)/(filtered_SARS_mutation$expected_count + 0.5)
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
 ```
 
-## Visualize Filtered Data
+## Characterize the Filtered Data
+
+Take a look at the first 5 rows of the data.
+
+```{r head_data}
+head(filtered_SARS_mutation, n=5)
+```
+
+Describe the entire data set.
+
+```{r describe_data}
+# How many rows? How many columns?
+dim(filtered_SARS_mutation)
+
+# How many phylogenetic clades?
+length(unique(filtered_SARS_mutation$clade))
+
+# How many nucleotide sites?
+length(unique(filtered_SARS_mutation$nt_site))
+```
+
+## Visualize the Filtered Data
 
@@
 # Create a subset of the data using only the clades that are going to be compared.
-clade_SARS_mutation <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade=="20A" | filtered_SARS_mutation$clade=="20B"),]
-
-# Identify unique clades in the dataset.
-clades <- unique(clade_SARS_mutation$clade)
+clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
 ```

### 2022-10-26 notebooks/expected_and_actual_counts_ratio.Rmd
Add an additional method for calculating p-values to compare speed

@@
 output: html_document
-date: Sys.time()
+date: "`r format(Sys.Date(), '%d.%m.%Y')`"
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(warning=FALSE, message=FALSE)
 
@@
 
-# Define a minimum threshold for the number of expected counts. All mutations below this threshold will be removed.
+# Define a minimum threshold for the number of expected counts. Mutations below this threshold will be removed.
 min_expected <- 10
@@
 
-For the analyses conducted in this notebook, subset the data by "all" to include data collected from all countries.
+For the analyses conducted in this notebook, subset the data by the country(countries) the sequence data came from.
 Additionally, remove low-quality mutations or branches with abnormally high mutation rates.
@@
 ```{r filter_all_data}
-# Include data from all countries AND exclude branches with very high mutation rates and questionable mutations.
-filtered_SARS_mutation <- subset(SARS_mutation, subset=="all" & exclude=="FALSE")
+# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
+filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
 
@@
 ```{r pseudocount}
-# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with pseudocount of 0.5.
+# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with a pseudocount.
 filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
@@
 
-Describe the entire data set.
+Describe the filtered data set.
 
@@
 ```{r nucleotide_site, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=actual_count/expected_count, x=nt_site))+
+ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
   geom_point()+
@@
 
+```{r temp}
+filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
+filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
+filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
+filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]
+
+# alternate p-value calculations between sequential rows
+p_values_alt <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(data.frame(clade1=c(as.numeric(c[3]),as.numeric(c[4])),clade2=c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))
+```
+
 For each comparison, calculate the probability of obtaining the observed and expected counts assuming there is no difference in selection on mutations between the clades being compared (i.e., p-value) using the Fisher's Exact test.
@@
 ```{r contingency_analysis}
-# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
-'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))'
+# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. Mac Desktop Local: "334.273  41.641 375.678"
+p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))
 

### 2022-10-26 notebooks/expected_and_actual_counts_ratio.Rmd
Modify code and restructure data frames to speed up time-consuming commands

@@
 # Define a minimum threshold for the number of expected counts. Mutations below this threshold will be removed.
-min_expected <- 10
+min_expected <- 0
 
@@
 
-```{r temp}
-filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
+To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed and expected values for clade 1 and clade 2.
+
+```{r contingency_matrix}
+# Add observed and expected values for each comparison to filtered comparisons data frame.
+{filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
 filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
 filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
-filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]
+filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]}
 
-# alternate p-value calculations between sequential rows
-p_values_alt <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(data.frame(clade1=c(as.numeric(c[3]),as.numeric(c[4])),clade2=c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))
+#
+{filtered_comparisons$clade1_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed_expected"]
+filtered_comparisons$clade2_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed_expected"]}
 ```
@@
 
-```{r contingency_analysis}
-# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. Mac Desktop Local: "334.273  41.641 375.678"
-p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))
+```{r contingency_table}
+# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
+# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
+'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'
 
@@
 load("fisher_pvalues.Rda")
+```
 
+Combine the p-values calculated using the Fisher's Exact test with the data frame of filtered comparisons.
+
+```{r combine_p_values}
 # Combine p-value list with the clade comparison list.
@@
 # False Discovery Rate and Bonferroni Corrections
-
+# NEED TO CORRECT FOR MORE THAN 2 CLADE COMPARISONS
+comparison_p_values$p_values_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
 ```
@@
 # Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
-comparison_p_values$fold_change <- apply(comparison_p_values[,c("clade1","clade2")], 1, FUN=function(c) {foldchange(fisher_data[which(fisher_data$nt_mutation_clade==c[1]), "observed_expected"], fisher_data[which(fisher_data$nt_mutation_clade==c[2]), "observed_expected"])})
+comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
 ```

### 2022-10-27 notebooks/expected_and_actual_counts_ratio.Rmd
Add Bonferroni corrected p-values and ensure code works >2 clades

@@
 # Create a list of the clades to include in comparisons.
-clades <- c("20A","20B")
+clades <- c("20A","20B","20C")
 
@@
 # Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
+# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
 filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
@@
 
-To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed and expected values for clade 1 and clade 2.
+To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed counts, expected counts and observed/expected ratios for clade 1 and clade 2.
 
 ```{r contingency_matrix}
-# Add observed and expected values for each comparison to filtered comparisons data frame.
-{filtered_comparisons$clade1_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed"]
-filtered_comparisons$clade1_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"expected"]
-filtered_comparisons$clade2_observed <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed"]
-filtered_comparisons$clade2_expected <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"expected"]}
+# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
+filtered_comparisons$id <- c(1:nrow(filtered_comparisons))
 
-#
-{filtered_comparisons$clade1_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade1),"observed_expected"]
-filtered_comparisons$clade2_obs_exp <- fisher_data[which(fisher_data$nt_mutation_clade %in% filtered_comparisons$clade2),"observed_expected"]}
+# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
+clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
+clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")
+
+# Add observed, expected and observed/expected data to the comparison matrix.
+{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
+filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
+filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
+filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
+filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
+filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
 ```
@@
 # Save and load variable because calculating p-values takes a long time.
-'save(p_values, file="fisher_pvalues.Rda")'
-load("fisher_pvalues.Rda")
+'save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))'
+load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
 ```
@@
 ```{r p_value_correction}
-# False Discovery Rate and Bonferroni Corrections
-# NEED TO CORRECT FOR MORE THAN 2 CLADE COMPARISONS
-comparison_p_values$p_values_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
+# Identify clades from the two columns in the comparison matrix and combine into one column.
+comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))
+
+# Count the number of rows for each clade (i.e., number of hypothesis tests).
+clade.table <- table(comparison_p_values$clades)
+
+# Add a column for the number of hypothesis tests conducted for each pair of clades.
+comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
+
+# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
+comparison_p_values$p_values_corrected <- comparison_p_values$p_value/comparison_p_values$num_tests
 ```
@@
 ```{r volcano_plot}
-ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+# Only plot data for the two clades being compared.
+plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
+ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
   geom_point()

### 2022-11-01 notebooks/expected_and_actual_counts_ratio.Rmd
Add alternative code that can only handle 2 clades and round before calculating p-values

@@
 library(stringr)
-library(gtools)
 ```
@@
 # Create a list of the clades to include in comparisons.
-clades <- c("20A","20B","20C")
+clades <- c("22A","22B")
 
@@
 # Add column with the mutation type (clade founder.mutation).
-filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
+#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 ```
@@
 
+```{r round_expected}
+# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
+filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
+```
+
 ## Characterize the Filtered Data
@@
 ```{r head_data}
-head(filtered_SARS_mutation, n=5)
+#head(filtered_SARS_mutation, n=5)
 ```
@@
 # How many rows? How many columns?
-dim(filtered_SARS_mutation)
+#dim(filtered_SARS_mutation)
 
 # How many phylogenetic clades?
-length(unique(filtered_SARS_mutation$clade))
+#length(unique(filtered_SARS_mutation$clade))
 
 # How many nucleotide sites?
-length(unique(filtered_SARS_mutation$nt_site))
+#length(unique(filtered_SARS_mutation$nt_site))
 ```
@@
 # Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
-ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
-  geom_bar(position="stack", stat="identity")+
-  theme_bw()+
-  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
+#ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
+  #geom_bar(position="stack", stat="identity")+
+  #theme_bw()+
+  #labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
 ```
@@
 ```{r nucleotide_site, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
-  geom_point()+
-  theme_bw()+
-  labs(x="", y="")
+#ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
+  #geom_point()+
+  #theme_bw()+
+  #labs(x="", y="")
 ```
@@
 # Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
-fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$expected_count, observed_expected=clade_SARS_mutation$observed_expected)
+fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_mutation, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$rounded_expected, observed_expected=clade_SARS_mutation$observed_expected)
 ```
@@
 
-For each mutation type (e.g., A -> T) at each nucleotide site, this notebook will generate contingency table analyses for each comparison in the set of clade comparisons.
-To expand the clade comparison matrix to include mutation types at nucleotide sites, generate a unique list of each mutation type (i.e., three mutations for one clade founder) at each nucleotide site.
-
-```{r unique_mutations}
-# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
-nt_mutation_list <- unique(paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, sep="_"))
-```
-
 Expand both the clade comparison matrix and the mutation list to create a combined matrix of all possible clade comparisons for each mutation type at each nucleotide site.
@@
 # In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
-rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))
+rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_SARS_mutation$nt_mutation), each=nrow(clade_comparisons)))
 
 # In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
-rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
+rep_clades <- do.call("rbind", replicate(length(unique(filtered_SARS_mutation$nt_mutation)), clade_comparisons, simplify=FALSE))
 
@@
 # Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
-# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
-filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
+filtered_comparisons <- comparisons[(comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade),]
 ```
@@
 # NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
-'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'
+p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))
 
@@
 # Identify clades from the two columns in the comparison matrix and combine into one column.
-comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))
+comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))
 
@@
 # Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
-comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
+# Double check calculation with Jesse's paper
+comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_obs_exp >= comparison_p_values$clade2_obs_exp, comparison_p_values$clade1_obs_exp/comparison_p_values$clade2_obs_exp, -comparison_p_values$clade2_obs_exp/comparison_p_values$clade1_obs_exp))
 ```
@@
 # Only plot data for the two clades being compared.
-plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
-ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
+#plot_data <- comparison_p_values[which(comparison_p_values$clades=="20A.22A"),]
+ggplot(comparison_p_values, aes(x=other_fold, y=-log10(p_values_corrected)))+
   geom_point()
@@
 ```
+
+## Shorted Script That Can Handle 2 Clades
+```{r two_clades}
+# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
+# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
+{start_two <- Sys.time()
+SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
+country <- "all"
+clades <- c("20A","20B")
+min_expected <- 0
+pseudo <- 0.5
+filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
+clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
+clade1 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[1]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
+clade2 <- clade_SARS_mutation[which(clade_SARS_mutation$clade==clades[2]),c("nt_mutation","clade","actual_count","expected_count","observed_expected")]
+m <- merge(clade1, clade2, by="nt_mutation")
+filtered_comparisons <- data.frame(clade1=paste(m$nt_mutation, m$clade.x, sep="_"), clade2=paste(m$nt_mutation, m$clade.y, sep="_"), clade1_observed=m$actual_count.x, clade1_expected=m$expected_count.x, clade2_observed=m$actual_count.y, clade2_expected=m$expected_count.y, clade1_obs_exp=m$observed_expected.x, clade2_obs_exp=m$observed_expected.y)
+end_two <- Sys.time()}
+end_two - start_two
+two <- filtered_comparisons
+```

### 2022-11-02 notebooks/expected_and_actual_counts_ratio.Rmd
Plot histogram of p-values

@@
 library(stringr)
+library(gtools)
 ```
@@
 # Create a list of the clades to include in comparisons.
-clades <- c("22A","22B")
+clades <- c("20A","22B")
 
@@
 filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
-
-# Add column with the mutation type (clade founder.mutation).
-#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 ```
@@
 ```
+
+The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).
 
@@
 ```{r stacked_bar, echo=FALSE}
+# Add column with the mutation type (clade founder.mutation).
+#filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
+
 # Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
@@
 # Create a subset of the data using only the clades that are going to be compared.
-clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
+filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
 ```
@@
 # Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
-fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_mutation, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$rounded_expected, observed_expected=clade_SARS_mutation$observed_expected)
+filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")
 ```
@@
 # In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
-rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_SARS_mutation$nt_mutation), each=nrow(clade_comparisons)))
+rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))
 
 # In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
-rep_clades <- do.call("rbind", replicate(length(unique(filtered_SARS_mutation$nt_mutation)), clade_comparisons, simplify=FALSE))
+rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))
 
@@
 # Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
-filtered_comparisons <- comparisons[(comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade),]
+filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]
 ```
@@
 # Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
-clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
-clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")
+clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
+clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")
 
-# Add observed, expected and observed/expected data to the comparison matrix.
-{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
-filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
-filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
-filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
-filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
-filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
+# Change column names so that clade 1 and clade 2 data frames can be combined
+colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
+colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")
+
+# Order clade 1 and clade 2 data frames by id and combine
+fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
+fisher_data[,which(colnames(fisher_data) == "id")] <- NULL
 ```
@@
 # NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
-p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))
+'p_values <- data.frame(p_value=apply(fisher_data, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'
 
@@
 # Combine p-value list with the clade comparison list.
-comparison_p_values <- cbind(filtered_comparisons, p_values)
+comparison_p_values <- cbind(fisher_data, p_values)
 ```
@@
 # Double check calculation with Jesse's paper
-comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_obs_exp >= comparison_p_values$clade2_obs_exp, comparison_p_values$clade1_obs_exp/comparison_p_values$clade2_obs_exp, -comparison_p_values$clade2_obs_exp/comparison_p_values$clade1_obs_exp))
+comparison_p_values$fold_change <- log2(ifelse(comparison_p_values$clade1_observed_expected >= comparison_p_values$clade2_observed_expected, comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected, -comparison_p_values$clade2_observed_expected/comparison_p_values$clade1_observed_expected))
 ```
@@
 #plot_data <- comparison_p_values[which(comparison_p_values$clades=="20A.22A"),]
-ggplot(comparison_p_values, aes(x=other_fold, y=-log10(p_values_corrected)))+
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_values_corrected)))+
   geom_point()
@@
 # sanity check analyses (distribution of synonymous p-values)
+plot_data <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
+nb <- sqrt(nrow(plot_data))
+bw <- (max(comparison_p_values$p_values_corrected)-min(comparison_p_values$p_values_corrected))/nb
+ggplot(plot_data, aes(x=p_values_corrected))+
+  geom_histogram(bins = nb, binwidth = bw)
 
+ggplot(comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),], aes(x=p_values_corrected, y=p_values_corrected))+
+  geom_point()
 ```
@@
 # Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
-{start_two <- Sys.time()
+'{start_two <- Sys.time()
 SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
@@
 end_two - start_two
-two <- filtered_comparisons
+two <- filtered_comparisons'
 ```

### 2022-11-03 notebooks/expected_and_actual_counts_ratio.Rmd
Modify and simplify code to only handle two clades

--- 2022-10-27 notebooks/expected_and_actual_counts_ratio.Rmd
@@
 
-# Create a list of the clades to include in comparisons.
-clades <- c("20A","20B","20C")
+# Identify the two clades that will be included in comparisons.
+clades <- c("20A","22B")
 
@@
 filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
-
-# Add column with the mutation type (clade founder.mutation).
-filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 ```
@@
 filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + pseudo)/(filtered_SARS_mutation$expected_count + pseudo)
+```
+
+The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).
+
+```{r round_expected}
+# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
+filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
 ```
@@
 ```{r stacked_bar, echo=FALSE}
+# Add column with the mutation type (clade founder.mutation).
+add_mutation <- filtered_SARS_mutation
+add_mutation$mutation_type <- paste(str_sub(add_mutation$nt_mutation,1,1), ".", str_sub(add_mutation$nt_mutation,-1,-1), sep="")
+
 # Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
-ggplot(filtered_SARS_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
+ggplot(add_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
   geom_bar(position="stack", stat="identity")+
@@
 ```{r nucleotide_site, echo=FALSE}
-ggplot(filtered_SARS_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
+ggplot(add_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
   geom_point()+
@@
   labs(x="", y="")
-```
-
-[INSERT] information about the plot. Plot to visualize data.
-
-```{r INSERT, echo=FALSE}
-#ggplot(filtered_SARS_mutation, aes(fill=INSERT, y=INSERT, x=INSERT))+
-  #geom_tile()+
-  #theme_bw()+
-  #labs(x="", y="")
 ```
@@
 # Create a subset of the data using only the clades that are going to be compared.
-clade_SARS_mutation <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
+filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
 ```
 
-Combine nucleotide site, mutation type and clade to create a reference column in the data set.
-This column will be referenced to obtain observed and expected counts for clade comparisons.
-Additionally, remove columns from the data set not necessary for contingency table analyses or creating volcano plots.
+```{r restructure_data}
+# Create two data frames with data for clade 1 and clade 2.
+clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
+clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
 
-```{r filter_contingency_data}
-# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
-fisher_data <- data.frame(nt_mutation_clade=paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, clade_SARS_mutation$clade, sep="_"), observed=clade_SARS_mutation$actual_count, expected=clade_SARS_mutation$expected_count, observed_expected=clade_SARS_mutation$observed_expected)
-```
+# Change column names so that clade 1 and clade 2 data frames can be combined
+colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
+colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
+colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
+colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"
 
-Create a two-column matrix of all possible clade comparisons, each column listing one clade in the comparison. This matrix should have [n x (n-1)]/2 rows where n is the number of clades.
-
-```{r clade_comparisons}
-# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
-clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
-```
-
-For each mutation type (e.g., A -> T) at each nucleotide site, this notebook will generate contingency table analyses for each comparison in the set of clade comparisons.
-To expand the clade comparison matrix to include mutation types at nucleotide sites, generate a unique list of each mutation type (i.e., three mutations for one clade founder) at each nucleotide site.
-
-```{r unique_mutations}
-# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
-nt_mutation_list <- unique(paste(clade_SARS_mutation$nt_site, clade_SARS_mutation$mutation_type, sep="_"))
-```
-
-Expand both the clade comparison matrix and the mutation list to create a combined matrix of all possible clade comparisons for each mutation type at each nucleotide site.
-To do this, repeat the clade comparison matrix by the number of unique mutations, and repeat each mutation in the mutation list by the number of clade comparisons.
-
-```{r combined_clade_comparisons}
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
-rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))
-
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
-rep_clades <- do.call("rbind", replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))
-
-# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
-comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
-```
-
-Not every mutation will exist in each clade. Therefore, remove comparisons where observed and expected count data does not exist for one or both clades in the comparison.
-
-```{r filter_for_comparisons}
-# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
-# DOUBLE CHECK THIS: NOT CONVINCED IT IS WORKING
-filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]
-```
-
-To decrease the amount of time it takes to run the contingency tests, add columns to the filtered comparisons data frame with the observed counts, expected counts and observed/expected ratios for clade 1 and clade 2.
-
-```{r contingency_matrix}
-# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
-filtered_comparisons$id <- c(1:nrow(filtered_comparisons))
-
-# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
-clade1 <- merge(filtered_comparisons[,c("id","clade1")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade1", by.y="nt_mutation_clade")
-clade2 <- merge(filtered_comparisons[,c("id","clade2")], fisher_data[,c("nt_mutation_clade","observed","expected","observed_expected")], by.x="clade2", by.y="nt_mutation_clade")
-
-# Add observed, expected and observed/expected data to the comparison matrix.
-{filtered_comparisons$clade1_observed <- clade1[order(clade1$id),"observed"]
-filtered_comparisons$clade1_expected <- clade1[order(clade1$id),"expected"]
-filtered_comparisons$clade2_observed <- clade2[order(clade2$id),"observed"]
-filtered_comparisons$clade2_expected <- clade2[order(clade2$id),"expected"]
-filtered_comparisons$clade1_obs_exp <- clade1[order(clade1$id),"observed_expected"]
-filtered_comparisons$clade2_obs_exp <- clade2[order(clade2$id),"observed_expected"]}
+# Merge the two data frames by mutation so that each row represents one comparison.
+filtered_comparisons <- merge(clade1, clade2, by="nt_mutation")
 ```
@@
 ```{r p_value_correction}
-# Identify clades from the two columns in the comparison matrix and combine into one column.
-comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,3], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,3], sep="."))
-
-# Count the number of rows for each clade (i.e., number of hypothesis tests).
-clade.table <- table(comparison_p_values$clades)
-
-# Add a column for the number of hypothesis tests conducted for each pair of clades.
-comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
-
 # Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
-comparison_p_values$p_values_corrected <- comparison_p_values$p_value/comparison_p_values$num_tests
+comparison_p_values$p_value_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
 ```
@@
 # Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
-comparison_p_values$fold_change <- foldchange(comparison_p_values$clade1_obs_exp, comparison_p_values$clade2_obs_exp)
+# Double check calculation with Jesse's paper
+comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
 ```
@@
 # Only plot data for the two clades being compared.
-plot_data <- comparison_p_values[which(comparison_p_values$clades=="20B.20C"),]
-ggplot(plot_data, aes(x=fold_change, y=-log10(p_value)))+
-  geom_point()
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
+  geom_point()+
+  theme_bw()
 ```
@@
 # sanity check analyses (distribution of synonymous p-values)
+verify_analyses <- comparison_p_values
+verify_analyses <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
+nb <- sqrt(nrow(verify_analyses))
+bw <- (max(verify_analyses$p_value_corrected)-min(verify_analyses$p_value_corrected))/nb
+ggplot(verify_analyses, aes(x=p_value_corrected))+
+  geom_histogram(bins = nb, binwidth = bw)
+```
 
+```{r alternative, echo=FALSE}
+# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
+# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
+
+# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
+#filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")
+
+# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
+#clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
+
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
+#rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))
+
+# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
+#rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))
+
+# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
+#comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
+
+# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
+#filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]
+
+# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
+#filtered_comparisons$id <- c(1:nrow(filtered_comparisons))
+
+# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
+#clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
+#clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")
+
+# Change column names so that clade 1 and clade 2 data frames can be combined
+#colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
+#colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")
+
+# Order clade 1 and clade 2 data frames by id and combine
+#fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
+#fisher_data[,which(colnames(fisher_data) == "id")] <- NULL
+
+# Identify clades from the two columns in the comparison matrix and combine into one column.
+#comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))
+
+# Count the number of rows for each clade (i.e., number of hypothesis tests).
+#clade.table <- table(comparison_p_values$clades)
+
+# Add a column for the number of hypothesis tests conducted for each pair of clades.
+#comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
 ```

### 2022-11-04 notebooks/expected_and_actual_counts_ratio.Rmd
Jitter plot to verify analyses and fix Bonferroni correction

@@
 output: html_document
-date: "`r format(Sys.Date(), '%d.%m.%Y')`"
+date: "`r format(Sys.Date(), '%Y-%m-%d')`"
 ---
@@
 
-[INSERT] Discuss purpose of notebook and steps to achieve that goal.
+The purpose of this notebook is to generate volcano plots comparing any two SARS-CoV-2 phylogenetic clades.
+Using the volcano plots, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified.
+Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing the 78 clade comparisons to be computed in parallel.
 
-First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using either access to the server or from github.com.
+Listed below are the general steps the notebook uses to generate a volcano plot:\
+1. Read in the SARS-CoV-2 data containing observed and expected mutation counts.\
+2. Set parameters for how to compute comparisons, including what clades to compare and what countries to accept data from.\
+3. Filter data based on parameters.\
+4. Restructure data such that there are columns for clade 1 variables and clade 2 variables.\
+5. Using contingency table analyses (e.g., Fisher's Exact test), calculate p-values for each mutation at every nucleotide site between the two clades.\
+6. Adjust p-values to account for the large number of hypothesis tests.\
+7. Using the ratio of observed counts to expected counts for each clade and a pseudocount, calculate log2 fold-change.\
+8. Generate a volcano plot with log2 fold-change on the X-axis and -log10 p-values on the Y-axis.\
+9. Verify the notebook is computing p-values correctly by generating plots with expected outcomes (e.g., synonymous mutations).\
+
+First, read in the comma-delimited data from the 'kfeldmann' computational notebook in the Bloom Lab folder on the Fred Hutch server using access to the server or from the Bloom Lab github.com.
 
@@
 # Identify the two clades that will be included in comparisons.
-clades <- c("20A","22B")
+clades <- c("21J","22B")
 
@@
 
-For the analyses conducted in this notebook, subset the data by the country(countries) the sequence data came from.
+For the analyses conducted in this notebook, subset the data by the country (countries) the sequence data came from.
 Additionally, remove low-quality mutations or branches with abnormally high mutation rates.
-Finally, create a column with mutation type (there are 12 possible mutations).
 
@@
 
-To compare mutations between phylogenetic clades, this notebook will plot volcano plots with p-value on the Y-axis and fold-change on the X-axis.
+To compare mutations between phylogenetic clades, this notebook will plot volcano plots with -log10 p-value on the Y-axis and log2 fold-change on the X-axis.
 Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated.
@@
 
-The Fisher's Exact test can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).
+Contingency table analyses can only accept integers, so round the expected counts prior to running analyses (i.e., gives control over how values are rounded).
 
 ```{r round_expected}
-# Round expected counts because contingency table analyses (i.e., Fisher's Exact test) can only analyze integers
+# Round expected counts because contingency table analyses (e.g., Fisher's Exact test) can only analyze integers.
 filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
@@
 
-## Characterize the Filtered Data
+## Describe and Characterize the Data
 
@@
 
-# How many phylogenetic clades?
-length(unique(filtered_SARS_mutation$clade))
-
 # How many nucleotide sites?
@@
 
-## Visualize the Filtered Data
+## Visualize the Data
 
@@
 
-## Contingency Table Analyses
+## Restructure Data for Analyses
 
-With such a large data set, it will take potentially hours to calculate the p-values for clade comparisons.
-To decrease the amount of time it takes to conduct analyses, create a subset of the data with only the clades to be compared.
+With such a large data set, it could take hours to calculate the p-values for all clade comparisons.
+To decrease the amount of time it takes to conduct analyses, this notebook will only compare two clades at a time and the script can be run in parallel to generate all comparisons at once.
+Therefore, filter the data to only include mutation data from the two clades being compared.
 
@@
 
+To quickly calculate the p-value for each contingency table analysis and ensure the user can easily understand what is being compared in each hypothesis test, restructure the data such that all of the variables necessary to run the statistical test are in one row.
+That includes observed and expected counts for clade 1 and observed and expected counts for clade 2.
+
 ```{r restructure_data}
@@
 
-# Change column names so that clade 1 and clade 2 data frames can be combined
+# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
 colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
@@
 # Merge the two data frames by mutation so that each row represents one comparison.
-filtered_comparisons <- merge(clade1, clade2, by="nt_mutation")
+comparisons <- merge(clade1, clade2, by="nt_mutation")
 ```
+
+## Contingency Table Analyses
 
@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-# NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test. 
-'p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c[3]),as.numeric(c[4])),c(as.numeric(c[5]),as.numeric(c[6]))))$p.value}))'
+p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["clade1_actual_count"]),as.numeric(c["clade1_rounded_expected"])),c(as.numeric(c["clade2_actual_count"]),as.numeric(c["clade2_rounded_expected"]))))$p.value}))
 
-# Save and load variable because calculating p-values takes a long time.
-'save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))'
+# Save and load variable because calculating p-values may take a long time.
+#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
 load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
@@
 
-Combine the p-values calculated using the Fisher's Exact test with the data frame of filtered comparisons.
+Combine the p-values calculated using the contingency table analyses with the data frame of comparisons.
 
 ```{r combine_p_values}
-# Combine p-value list with the clade comparison list.
-comparison_p_values <- cbind(filtered_comparisons, p_values)
+# Combine the p-value list with the comparison list.
+comparison_p_values <- cbind(comparisons, p_values)
 ```
@@
 # Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
-comparison_p_values$p_value_corrected <- comparison_p_values$p_value/nrow(comparison_p_values)
+comparison_p_values$p_value_corrected <- comparison_p_values$p_value*nrow(comparison_p_values)
 ```
@@
 
-Use a function to calculate fold-change using the observed-expected ratios calculated with a pseudocount.
-This value will be the X-axis for the volcano plot.
-P-values clustering less than zero indicate a difference in degree of selection for one clade, whereas p-values clustering greater than zero indicate a similar difference in degree of selection but for the other clade.
+Calculate fold-change using the observed-expected ratios calculated with a pseudocount.
+Fold-change is calculated as log2 of the clade 1 ratio divided by the clade 2 ratio.
+Therefore, if fold-change is positive then the clade 1 ratio is greater than the clade 2 ratio, and vise versa if fold-change is negative.
 
 ```{r fold-change}
-# Calculate fold change for plotting volcano plots [clade 1 observed/clade 1 expected]/[clade 2 observed/clade 2 expected].
-# Double check calculation with Jesse's paper
+# Calculate fold change for plotting volcano plots.
 comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
@@
 
-Plot the volcano plot with -log10 p-values on the Y-axis and log fold-change on the X-axis.
+Generate a volcano plot with -log10 p-values on the Y-axis and log2 fold-change on the X-axis.
 
-```{r volcano_plot}
-# Only plot data for the two clades being compared.
+```{r volcano_plot, echo=FALSE}
 ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
   geom_point()+
+  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
   theme_bw()
@@
 
-## Verify Analyses Using Analyses With Expected Outcomes
+## Verify Analyses
+
+To verify the analyses are generating expected results, compare the p-values for synonymous mutations to non-synonymous mutations.
+Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures.
+Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-value close to 0), there should be insignificant differences between clades on synonymous mutations (i.e., p-values close to 1).
 
@@
 verify_analyses <- comparison_p_values
-verify_analyses <- comparison_p_values[which(comparison_p_values$clade1_four_fold_degenerate=="TRUE"),]
-nb <- sqrt(nrow(verify_analyses))
-bw <- (max(verify_analyses$p_value_corrected)-min(verify_analyses$p_value_corrected))/nb
-ggplot(verify_analyses, aes(x=p_value_corrected))+
-  geom_histogram(bins = nb, binwidth = bw)
+verify_analyses$synonymous <- paste(comparison_p_values$clade1_synonymous, comparison_p_values$clade2_synonymous, sep=".")
+
+ggplot(verify_analyses, aes(x=synonymous, y=-log10(p_value_corrected)))+
+  geom_jitter(alpha=0.5)+
+  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)
 ```
-
-```{r alternative, echo=FALSE}
-# Script that can handle 2+ clades: 1.380048, 1.151285, 1.034135 minutes
-# Script that can only handle 2 clades: 50.01184, 58.11915, 58.6249 seconds
-
-# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
-#filtered_clades$nt_mutation_clade <- paste(filtered_clades$nt_mutation, filtered_clades$clade, sep="_")
-
-# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
-#clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)
-
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
-#rep_nt_mutation <- data.frame(nt_mutation=rep(unique(filtered_clades$nt_mutation), each=nrow(clade_comparisons)))
-
-# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
-#rep_clades <- do.call("rbind", replicate(length(unique(filtered_clades$nt_mutation)), clade_comparisons, simplify=FALSE))
-
-# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
-#comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))
-
-# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
-#filtered_comparisons <- comparisons[(comparisons$clade1 %in% filtered_clades$nt_mutation_clade & comparisons$clade2 %in% filtered_clades$nt_mutation_clade),]
-
-# Create an ID to maintain the order of rows (and thus comparisons) after merging data frames.
-#filtered_comparisons$id <- c(1:nrow(filtered_comparisons))
-
-# Merge each of the clades in the comparison matrix with the data to align observed, expected and observed/expected values.
-#clade1 <- merge(filtered_comparisons[,c("id","clade1")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade1", by.y="nt_mutation_clade")
-#clade2 <- merge(filtered_comparisons[,c("id","clade2")], filtered_clades[,c("nt_mutation_clade","actual_count","rounded_expected","expected_count","observed_expected","synonymous","four_fold_degenerate")], by.x="clade2", by.y="nt_mutation_clade")
-
-# Change column names so that clade 1 and clade 2 data frames can be combined
-#colnames(clade1)[3:ncol(clade1)] <- paste("clade1", colnames(clade1)[3:ncol(clade1)], sep="_")
-#colnames(clade2)[3:ncol(clade2)] <- paste("clade2", colnames(clade2)[3:ncol(clade2)], sep="_")
-
-# Order clade 1 and clade 2 data frames by id and combine
-#fisher_data <- cbind(clade1[order(clade1$id),], clade2[order(clade2$id),])
-#fisher_data[,which(colnames(fisher_data) == "id")] <- NULL
-
-# Identify clades from the two columns in the comparison matrix and combine into one column.
-#comparison_p_values$clades <- data.frame(clades=paste(str_split(comparison_p_values$clade1, "_", simplify=TRUE)[,2], str_split(comparison_p_values$clade2, "_", simplify=TRUE)[,2], sep="."))
-
-# Count the number of rows for each clade (i.e., number of hypothesis tests).
-#clade.table <- table(comparison_p_values$clades)
-
-# Add a column for the number of hypothesis tests conducted for each pair of clades.
-#comparison_p_values$num_tests <- apply(comparison_p_values$clades, 1, FUN=function(c){clade.table[[c[1]]]})
-```

### 2022-11-07 notebooks/expected_and_actual_counts_ratio.Rmd
Compare synonymous and non-synonymous mutations on a regular scale

@@
 #install.packages("ggplot2")
+#install.packages("ggforce")
 #install.packages("stringr")
@@
 library(ggplot2)
+library(ggforce)
 library(stringr)
@@
 
-The purpose of this notebook is to generate volcano plots comparing any two SARS-CoV-2 phylogenetic clades.
-Using the volcano plots, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified.
-Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing the 78 clade comparisons to be computed in parallel.
+The purpose of this notebook is to generate a volcano plot comparing any two SARS-CoV-2 phylogenetic clades.
+Using the volcano plot, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified.
+Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing all clade comparisons to be computed in parallel.
 
@@
 # Identify the two clades that will be included in comparisons.
-clades <- c("21J","22B")
+clades <- c("20A","22B")
 
@@
 filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
+
+# Remove columns that were already used to filter the data or than contain redundant information.
+filtered_clades$subset <- NULL
+filtered_clades$exclude <- NULL
+filtered_clades$nt_site <- NULL
 ```
@@
 # Create two data frames with data for clade 1 and clade 2.
-clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
-clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),c("nt_mutation","clade","actual_count","rounded_expected","observed_expected","synonymous","four_fold_degenerate")]
+clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),]
+clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),]
 
 # Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
-colnames(clade1)[2:ncol(clade1)] <- paste("clade1", colnames(clade1)[2:ncol(clade1)], sep="_")
-colnames(clade2)[2:ncol(clade2)] <- paste("clade2", colnames(clade2)[2:ncol(clade2)], sep="_")
-colnames(clade1)[which(colnames(clade1)=="clade1_clade")] <- "clade1"
-colnames(clade2)[which(colnames(clade2)=="clade2_clade")] <- "clade2"
+colnames(clade1) <- paste("c1", colnames(clade1), sep="_")
+colnames(clade2) <- paste("c2", colnames(clade2), sep="_")
+
+# Modify nt_mutation, clade_founder_nt and clade to make merging easier and make the data frame more readable.
+{colnames(clade1)[which(colnames(clade1)=="c1_nt_mutation")] <- "nt_mutation"
+colnames(clade2)[which(colnames(clade2)=="c2_nt_mutation")] <- "nt_mutation"
+colnames(clade1)[which(colnames(clade1)=="c1_clade_founder_nt")] <- "clade_founder_nt"
+colnames(clade2)[which(colnames(clade2)=="c2_clade_founder_nt")] <- "clade_founder_nt"
+colnames(clade1)[which(colnames(clade1)=="c1_clade")] <- "clade1"
+colnames(clade2)[which(colnames(clade2)=="c2_clade")] <- "clade2"}
 
 # Merge the two data frames by mutation so that each row represents one comparison.
-comparisons <- merge(clade1, clade2, by="nt_mutation")
+comparisons <- merge(clade1, clade2, by=c("nt_mutation","clade_founder_nt"))
 ```
@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["clade1_actual_count"]),as.numeric(c["clade1_rounded_expected"])),c(as.numeric(c["clade2_actual_count"]),as.numeric(c["clade2_rounded_expected"]))))$p.value}))
+p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["c1_actual_count"]),as.numeric(c["c1_rounded_expected"])),c(as.numeric(c["c2_actual_count"]),as.numeric(c["c2_rounded_expected"]))))$p.value}))
 
@@
 ```{r p_value_correction}
-# Using Bonferroni Correction, adjust the p-values by the number of hypothesis tests.
-comparison_p_values$p_value_corrected <- comparison_p_values$p_value*nrow(comparison_p_values)
+# Using False Discovery Rate Correction, adjust the p-values by the number of hypothesis tests.
+comparison_p_values$p_value_corrected <- p.adjust(comparison_p_values$p_value, method="fdr", n=nrow(comparison_p_values))
 ```
@@
 # Calculate fold change for plotting volcano plots.
-comparison_p_values$fold_change <- log2(comparison_p_values$clade1_observed_expected/comparison_p_values$clade2_observed_expected)
+comparison_p_values$fold_change <- log2(comparison_p_values$c1_observed_expected/comparison_p_values$c2_observed_expected)
 ```
@@
 ```{r volcano_plot, echo=FALSE}
-ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value_corrected)))+
+# Plot the uncorrected p-values.
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+  geom_point(alpha=0.5)+
+  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
+  theme_bw()
+
+# After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
+volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
+ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=nt_mutation, color=clade_founder_nt))+
   geom_point()+
+  geom_text(color="black", size=2, nudge_y=0.2)+
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
+  scale_color_brewer(palette="Set2")+
   theme_bw()
@@
 
-To verify the analyses are generating expected results, compare the p-values for synonymous mutations to non-synonymous mutations.
+To verify the analyses are generating expected results, compare the corrected p-values for synonymous mutations to non-synonymous mutations.
 Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures.
-Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-value close to 0), there should be insignificant differences between clades on synonymous mutations (i.e., p-values close to 1).
+Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-values close to 0), there should be insignificant differences between clades for synonymous mutations (i.e., p-values close to 1).
 
 ```{r verify_analyses}
-# sanity check analyses (distribution of synonymous p-values)
+# Plot the corrected p-values for synonymous and non-synonymous mutations and compare to a significance value of 0.05.
 verify_analyses <- comparison_p_values
-verify_analyses$synonymous <- paste(comparison_p_values$clade1_synonymous, comparison_p_values$clade2_synonymous, sep=".")
+verify_analyses$synonymous <- ifelse(verify_analyses$c1_synonymous=="TRUE" & verify_analyses$c2_synonymous=="TRUE", "TRUE", "FALSE")
+verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
 
-ggplot(verify_analyses, aes(x=synonymous, y=-log10(p_value_corrected)))+
-  geom_jitter(alpha=0.5)+
-  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)
+p <- ggplot(verify_analyses, aes(x=p_value_corrected, fill=synonymous, label=..count..))+
+  geom_histogram()+
+  geom_text(size=2, vjust=-0.5, stat="bin", position="stack")+
+  geom_vline(data=data.frame(xint=0.05, p_one="P<1"), aes(xintercept=xint), linetype=2)+
+  scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1))+
+  labs(x="FDR corrected p-values")+
+  theme_bw()
+
+p + ggforce::facet_row(vars(p_one), scales="free", space="free")
 ```

### 2022-11-09 notebooks/expected_and_actual_counts_ratio.Rmd
Add ability to compare clades using amino acids

@@
 library(gtools)
+
+`%notin%` <- Negate(`%in%`)
 ```
@@
 pseudo <- 0.5
+
+option <- "aa" # nt or aa
+merge_columns <- c("nt_mutation","nt_site","clade_founder_nt","gene","codon_position")
+merge_columns <- c("aa_mutation","gene")
 ```
@@
 ## Filter and Prepare the SARS-CoV-2 Mutation Data
+
+With the data set, both nucleotide and amino acid mutations can be compared between clades.
+However, there is no column identifying the amino acid mutation (i.e., starting amino acid, codon position along the genome and mutated amino acid).
+Create a column with this information using the nucleotide site, codon position (i.e., 1, 2, or 3), clade founder amino acid and mutant amino acid columns.
+
+```{r aa_mutation}
+# Create a list with the starting nucleotide site for an amino acid for all rows in the data set.
+starting_nt <- SARS_mutation$nt_site-as.numeric(str_sub(SARS_mutation$codon_position,1,1))+1
+
+# Create column with amino acid mutation (e.g., M.266.268.L)
+SARS_mutation$aa_mutation <- paste(str_sub(SARS_mutation$clade_founder_aa,1,1), starting_nt, starting_nt+2, str_sub(SARS_mutation$mutant_aa,1,1), sep=".")
+```
 
@@
 
+```{r clean_columns}
+# Remove columns that were already used to filter the data.
+filtered_SARS_mutation$subset <- NULL
+filtered_SARS_mutation$exclude <- NULL
+```
+
 ## Describe and Characterize the Data
@@
 
-Describe the filtered data set.
+Look at a summary of the data.\
+-What are the data types for each column?\
+-Is there a similar number of TRUEs and FALSEs for logical variables?\
+-Describe the numeric variables (e.g., minimum value).\
+
+```{r summary_data}
+summary(filtered_SARS_mutation)
+```
+
+Take a closer look at how many nucleotide sites and the proportion of mutations in the data set.
 
 ```{r describe_data}
-# How many rows? How many columns?
-dim(filtered_SARS_mutation)
-
 # How many nucleotide sites?
 length(unique(filtered_SARS_mutation$nt_site))
+
+# What percentage of the data set is each type of mutation? (e.g., A -> T)
+mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
+(table(mutation_type)/length(mutation_type))*100
 ```
@@
 
-[INSERT] information about the plot. Plot to visualize data.
+The four nucleotides (e.g., A) and their possible mutations (e.g., A -> T) may be under different directions and degrees of selection between clades leading to an unequal distribution of nucleotides in the genome.
+Look at how many mutations (normalized by the expected count) there are for each clade.
 
-```{r stacked_bar, echo=FALSE}
-# Add column with the mutation type (clade founder.mutation).
-add_mutation <- filtered_SARS_mutation
-add_mutation$mutation_type <- paste(str_sub(add_mutation$nt_mutation,1,1), ".", str_sub(add_mutation$nt_mutation,-1,-1), sep="")
+```{r stacked_nucleotide, echo=FALSE}
+stacked_nt <- filtered_SARS_mutation
+stacked_nt$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
 
-# Stacked bar graph for the mutation counts normalized by the expected mutation rate for each clade
-ggplot(add_mutation, aes(fill=mutation_type, y=actual_count/expected_count, x=clade))+
+ggplot(stacked_nt, aes(x=clade, y=actual_count/expected_count, fill=mutation_type))+
   geom_bar(position="stack", stat="identity")+
+  scale_fill_brewer(palette="Paired")+
   theme_bw()+
@@
 
-[INSERT] information about the plot. Plot to visualize data.
+Amino acids occur at different proportions throughout the SARS-CoV-2 genome and between phylogenetic clades. [INSERT]
 
-```{r nucleotide_site, echo=FALSE}
-ggplot(add_mutation, aes(color=mutation_type, y=log(actual_count/expected_count), x=nt_site))+
-  geom_point()+
+```{r stacked_amino_acid, echo=FALSE}
+stacked_aa <- filtered_SARS_mutation
+stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
+stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)
+
+ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
+  geom_bar(position="stack", stat="identity")+
+  facet_wrap(~clade_founder, ncol=7)+
   theme_bw()+
-  labs(x="", y="")
+  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
 ```
@@
 filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
-
-# Remove columns that were already used to filter the data or than contain redundant information.
-filtered_clades$subset <- NULL
-filtered_clades$exclude <- NULL
-filtered_clades$nt_site <- NULL
 ```
@@
 # Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
-colnames(clade1) <- paste("c1", colnames(clade1), sep="_")
-colnames(clade2) <- paste("c2", colnames(clade2), sep="_")
-
-# Modify nt_mutation, clade_founder_nt and clade to make merging easier and make the data frame more readable.
-{colnames(clade1)[which(colnames(clade1)=="c1_nt_mutation")] <- "nt_mutation"
-colnames(clade2)[which(colnames(clade2)=="c2_nt_mutation")] <- "nt_mutation"
-colnames(clade1)[which(colnames(clade1)=="c1_clade_founder_nt")] <- "clade_founder_nt"
-colnames(clade2)[which(colnames(clade2)=="c2_clade_founder_nt")] <- "clade_founder_nt"
-colnames(clade1)[which(colnames(clade1)=="c1_clade")] <- "clade1"
-colnames(clade2)[which(colnames(clade2)=="c2_clade")] <- "clade2"}
+colnames(clade1)[colnames(clade1) %notin% merge_columns] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_columns], "1", sep="_")
+colnames(clade2)[colnames(clade2) %notin% merge_columns] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_columns], "2", sep="_")
 
 # Merge the two data frames by mutation so that each row represents one comparison.
-comparisons <- merge(clade1, clade2, by=c("nt_mutation","clade_founder_nt"))
+comparisons <- merge(clade1, clade2, by=merge_columns)
 ```
@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["c1_actual_count"]),as.numeric(c["c1_rounded_expected"])),c(as.numeric(c["c2_actual_count"]),as.numeric(c["c2_rounded_expected"]))))$p.value}))
+#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
 
 # Save and load variable because calculating p-values may take a long time.
-#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
-load(paste("pvalues", paste(clades, collapse="_"), ".Rda", sep=""))
+#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+load(paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
 ```
@@
 # Calculate fold change for plotting volcano plots.
-comparison_p_values$fold_change <- log2(comparison_p_values$c1_observed_expected/comparison_p_values$c2_observed_expected)
+comparison_p_values$fold_change <- log2(comparison_p_values$observed_expected_1/comparison_p_values$observed_expected_2)
 ```
@@
 
-```{r volcano_plot, echo=FALSE}
+```{r volcano_all, echo=FALSE}
 # Plot the uncorrected p-values.
@@
   theme_bw()
+```
 
+```{r volcano_corrected, echo=FALSE}
 # After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=nt_mutation, color=clade_founder_nt))+
+ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(paste(option, "mutation", sep="_")), color=str_sub(get(paste(option, "mutation", sep="_")),1,1)))+
   geom_point()+
@@
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
-  scale_color_brewer(palette="Set2")+
+  #scale_color_brewer(palette="Set2", name="mutation")+
   theme_bw()
@@
 verify_analyses <- comparison_p_values
-verify_analyses$synonymous <- ifelse(verify_analyses$c1_synonymous=="TRUE" & verify_analyses$c2_synonymous=="TRUE", "TRUE", "FALSE")
+verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="TRUE" & verify_analyses$synonymous_2=="TRUE", "TRUE", "FALSE")
 verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")

### 2022-11-11 notebooks/expected_and_actual_counts_ratio.Rmd
Clean up notebook and fix plotting errors

@@
 #install.packages("ggforce")
+#install.packages("RColorBrewer")
 #install.packages("stringr")
@@
 library(ggforce)
+library(RColorBrewer)
 library(stringr)
@@
 The purpose of this notebook is to generate a volcano plot comparing any two SARS-CoV-2 phylogenetic clades.
-Using the volcano plot, mutations at certain nucleotide sites that are undergoing very different directions or degrees of selection between the two clades can be identified.
+Using the volcano plot, nucleotide mutations or amino acid mutations at certain sites along the genome that are undergoing very different directions or degrees of selection between the two clades can be identified.
 Once this notebook has been finalized, sections of the code will be saved as scripts and run through a Snakemake pipeline, allowing all clade comparisons to be computed in parallel.
@@
 4. Restructure data such that there are columns for clade 1 variables and clade 2 variables.\
-5. Using contingency table analyses (e.g., Fisher's Exact test), calculate p-values for each mutation at every nucleotide site between the two clades.\
+5. Using contingency table analyses (e.g., Fisher's Exact test), calculate a p-value for each mutation at every nucleotide or amino acid site between the two clades.\
 6. Adjust p-values to account for the large number of hypothesis tests.\
 7. Using the ratio of observed counts to expected counts for each clade and a pseudocount, calculate log2 fold-change.\
-8. Generate a volcano plot with log2 fold-change on the X-axis and -log10 p-values on the Y-axis.\
-9. Verify the notebook is computing p-values correctly by generating plots with expected outcomes (e.g., synonymous mutations).\
+8. Generate a volcano plot with log2 fold-change on the x-axis and -log10 p-values on the y-axis.\
+9. Verify the notebook is computing p-values correctly by generating plots with expected outcomes (e.g., synonymous vs non-synonymous mutations).\
 
@@
 
-option <- "aa" # nt or aa
-merge_columns <- c("nt_mutation","nt_site","clade_founder_nt","gene","codon_position")
-merge_columns <- c("aa_mutation","gene")
+# Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
+option <- "nt"
+merge_column <- paste(option, "mutation", sep="_")
 ```
@@
 ```{r aa_mutation}
+if(option=='aa'){
   # Create a list with the starting nucleotide site for an amino acid for all rows in the data set.
@@
 
-# Create column with amino acid mutation (e.g., M.266.268.L)
+  # Create a column with amino acid mutation (e.g., M.266.268.L).
   SARS_mutation$aa_mutation <- paste(str_sub(SARS_mutation$clade_founder_aa,1,1), starting_nt, starting_nt+2, str_sub(SARS_mutation$mutant_aa,1,1), sep=".")
+}
+
+# NEED TO SUM TOGETHER DUPLICATE AMINO ACIDS
+# consider entire gene column
+# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
 ```
@@
 filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
+
+# Remove columns that were used to filter the data.
+filtered_SARS_mutation$subset <- NULL
+filtered_SARS_mutation$exclude <- NULL
 ```
 
-To compare mutations between phylogenetic clades, this notebook will plot volcano plots with -log10 p-value on the Y-axis and log2 fold-change on the X-axis.
+To compare mutations between phylogenetic clades, this notebook will plot volcano plots with -log10 p-value on the y-axis and log2 fold-change on the x-axis.
 Prior to filtering the data for contingency table analyses, calculate the ratio of observed to expected counts and add a pseudocount to prevent infinities when fold-change is calculated.
-Mutations with a small count (i.e., 1) are going to be affected by the pseudocount more than mutations with a large count (i.e., 100).
+Mutations with a small count (e.g., 1) are going to be affected by the pseudocount more than mutations with a large count (e.g., 100).
 
@@
 
-```{r clean_columns}
-# Remove columns that were already used to filter the data.
-filtered_SARS_mutation$subset <- NULL
-filtered_SARS_mutation$exclude <- NULL
-```
-
 ## Describe and Characterize the Data
@@
 
-Look at a summary of the data.\
+Look at a summary of the data:
+
 -What are the data types for each column?\
@@
 
-# What percentage of the data set is each type of mutation? (e.g., A -> T)
-mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
-(table(mutation_type)/length(mutation_type))*100
+# What percentage of the data set is each type of nucleotide mutation? (e.g., A -> T)
+nt_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
+(table(nt_type)/length(nt_type))*100
+
+# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
+aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,-1,-1), sep="")
+(table(aa_type)/length(aa_type))*100
 ```
@@
 
-The four nucleotides (e.g., A) and their possible mutations (e.g., A -> T) may be under different directions and degrees of selection between clades leading to an unequal distribution of nucleotides in the genome.
-Look at how many mutations (normalized by the expected count) there are for each clade.
+After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade.
+The number of mutations is normalized by the expected count.
 
@@
   geom_bar(position="stack", stat="identity")+
-  scale_fill_brewer(palette="Paired")+
+  scale_fill_brewer(palette="Paired", name="Nucleotide Mutation")+
   theme_bw()+
@@
 
-Amino acids occur at different proportions throughout the SARS-CoV-2 genome and between phylogenetic clades. [INSERT]
+In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to.
 
 ```{r stacked_amino_acid, echo=FALSE}
+stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)
+
 stacked_aa <- filtered_SARS_mutation
@@
   facet_wrap(~clade_founder, ncol=7)+
+  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
+  scale_x_discrete(guide=guide_axis(n.dodge=5))+
   theme_bw()+
@@
 
-# Append 'clade1' or 'clade2' to the column names so that the two data frames can be combined with easy to interpret columns.
-colnames(clade1)[colnames(clade1) %notin% merge_columns] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_columns], "1", sep="_")
-colnames(clade2)[colnames(clade2) %notin% merge_columns] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_columns], "2", sep="_")
+# Append '1' or '2' to the column names so that the two data frames can be easily combined.
+colnames(clade1)[colnames(clade1) %notin% merge_column] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_column], "1", sep="_")
+colnames(clade2)[colnames(clade2) %notin% merge_column] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_column], "2", sep="_")
 
 # Merge the two data frames by mutation so that each row represents one comparison.
-comparisons <- merge(clade1, clade2, by=merge_columns)
+comparisons <- merge(clade1, clade2, by=merge_column)
+
+# Determine if any of the columns from 'clade1' and 'clade2' are the same.
+col_names <- colnames(filtered_clades)[colnames(filtered_clades) %notin% merge_column]
+is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[, get(paste(c[1], 1, sep="_"))], comparisons[, get(paste(c[1], 2, sep="_"))])}))
+col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]
+
+# Rename one of the duplicate columns and remove the other.
+colnames(comparisons)[colnames(comparisons) %in% paste(col_equal, "1", sep="_")] <- col_equal
+comparisons[,paste(col_equal, "2", sep="_")] <- NULL
 ```
@@
 
-# Save and load variable because calculating p-values may take a long time.
+# Save and load the variable because calculating p-values may take a long time.
 #save(p_values, file=paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
@@
 
+By calculating a p-value for every nucleotide or amino acid mutation, comparing the two clades generates a lot of hypothesis tests.
+Account for this large number of hypothesis tests by correcting the p-values using a false discovery rate correction.
+
 ```{r p_value_correction}
-# Using False Discovery Rate Correction, adjust the p-values by the number of hypothesis tests.
+# Using a false discovery rate correction, adjust the p-values by the number of hypothesis tests.
 comparison_p_values$p_value_corrected <- p.adjust(comparison_p_values$p_value, method="fdr", n=nrow(comparison_p_values))
@@
 ```{r fold-change}
-# Calculate fold change for plotting volcano plots.
+# Calculate fold-change for plotting volcano plots.
 comparison_p_values$fold_change <- log2(comparison_p_values$observed_expected_1/comparison_p_values$observed_expected_2)
@@
 
-Generate a volcano plot with -log10 p-values on the Y-axis and log2 fold-change on the X-axis.
+Generate volcano plots with -log10 p-values on the y-axis and log2 fold-change on the x-axis:
+
+The first volcano plot uses the uncorrected p-values, and the red line indicates a significance value of P = 0.05.
 
@@
 
+After correcting the p-values, a large portion of the p-values equal one.
+The second volcano plot drops P = 1 and uses the remaining corrected p-values.
+Points are colored by the starting nucleotide or amino acid mutation (i.e., clade founder) and labels indicate the mutation.
+The red line indicates a significance value of P = 0.05.
+
 ```{r volcano_corrected, echo=FALSE}
-# After dropping p = 1, plot FDR corrected p-values (color by nucleotide clade founder and label by mutation).
+# After dropping P = 1, plot the corrected p-values.
+color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(comparison_p_values[,str_sub(get(paste(option, "mutation", sep="_")),1,1)])))
+
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
@@
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
-  #scale_color_brewer(palette="Set2", name="mutation")+
+  scale_color_manual(values=color, name="mutation")+
   theme_bw()
@@
 Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-values close to 0), there should be insignificant differences between clades for synonymous mutations (i.e., p-values close to 1).
+The black line indicates a significance value of P = 0.05.
 
-```{r verify_analyses}
-# Plot the corrected p-values for synonymous and non-synonymous mutations and compare to a significance value of 0.05.
+```{r verify_analyses, echo=FALSE}
+# Plot the corrected p-values for synonymous and non-synonymous mutations.
 verify_analyses <- comparison_p_values
+if("synonymous" %notin% colnames(verify_analyses)){
   verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="TRUE" & verify_analyses$synonymous_2=="TRUE", "TRUE", "FALSE")
+}
 verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
 
-p <- ggplot(verify_analyses, aes(x=p_value_corrected, fill=synonymous, label=..count..))+
+p <- ggplot(verify_analyses, aes(x=p_value_corrected, fill=synonymous, label=after_stat(count)))+
   geom_histogram()+
@@
   scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1))+
-  labs(x="FDR corrected p-values")+
+  labs(x="corrected p-values")+
   theme_bw()

### 2022-11-14 notebooks/expected_and_actual_counts_ratio.Rmd
Pull updated CSV file and modify amino acid analyses to use updated file

@@
 
-#install.packages("data.table")
+#install.packages("dplyr")
 #install.packages("ggplot2")
@@
 
-library(data.table)
+library(dplyr)
 library(ggplot2)
@@
 # Read data using access to the Fred Hutch server.
-SARS_mutation <- fread("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
+SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
 
 # Read data from github.com.
-#SARS_mutation <- fread("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
+#SARS_mutation <- read.csv("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
 ```
@@
 
-With the data set, both nucleotide and amino acid mutations can be compared between clades.
-However, there is no column identifying the amino acid mutation (i.e., starting amino acid, codon position along the genome and mutated amino acid).
-Create a column with this information using the nucleotide site, codon position (i.e., 1, 2, or 3), clade founder amino acid and mutant amino acid columns.
-
-```{r aa_mutation}
-if(option=='aa'){
-  # Create a list with the starting nucleotide site for an amino acid for all rows in the data set.
-  starting_nt <- SARS_mutation$nt_site-as.numeric(str_sub(SARS_mutation$codon_position,1,1))+1
-
-  # Create a column with amino acid mutation (e.g., M.266.268.L).
-  SARS_mutation$aa_mutation <- paste(str_sub(SARS_mutation$clade_founder_aa,1,1), starting_nt, starting_nt+2, str_sub(SARS_mutation$mutant_aa,1,1), sep=".")
-}
-
-# NEED TO SUM TOGETHER DUPLICATE AMINO ACIDS
-# consider entire gene column
-# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
-```
-
 For the analyses conducted in this notebook, subset the data by the country (countries) the sequence data came from.
@@
 # Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
-filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="FALSE" & expected_count>=min_expected)
+filtered_SARS_mutation <- subset(SARS_mutation, subset==country & exclude=="False" & expected_count>=min_expected)
+```
 
-# Remove columns that were used to filter the data.
-filtered_SARS_mutation$subset <- NULL
-filtered_SARS_mutation$exclude <- NULL
+```{r aa_mutation}
+if(option=='aa'){
+  aa_group <- filtered_SARS_mutation
+  filtered_SARS_mutation <- as.data.frame(aa_group %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
+}else{
+  nt_group <- filtered_SARS_mutation
+  filtered_SARS_mutation <- nt_group[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","actual_count","expected_count")]
+}
+
+# consider entire gene column
+# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
 ```
@@
 # What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
-aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,-1,-1), sep="")
+aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,1,1), sep="")
+aa_type <- paste(filtered_SARS_mutation$clade_founder_aa, ".", filtered_SARS_mutation$mutant_aa, sep="")
 (table(aa_type)/length(aa_type))*100
@@
 ```{r stacked_amino_acid, echo=FALSE}
-stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)
+stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(79)
 
 stacked_aa <- filtered_SARS_mutation
-stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
-stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)
+#stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
+#stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)
 
-ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
+ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
   geom_bar(position="stack", stat="identity")+
-  facet_wrap(~clade_founder, ncol=7)+
+  facet_wrap(~clade_founder_aa, ncol=7)+
   scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
@@
 
-With such a large data set, it could take hours to calculate the p-values for all clade comparisons.
-To decrease the amount of time it takes to conduct analyses, this notebook will only compare two clades at a time and the script can be run in parallel to generate all comparisons at once.
-Therefore, filter the data to only include mutation data from the two clades being compared.
-
-```{r subset_by_clade}
-# Create a subset of the data using only the clades that are going to be compared.
-filtered_clades <- filtered_SARS_mutation[filtered_SARS_mutation$clade %in% clades,]
-```
-
 To quickly calculate the p-value for each contingency table analysis and ensure the user can easily understand what is being compared in each hypothesis test, restructure the data such that all of the variables necessary to run the statistical test are in one row.
@@
 # Create two data frames with data for clade 1 and clade 2.
-clade1 <- filtered_clades[which(filtered_clades$clade==clades[1]),]
-clade2 <- filtered_clades[which(filtered_clades$clade==clades[2]),]
+clade1 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[1]),]
+clade2 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[2]),]
 
@@
 # Determine if any of the columns from 'clade1' and 'clade2' are the same.
-col_names <- colnames(filtered_clades)[colnames(filtered_clades) %notin% merge_column]
-is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[, get(paste(c[1], 1, sep="_"))], comparisons[, get(paste(c[1], 2, sep="_"))])}))
+col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% merge_column]
+is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
 col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]
@@
 # After dropping P = 1, plot the corrected p-values.
-color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(comparison_p_values[,str_sub(get(paste(option, "mutation", sep="_")),1,1)])))
+color <- colorRampPalette(brewer.pal(8,"Set1"))(comparison_p_values[,paste("clade_founder", option, sep="_")] %>% unique() %>% length())
 
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(paste(option, "mutation", sep="_")), color=str_sub(get(paste(option, "mutation", sep="_")),1,1)))+
+ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=get(paste("clade_founder", option, sep="_"))))+
   geom_point()+
@@
 if("synonymous" %notin% colnames(verify_analyses)){
-  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="TRUE" & verify_analyses$synonymous_2=="TRUE", "TRUE", "FALSE")
+  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="True" & verify_analyses$synonymous_2=="True", "True", "False")
 }

### 2022-11-15 notebooks/expected_and_actual_counts_ratio.Rmd
Group by amino acid and add counts of identical mutations

@@
 # Identify the two clades that will be included in comparisons.
-clades <- c("20A","22B")
+clades <- c("22A","22B")
 
@@
 # Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
-option <- "nt"
+option <- "aa"
 merge_column <- paste(option, "mutation", sep="_")
@@
 
-## Filter and Prepare the SARS-CoV-2 Mutation Data
+## Filter the SARS-CoV-2 Mutation Data
 
@@
 
-```{r aa_mutation}
-if(option=='aa'){
-  aa_group <- filtered_SARS_mutation
-  filtered_SARS_mutation <- as.data.frame(aa_group %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
-}else{
-  nt_group <- filtered_SARS_mutation
-  filtered_SARS_mutation <- nt_group[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","actual_count","expected_count")]
-}
+Create nucleotide mutation and amino acid mutation data sets.
+Actual and expected counts for rows with identical amino acid mutations are added together.
+Both data sets only include columns relevant for that mutation to make understanding restructuring the data and analyses easier.
+
+```{r group_mutation}
+# Group by amino acid mutation and add counts for identical amino acid mutations.
+# NOTE: Grouping with and without 'gene' creates a different number of rows because some equivalent aa mutations belong to different genes. There are 2,160 aa mutations that have duplicates.
+aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
+
+# Reduce columns for nucleotide mutation analyses to make workflow easier to follow.
+nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
 
@@
 # Should I compare mutations with different wild types? (A200T vs C200T) Plot?
+```
+
+## Describe and Characterize the Data
+
+Take a look at the first 5 rows of the data.
+
+```{r head_data}
+head(get(paste(option, "group", sep="_")), n=5)
+```
+
+Look at a summary of the data:
+
+-What are the data types for each column?\
+-Is there a similar number of TRUEs and FALSEs for logical variables?\
+-Describe the numeric variables (e.g., minimum value).\
+
+```{r summary_data}
+summary(get(paste(option, "group", sep="_")))
+```
+
+Take a closer look at how many nucleotide sites, amino acid sites and the proportion of mutations in the data set.
+
+```{r describe_data}
+# How many nucleotide sites?
+length(unique(nt_group$nt_site))
+
+# How many amino acid sites?
+length(unique(aa_group$codon_site))
+
+# What percentage of the data set is each type of nucleotide mutation? (e.g., A -> T)
+nt_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")
+(table(nt_type)/length(nt_type))*100
+
+# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
+aa_type <- paste(aa_group$clade_founder_aa, ".", aa_group$mutant_aa, sep="")
+(table(aa_type)/length(aa_type))*100
+```
+
+## Visualize the Data
+
+After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade.
+The number of mutations is normalized by the expected count.
+
+```{r stacked_nucleotide, echo=FALSE}
+stacked_nt <- nt_group
+stacked_nt$mutation_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")
+
+ggplot(stacked_nt, aes(x=clade, y=actual_count/expected_count, fill=mutation_type))+
+  geom_bar(position="stack", stat="identity")+
+  scale_fill_brewer(palette="Paired", name="Nucleotide Mutation")+
+  theme_bw()+
+  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
+```
+
+In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to.
+
+```{r stacked_amino_acid, echo=FALSE}
+stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)
+
+stacked_aa <- aa_group
+stacked_aa$clade_founder <- str_sub(aa_group$clade_founder_aa,1,1)
+stacked_aa$mutant <- str_sub(aa_group$mutant_aa,1,1)
+
+ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
+  geom_bar(position="stack", stat="identity")+
+  facet_wrap(~clade_founder, ncol=7)+
+  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
+  scale_x_discrete(guide=guide_axis(n.dodge=5))+
+  theme_bw()+
+  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
+```
+
+## Prepare and Restructure Data for Analyses
+
+```{r rename_variable}
+# Rename variable for analyses to make code more readable.
+filtered_SARS_mutation <- get(paste(option, "group", sep="_"))
 ```
@@
 ```
-
-## Describe and Characterize the Data
-
-Take a look at the first 5 rows of the data.
-
-```{r head_data}
-head(filtered_SARS_mutation, n=5)
-```
-
-Look at a summary of the data:
-
--What are the data types for each column?\
--Is there a similar number of TRUEs and FALSEs for logical variables?\
--Describe the numeric variables (e.g., minimum value).\
-
-```{r summary_data}
-summary(filtered_SARS_mutation)
-```
-
-Take a closer look at how many nucleotide sites and the proportion of mutations in the data set.
-
-```{r describe_data}
-# How many nucleotide sites?
-length(unique(filtered_SARS_mutation$nt_site))
-
-# What percentage of the data set is each type of nucleotide mutation? (e.g., A -> T)
-nt_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
-(table(nt_type)/length(nt_type))*100
-
-# What percentage of the data set is each type of amino acid mutation? (e.g., K -> M)
-aa_type <- paste(str_sub(filtered_SARS_mutation$clade_founder_aa,1,1), ".", str_sub(filtered_SARS_mutation$mutant_aa,1,1), sep="")
-aa_type <- paste(filtered_SARS_mutation$clade_founder_aa, ".", filtered_SARS_mutation$mutant_aa, sep="")
-(table(aa_type)/length(aa_type))*100
-```
-
-## Visualize the Data
-
-After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade.
-The number of mutations is normalized by the expected count.
-
-```{r stacked_nucleotide, echo=FALSE}
-stacked_nt <- filtered_SARS_mutation
-stacked_nt$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")
-
-ggplot(stacked_nt, aes(x=clade, y=actual_count/expected_count, fill=mutation_type))+
-  geom_bar(position="stack", stat="identity")+
-  scale_fill_brewer(palette="Paired", name="Nucleotide Mutation")+
-  theme_bw()+
-  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
-```
-
-In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to.
-
-```{r stacked_amino_acid, echo=FALSE}
-stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(79)
-
-stacked_aa <- filtered_SARS_mutation
-#stacked_aa$clade_founder <- str_sub(filtered_SARS_mutation$clade_founder_aa,1,1)
-#stacked_aa$mutant <- str_sub(filtered_SARS_mutation$mutant_aa,1,1)
-
-ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
-  geom_bar(position="stack", stat="identity")+
-  facet_wrap(~clade_founder_aa, ncol=7)+
-  scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
-  scale_x_discrete(guide=guide_axis(n.dodge=5))+
-  theme_bw()+
-  labs(x="Phylogenetic Clade", y="Normalized Mutation Count")
-```
-
-## Restructure Data for Analyses
 
@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
+p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
 
 # Save and load the variable because calculating p-values may take a long time.
-#save(p_values, file=paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
-load(paste("pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+#save(p_values, file=paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+load(paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
 ```
@@
 # After dropping P = 1, plot the corrected p-values.
-color <- colorRampPalette(brewer.pal(8,"Set1"))(comparison_p_values[,paste("clade_founder", option, sep="_")] %>% unique() %>% length())
+volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
+if("gene" %notin% colnames(volcano)){
+  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep=";"))
+}
+color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
 
-volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=get(paste("clade_founder", option, sep="_"))))+
+p <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
   geom_point()+
@@
   theme_bw()
+
+p + theme(legend.position="none")
+```
+
+```{r volcano_corrected_legend}
+p + theme(legend.position="bottom")
 ```

### 2022-11-15 expected_and_actual_counts_ratio/scripts/filter_clade_data.R
Create script that will write CSV files for clade comparisons

@@ -1,0 +1,24 @@
+library(dplyr)
+
+# Read data using access to the Fred Hutch server.
+SARS_mutation <- read.csv(snakemake@input[[1]], sep=",", header=TRUE)
+
+# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
+filtered_SARS_mutation <- subset(SARS_mutation, subset==snakemake@input[[2]] & exclude=="False" & expected_count>=snakemake@input[[3]])
+
+# Make a list of the unique clades in the dataset.
+clades <- unique(filtered_SARS_mutation$clade)
+
+## NUCLEOTIDE MUTATION
+# Filter columns to make workflow easier to interpret.
+nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
+
+# Save individual CSV files for each clade with the filtered data.
+sapply(clades, FUN=function(c){ write.csv(nt_group[which(nt_group$clade==c[1]),], file=paste("results/", c[1], "_nt.csv", sep="")) })
+
+## AMINO ACID MUTATION
+# Group by amino acid mutation and add counts for identical amino acid mutations.
+aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
+
+# Save individual CSV files for each clade with the filtered data.
+sapply(clades, FUN=function(c){ write.csv(aa_group[which(aa_group$clade==c[1]),], file=paste("results/", c[1], "_aa.csv", sep="")) })

### 2022-11-16 expected_and_actual_counts_ratio/scripts/calculate_p_values.R
Create script for calculating p-values

@@ -1,0 +1,68 @@
+library(dplyr)
+
+# Read data using access to the Fred Hutch server.
+SARS_mutation <- read.csv(snakemake@input[[1]], sep=",", header=TRUE)
+
+# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
+filtered_SARS_mutation <- subset(SARS_mutation, subset==snakemake@params@subset & exclude=="False" & expected_count>=snakemake@params@min_expected)
+
+# Make a list of the unique clades in the dataset.
+clades <- unique(filtered_SARS_mutation$clade)
+
+# Filter columns to make workflow easier to interpret.
+if(snakemake@wildcards[["nt"]]=="nt"){
+   nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
+}
+
+# Group by amino acid mutation and add counts for identical amino acid mutations.
+if(snakemake@wildcards[["aa"]]=="aa"){
+   aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
+}
+
+#
+filtered_SARS_mutation <-  
+
+
+
+
+
+
+
+
+# Calculate the ratio of actual counts (i.e., observed) to expected counts (i.e., expected) with a pseudocount.
+filtered_SARS_mutation$observed_expected <- (filtered_SARS_mutation$actual_count + snakemake@input[[2]])/(filtered_SARS_mutation$expected_count + snakemake@input[[2]])
+
+# Round expected counts because contingency table analyses (e.g., Fisher's Exact test) can only analyze integers.
+filtered_SARS_mutation$rounded_expected <- round(filtered_SARS_mutation$expected_count)
+
+# Create two data frames with data for clade 1 and clade 2.
+clade1 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[1]),]
+clade2 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[2]),]
+
+# Append '1' or '2' to the column names so that the two data frames can be easily combined.
[…]

### 2022-11-17 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Modify plots describing the data to only use spike mutations and fix coloring of volcano plots

--- 2022-11-15 notebooks/expected_and_actual_counts_ratio.Rmd
@@
 # Read data using access to the Fred Hutch server.
-SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
+SARS_mutation <- read.csv("../../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", sep=",", header=TRUE)
 
@@
 # Identify the two clades that will be included in comparisons.
-clades <- c("22A","22B")
+clades <- c("20A","22B")
 
@@
 
-# consider entire gene column
 # Should I compare mutations with different wild types? (A200T vs C200T) Plot?
@@
 The number of mutations is normalized by the expected count.
+To simplify the plot, only the nucleotide mutations in the spike protein are plotted.
+Previous studies have shown that this protein is important for SARS-CoV-2 evolution.
 
 ```{r stacked_nucleotide, echo=FALSE}
-stacked_nt <- nt_group
-stacked_nt$mutation_type <- paste(str_sub(nt_group$nt_mutation,1,1), ".", str_sub(nt_group$nt_mutation,-1,-1), sep="")
+stacked_nt <- nt_group[which(nt_group$gene=="S"),]
+stacked_nt$mutation_type <- paste(str_sub(stacked_nt$nt_mutation,1,1), ".", str_sub(stacked_nt$nt_mutation,-1,-1), sep="")
 
@@
 In the following amino acid mutation plot, individual plots indicate the starting amino acid (i.e., clade founder) and stacked bars within each plot indicate what that amino acid mutated to.
+Again, only the amino acid mutations in the spike protein are plotted.
 
@@
 stacked_col <- colorRampPalette(brewer.pal(8,"Set1"))(21)
+stacked_aa <- aa_group[which(aa_group$gene=="S"),]
 
-stacked_aa <- aa_group
-stacked_aa$clade_founder <- str_sub(aa_group$clade_founder_aa,1,1)
-stacked_aa$mutant <- str_sub(aa_group$mutant_aa,1,1)
-
-ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant))+
+ggplot(stacked_aa, aes(x=clade, y=actual_count/expected_count, fill=mutant_aa))+
   geom_bar(position="stack", stat="identity")+
-  facet_wrap(~clade_founder, ncol=7)+
+  facet_wrap(~clade_founder_aa, ncol=5)+
   scale_fill_manual(values=stacked_col, name="Amino Acid Mutation")+
@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
+#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
 
 # Save and load the variable because calculating p-values may take a long time.
-#save(p_values, file=paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
-load(paste("../results/expected_and_actual_counts_ratio/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+#save(p_values, file=paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+load(paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
 ```
@@
 
-```{r volcano_corrected, echo=FALSE}
+```{r volcano_corrected, echo=FALSE, fig.height=10}
 # After dropping P = 1, plot the corrected p-values.
@@
 if("gene" %notin% colnames(volcano)){
-  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep=";"))
+  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
 }
@@
 
-p <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
+ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
+  geom_point()+
+  geom_text(color="black", size=2, nudge_y=0.2)+
+  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
+  scale_color_manual(values=color, name="mutation")+
+  theme_bw()+
+  theme(legend.position="bottom")
+```
+
+Plot the same volcano plot but color only the points where the mutation is in the spike protein for one or both clades.
+
+```{r volcano_corrected_spike, echo=FALSE}
+spike <- volcano$gene[grep("S", volcano$gene)]
+volcano$spike <- ifelse(volcano$gene %in% spike, volcano$gene, "non-spike")
+non_spike <- which(levels(as.factor(volcano$spike))=="non-spike")
+color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$spike)))
+color[non_spike] <- "grey"
+
+ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=spike))+
   geom_point()+
@@
   theme_bw()
-
-p + theme(legend.position="none")
-```
-
-```{r volcano_corrected_legend}
-p + theme(legend.position="bottom")
 ```

### 2022-11-18 expected_and_actual_counts_ratio/scripts/calculate_p_values.R
Fix snakemake object calls and verify all commands of script work

@@ -1,36 +1,35 @@
 library(dplyr)
+library(stringr)
+
+`%notin%` <- Negate(`%in%`)
+
+# Create the name of the column with the mutations.
+merge_column <- paste(snakemake@wildcards[["mutation_type"]], "mutation", sep="_")
+
+# Identify the clades that are going to be compared.
+clades <- str_split(snakemake@wildcards[["comparisons"]], "_", 2)[[1]]
 
 # Read data using access to the Fred Hutch server.
 SARS_mutation <- read.csv(snakemake@input[[1]], sep=",", header=TRUE)
 
 # Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
-filtered_SARS_mutation <- subset(SARS_mutation, subset==snakemake@params@subset & exclude=="False" & expected_count>=snakemake@params@min_expected)
-
-# Make a list of the unique clades in the dataset.
-clades <- unique(filtered_SARS_mutation$clade)
+filtered_SARS_mutation <- subset(SARS_mutation, subset==snakemake@params[["subset"]] & exclude=="False" & expected_count>=snakemake@params[["min_expected"]] & clade %in% clades)
 
 # Filter columns to make workflow easier to interpret.
-if(snakemake@wildcards[["nt"]]=="nt"){
-   nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
+if(snakemake@wildcards[["mutation_type"]]=="nt"){
+   temp <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
 }
 
 # Group by amino acid mutation and add counts for identical amino acid mutations.
-if(snakemake@wildcards[["aa"]]=="aa"){
-   aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
+if(snakemake@wildcards[["mutation_type"]]=="aa"){
[…]

### 2022-11-21 expected_and_actual_counts_ratio/scripts/volcano_plot.R
Add script to generate PDF plots for nt and aa mutations

@@ -1,0 +1,53 @@
+library(ggplot2)
+library(RColorBrewer)
+
+`%notin%` <- Negate(`%in%`)
+
+# Specify column for plotting mutation data (nt_mutation or aa_mutation).
+merge_column <- paste(snakemake@wildcards[["mutation_type"]], "mutation", sep="_")
+
+# Create filename for plotting mutation data from a two-clade comparison.
+input_name <- paste("results/pvalues", snakemake@wildcards[["comparisons"]], "_", snakemake@wildcards[["mutation_type"]], ".csv", sep="")
+
+# Read in the data for plotting.
+comparison_p_values <- read.csv(input_name, sep=",", header=TRUE)
+
+## After dropping P = 1, plot the corrected p-values. ##
+volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
+if("gene" %notin% colnames(volcano)){
+  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
+}
+
+if(merge_column == "nt_mutation"){
+	color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
+
+	plot <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
+  	  geom_point()+
+  	  geom_text(color="black", size=2, nudge_y=0.2)+
+  	  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
+  	  scale_color_manual(values=color, name="mutation")+
+  	  theme_bw()+
+  	  theme(legend.position="bottom")
+}
+
+## Aggregate all mutations that do not include the spike protein to 'non-spike'.
+if(merge_column == "aa_mutation"){
+	spike <- volcano$gene[grep("S", volcano$gene)]
+	volcano$spike <- ifelse(volcano$gene %in% spike, volcano$gene, "non-spike")
+	non_spike <- which(levels(as.factor(volcano$spike))=="non-spike")
+	color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$spike)))
+	color[non_spike] <- "grey"
+
+	plot <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=spike))+
+	  geom_point()+
[…]

### 2022-11-21 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Modify while test running notebook

@@
 # Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-#p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
+p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
 
@@
 #save(p_values, file=paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
-load(paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
+#load(paste("../results/pvalues", paste(clades, collapse="_"), option, ".Rda", sep=""))
 ```
@@
 # Plot the uncorrected p-values.
-ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+p <- ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
   geom_point(alpha=0.5)+
@@
 
-```{r volcano_corrected, echo=FALSE, fig.height=10}
+```{r volcano_corrected, echo=FALSE}
+#fig.height=10?
 # After dropping P = 1, plot the corrected p-values.

### 2022-11-22 expected_and_actual_counts_ratio/scripts/calculate_p_values.R
Merge with both mutation and gene

@@ -39,14 +39,14 @@ clade1 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[1]),
 clade2 <- filtered_SARS_mutation[which(filtered_SARS_mutation$clade==clades[2]),]
 
 # Append '1' or '2' to the column names so that the two data frames can be easily combined.
-colnames(clade1)[colnames(clade1) %notin% merge_column] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_column], "1", sep="_")
-colnames(clade2)[colnames(clade2) %notin% merge_column] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_column], "2", sep="_")
+colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")] <- paste(colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")], "1", sep="_")
+colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")] <- paste(colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")], "2", sep="_")
 
 # Merge the two data frames by mutation so that each row represents one comparison.
-comparisons <- merge(clade1, clade2, by=merge_column)
+comparisons <- merge(clade1, clade2, by=c(merge_column, "gene"))
 
 # Determine if any of the columns from 'clade1' and 'clade2' are the same.
-col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% merge_column]
+col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% c(merge_column, "gene")]
 is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
 col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]
 

### 2022-11-22 expected_and_actual_counts_ratio/scripts/volcano_plot.R
Fix plotting for readability

@@ -1,5 +1,6 @@
 library(ggplot2)
 library(RColorBrewer)
+library(grid)
 
 `%notin%` <- Negate(`%in%`)
 
@@ -14,40 +15,32 @@ comparison_p_values <- read.csv(input_name, sep=",", header=TRUE)
 
 ## After dropping P = 1, plot the corrected p-values. ##
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-if("gene" %notin% colnames(volcano)){
-  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
-}
+color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
 
-if(merge_column == "nt_mutation"){
-	color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
+neg_clade <- textGrob(unique(volcano$clade_2), gp=gpar(fontsize=13, fontface="bold"))
+pos_clade <- textGrob(unique(volcano$clade_1), gp=gpar(fontsize=13, fontface="bold"))
 
-	plot <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
-  	  geom_point()+
-  	  geom_text(color="black", size=2, nudge_y=0.2)+
-  	  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
-  	  scale_color_manual(values=color, name="mutation")+
-  	  theme_bw()+
-  	  theme(legend.position="bottom")
-}
+na <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
+  geom_point()+
+  geom_text(color="black", size=2, nudge_y=0.2)+
+  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
+  scale_color_manual(values=color, name="Mutation")+
+  theme_bw()+
+  labs(x="Log2 Fold Change", y="-log10 FDR Corrected P-Value")
 
-## Aggregate all mutations that do not include the spike protein to 'non-spike'.
-if(merge_column == "aa_mutation"){
-	spike <- volcano$gene[grep("S", volcano$gene)]
[…]

### 2022-11-22 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Fix merging for amino acid mutations and volcano plots

@@
 
-#install.packages("dplyr")
-#install.packages("ggplot2")
-#install.packages("ggforce")
-#install.packages("RColorBrewer")
-#install.packages("stringr")
-#install.packages("gtools")
-
 library(dplyr)
@@
 library(gtools)
+library(grid)
 
@@
 # Identify the two clades that will be included in comparisons.
-clades <- c("20A","22B")
+clades <- c("20A","22A")
 
@@
 # Analyze nucleotide mutations or amino acid mutations ('nt' or 'aa')?
-option <- "aa"
+option <- "nt"
 merge_column <- paste(option, "mutation", sep="_")
@@
 # Append '1' or '2' to the column names so that the two data frames can be easily combined.
-colnames(clade1)[colnames(clade1) %notin% merge_column] <- paste(colnames(clade1)[colnames(clade1) %notin% merge_column], "1", sep="_")
-colnames(clade2)[colnames(clade2) %notin% merge_column] <- paste(colnames(clade2)[colnames(clade2) %notin% merge_column], "2", sep="_")
+colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")] <- paste(colnames(clade1)[colnames(clade1) %notin% c(merge_column, "gene")], "1", sep="_")
+colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")] <- paste(colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")], "2", sep="_")
 
 # Merge the two data frames by mutation so that each row represents one comparison.
-comparisons <- merge(clade1, clade2, by=merge_column)
+comparisons <- merge(clade1, clade2, by=c(merge_column, "gene"))
 
 # Determine if any of the columns from 'clade1' and 'clade2' are the same.
-col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% merge_column]
+col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% c(merge_column, "gene")]
 is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
@@
 # Plot the uncorrected p-values.
-p <- ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
+ggplot(comparison_p_values, aes(x=fold_change, y=-log10(p_value)))+
   geom_point(alpha=0.5)+
@@
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-if("gene" %notin% colnames(volcano)){
-  volcano$gene <- ifelse(volcano$gene_1==volcano$gene_2, volcano$gene_1, paste(volcano$gene_1, volcano$gene_2, sep="_"))
-}
 color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
 
-ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
+neg_clade <- textGrob(clades[2], gp=gpar(fontsize=13, fontface="bold"))
+pos_clade <- textGrob(clades[1], gp=gpar(fontsize=13, fontface="bold"))
+
+na <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
   geom_point()+
@@
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
-  scale_color_manual(values=color, name="mutation")+
+  scale_color_manual(values=color, name="Mutation")+
   theme_bw()+
-  theme(legend.position="bottom")
-```
+  labs(x="Log2 Fold Change", y="-log10 FDR Corrected P-Value")
 
-Plot the same volcano plot but color only the points where the mutation is in the spike protein for one or both clades.
+x <- layer_scales(na)$x$range$range
+y <- layer_scales(na)$y$range$range
 
-```{r volcano_corrected_spike, echo=FALSE}
-spike <- volcano$gene[grep("S", volcano$gene)]
-volcano$spike <- ifelse(volcano$gene %in% spike, volcano$gene, "non-spike")
-non_spike <- which(levels(as.factor(volcano$spike))=="non-spike")
-color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$spike)))
-color[non_spike] <- "grey"
+if(is.null(x)){
+  na
+} else {
+  p <- na + coord_cartesian(xlim=c(-max(abs(x)), max(abs(x))))
+  x_mod <- p$coordinates$limits[[1]]
 
-ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=spike))+
-  geom_point()+
-  geom_text(color="black", size=2, nudge_y=0.2)+
-  geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
-  scale_color_manual(values=color, name="mutation")+
-  theme_bw()
+  p + annotation_custom(neg_clade,xmin=x_mod[1]*0.7,xmax=x_mod[1]*0.7,ymin=y[2]*0.95,ymax=y[2]*0.95)+ 
+      annotation_custom(pos_clade,xmin=x_mod[2]*0.7,xmax=x_mod[2]*0.7,ymin=y[2]*0.95,ymax=y[2]*0.95)
+}
 ```
@@
 # Plot the corrected p-values for synonymous and non-synonymous mutations.
+# PLOT UNCORRECTED P-VALUES
 verify_analyses <- comparison_p_values

### 2022-11-22 expected_and_actual_counts_ratio/scripts/filter_clade_data.R
Remove script not needed for Snakefile

@@ -1,24 +1,0 @@
-library(dplyr)
-
-# Read data using access to the Fred Hutch server.
-SARS_mutation <- read.csv(snakemake@input[[1]], sep=",", header=TRUE)
-
-# Include data from certain countries AND exclude branches with very high mutation rates and questionable mutations.
-filtered_SARS_mutation <- subset(SARS_mutation, subset==snakemake@input[[2]] & exclude=="False" & expected_count>=snakemake@input[[3]])
-
-# Make a list of the unique clades in the dataset.
-clades <- unique(filtered_SARS_mutation$clade)
-
-## NUCLEOTIDE MUTATION
-# Filter columns to make workflow easier to interpret.
-nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
-
-# Save individual CSV files for each clade with the filtered data.
-sapply(clades, FUN=function(c){ write.csv(nt_group[which(nt_group$clade==c[1]),], file=paste("results/", c[1], "_nt.csv", sep="")) })
-
-## AMINO ACID MUTATION
-# Group by amino acid mutation and add counts for identical amino acid mutations.
-aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
-
-# Save individual CSV files for each clade with the filtered data.
-sapply(clades, FUN=function(c){ write.csv(aa_group[which(aa_group$clade==c[1]),], file=paste("results/", c[1], "_aa.csv", sep="")) })

### 2022-11-23 expected_and_actual_counts_ratio/scripts/volcano_plot.R
Decrease distant between points and labels, fix plot dimensions and assign colors to genes

@@ -1,8 +1,12 @@
+library(stringr)
 library(ggplot2)
 library(RColorBrewer)
 library(grid)
 
 `%notin%` <- Negate(`%in%`)
+
+# Determine what the clades being compared are.
+clades <- str_split(snakemake@wildcards[["comparisons"]], "_", 2)[[1]]
 
 # Specify column for plotting mutation data (nt_mutation or aa_mutation).
 merge_column <- paste(snakemake@wildcards[["mutation_type"]], "mutation", sep="_")
@@ -14,15 +18,16 @@ input_name <- paste("results/pvalues", snakemake@wildcards[["comparisons"]], "_"
 comparison_p_values <- read.csv(input_name, sep=",", header=TRUE)
 
 ## After dropping P = 1, plot the corrected p-values. ##
-volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
 
-neg_clade <- textGrob(unique(volcano$clade_2), gp=gpar(fontsize=13, fontface="bold"))
-pos_clade <- textGrob(unique(volcano$clade_1), gp=gpar(fontsize=13, fontface="bold"))
+all_colors <- data.frame(color=colorRampPalette(brewer.pal(8,"Set1"))(length(unique(comparison_p_values$gene))), gene=levels(as.factor(comparison_p_values$gene)))
+volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
+color <- all_colors[all_colors$gene %in% unique(volcano$gene), "color"]
+
+neg_clade <- textGrob(clades[2], gp=gpar(fontsize=20, fontface="bold"))
+pos_clade <- textGrob(clades[1], gp=gpar(fontsize=20, fontface="bold"))
 
 na <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(merge_column), color=gene))+
   geom_point()+
-  geom_text(color="black", size=2, nudge_y=0.2)+
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
   scale_color_manual(values=color, name="Mutation")+
   theme_bw()+
[…]

### 2022-11-23 expected_and_actual_counts_ratio/scripts/p_distribution.R
Create script to plot uncorrected p-value distributions

@@ -1,0 +1,30 @@
+library(ggplot2)
+library(ggforce)
+
+`%notin%` <- Negate(`%in%`)
+
+# Create filename for plotting mutation data from a two-clade comparison.
+input_name <- paste("results/pvalues", snakemake@wildcards[["comparisons"]], "_", snakemake@wildcards[["mutation_type"]], ".csv", sep="")
+
+# Read in the data for plotting.
+verify_analyses <- read.csv(input_name, sep=",", header=TRUE)
+
+# Plot the corrected p-values for synonymous and non-synonymous mutations.
+if("synonymous" %notin% colnames(verify_analyses)){
+  verify_analyses$synonymous <- ifelse(verify_analyses$synonymous_1=="True" & verify_analyses$synonymous_2=="True", "True", "False")
+}
+verify_analyses$p_one <- ifelse(verify_analyses$p_value==1, "P=1", "P<1")
+
+p <- ggplot(verify_analyses, aes(x=p_value, fill=synonymous, label=after_stat(count)))+
+  geom_histogram()+
+  geom_text(size=2, vjust=-0.5, stat="bin", position="stack")+
+  geom_vline(data=data.frame(xint=0.05, p_one="P<1"), aes(xintercept=xint), linetype=2)+
+  scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1))+
+  labs(x="p-values")+
+  theme_bw()
+
+pdf(file=snakemake@output[[1]], width=12, height=10)
+  
+p + ggforce::facet_row(vars(p_one), scales="free", space="free")
+
+dev.off()

### 2022-11-23 expected_and_actual_counts_ratio/notebooks/expected_and_actual_counts_ratio.Rmd
Update text of notebook to reflect code changes

@@
 Create nucleotide mutation and amino acid mutation data sets.
-Actual and expected counts for rows with identical amino acid mutations are added together.
+Actual and expected counts for rows with identical amino acid mutations and genes are added together.
 Both data sets only include columns relevant for that mutation to make understanding restructuring the data and analyses easier.
@@
 ```{r group_mutation}
-# Group by amino acid mutation and add counts for identical amino acid mutations.
-# NOTE: Grouping with and without 'gene' creates a different number of rows because some equivalent aa mutations belong to different genes. There are 2,160 aa mutations that have duplicates.
+# Group by amino acid mutation and gene, and add counts for identical mutations.
 aa_group <- as.data.frame(filtered_SARS_mutation %>% group_by(clade, aa_mutation, clade_founder_aa, codon_site, mutant_aa, synonymous, gene) %>% summarise(actual_count=sum(actual_count), expected_count=sum(expected_count)))
@@
 nt_group <- filtered_SARS_mutation[,c("clade","nt_mutation","clade_founder_nt","nt_site","synonymous","four_fold_degenerate","gene","actual_count","expected_count")]
-
-# Should I compare mutations with different wild types? (A200T vs C200T) Plot?
 ```
@@
 
-After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visual the proportion of each mutation for each clade.
+After numerically describing the nucleotide and amino acid mutation data, use stacked bar plots to visualize the proportion of each mutation for each clade.
 The number of mutations is normalized by the expected count.
@@
 
-# Merge the two data frames by mutation so that each row represents one comparison.
+# Merge the two data frames by mutation and gene so that each row represents one comparison.
 comparisons <- merge(clade1, clade2, by=c(merge_column, "gene"))
@@
 Fold-change is calculated as log2 of the clade 1 ratio divided by the clade 2 ratio.
-Therefore, if fold-change is positive then the clade 1 ratio is greater than the clade 2 ratio, and vise versa if fold-change is negative.
+Therefore, if fold-change is positive then mutations are enriched in clade 1, and vise versa if fold-change is negative.
 
@@
 The second volcano plot drops P = 1 and uses the remaining corrected p-values.
-Points are colored by the starting nucleotide or amino acid mutation (i.e., clade founder) and labels indicate the mutation.
+Points are colored by gene and labels indicate the mutation.
 The red line indicates a significance value of P = 0.05.
@@
 ```{r volcano_corrected, echo=FALSE}
-#fig.height=10?
 # After dropping P = 1, plot the corrected p-values.
+all_colors <- data.frame(color=colorRampPalette(brewer.pal(8,"Set1"))(length(unique(comparison_p_values$gene))), gene=levels(as.factor(comparison_p_values$gene)))
 volcano <- comparison_p_values[which(comparison_p_values$p_value_corrected!=1),]
-color <- colorRampPalette(brewer.pal(8,"Set1"))(length(unique(volcano$gene)))
+color <- all_colors[all_colors$gene %in% unique(volcano$gene),"color"]
 
@@
   geom_point()+
-  geom_text(color="black", size=2, nudge_y=0.2)+
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
@@
 } else {
-  p <- na + coord_cartesian(xlim=c(-max(abs(x)), max(abs(x))))
+  p <- na + coord_cartesian(xlim=c(-max(abs(x)), max(abs(x)))) + geom_text(color="black", size=3, nudge_y=diff(y)/20)
   x_mod <- p$coordinates$limits[[1]]
@@
 
-To verify the analyses are generating expected results, compare the corrected p-values for synonymous mutations to non-synonymous mutations.
-Unlike non-synonymous mutations, synonymous mutations should not be experiencing selective pressures.
-Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-values close to 0), there should be insignificant differences between clades for synonymous mutations (i.e., p-values close to 1).
+To verify the analyses are generating expected results, compare the uncorrected p-values for synonymous mutations to non-synonymous mutations.
+Synonymous mutations are less likely to be experiencing selective pressures compared to non-synonymous mutations.
+Therefore, while there might be differences in the direction and degree of selection on non-synonymous (i.e., p-values close to 0), we are less likely to observe significant differences between clades for synonymous mutations (i.e., p-values close to 1).
 The black line indicates a significance value of P = 0.05.
@@
 ```{r verify_analyses, echo=FALSE}
-# Plot the corrected p-values for synonymous and non-synonymous mutations.
-# PLOT UNCORRECTED P-VALUES
+# Plot the uncorrected p-values for synonymous and non-synonymous mutations.
 verify_analyses <- comparison_p_values
@@
 }
-verify_analyses$p_one <- ifelse(verify_analyses$p_value_corrected==1, "P=1", "P<1")
+verify_analyses$p_one <- ifelse(verify_analyses$p_value==1, "P=1", "P<1")
 
-p <- ggplot(verify_analyses, aes(x=p_value_corrected, fill=synonymous, label=after_stat(count)))+
+p <- ggplot(verify_analyses, aes(x=p_value, fill=synonymous, label=after_stat(count)))+
   geom_histogram()+
@@
   scale_x_continuous(breaks=c(0,0.25,0.5,0.75,1))+
-  labs(x="corrected p-values")+
+  labs(x="p-values", y="count")+
   theme_bw()

### 2022-11-23 expected_and_actual_counts_ratio/scripts/volcano_plot.R
Minor plotting changes

@@ -31,7 +31,7 @@ na <- ggplot(volcano, aes(x=fold_change, y=-log10(p_value_corrected), label=get(
   geom_hline(yintercept=-log10(0.05), col="red", linetype=2)+
   scale_color_manual(values=color, name="Mutation")+
   theme_bw()+
-  labs(x="Log2 Fold Change", y="-log10 FDR Corrected P-Value")
+  labs(x="Log2 Fold-Change", y="-log10 FDR Corrected P-Value")
 
 x <- layer_scales(na)$x$range$range
 y <- layer_scales(na)$y$range$range

### 2022-11-23 expected_and_actual_counts_ratio/scripts/enriched_per_clade.R
Create script to plot mutations that are enriched in each clade

@@ -1,0 +1,49 @@
+library(ggplot2)
+library(RColorBrewer)
+
+# Create the column to label points with.
+mutation <- paste(snakemake@wildcards[["mutation_type"]], "mutation", sep="_")
+
+# Create filename for reading in all mutation data.
+input_name <- paste("results/pvalues", "_", snakemake@wildcards[["mutation_type"]], ".csv", sep="")
+
+# Read in the data for plotting.
+combined_mutations <- read.csv(input_name, sep=",", header=TRUE)
+
+# Remove corrected p-values equal to 1.
+filtered_mutations <- combined_mutations[which(combined_mutations$p_value_corrected != 1),]
+
+# Filter for mutations enriched in each clade (positive fold-change is enriched in clade 1 and negative fold-change is enriched in clade 2).
+clade1 <- filtered_mutations[which(filtered_mutations$fold_change > 0),c("clade_1","p_value_corrected","fold_change","gene", mutation)]
+clade2 <- filtered_mutations[which(filtered_mutations$fold_change < 0),c("clade_2","p_value_corrected","fold_change","gene", mutation)]
+
+# Change the clade column name so the two data frames have the same column names.
+colnames(clade1)[1] <- "clade"
+colnames(clade2)[1] <- "clade"
+
+# Combine the clade 1 and clade 2 data frames for plotting.
+combined_data <- rbind(clade1, clade2)
+
+# Assign colors to each possible gene in the data set.
+all_colors <- data.frame(color=colorRampPalette(brewer.pal(8,"Set1"))(length(unique(filtered_mutations$gene))), gene=levels(as.factor(filtered_mutations$gene)))
+
+# Only plot data for one clade.
+plot_data <- combined_data[which(combined_data$clade == snakemake@wildcards[["clades"]]),]
+
+# Filter the assigned colors for the genes in the plotting data set.
+color <- all_colors[all_colors$gene %in% unique(plot_data$gene), "color"]
+
+# Generate plot showing the enriched mutations for each clade using all relevant comparisons.
+p <- ggplot(plot_data, aes(x=abs(fold_change), y=-log10(p_value_corrected), color=gene, label=get(mutation)))+
+  geom_point()+
+  scale_color_manual(values=color, name="Mutation")+
+  labs(x="Log2 Fold-Change", y="-log10 FDR Corrected P-Value")+
+  theme_bw()
+
+y <- layer_scales(p)$y$range$range
[…]

### 2022-12-09 expected_and_actual_counts_ratio/scripts/calculate_p_values.R
Finalize at end of rotation

@@ -45,26 +45,32 @@ colnames(clade2)[colnames(clade2) %notin% c(merge_column, "gene")] <- paste(coln
 # Merge the two data frames by mutation so that each row represents one comparison.
 comparisons <- merge(clade1, clade2, by=c(merge_column, "gene"))
 
-# Determine if any of the columns from 'clade1' and 'clade2' are the same.
-col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% c(merge_column, "gene")]
-is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
-col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]
+if(nrow(comparisons)!=0){
+	# Determine if any of the columns from 'clade1' and 'clade2' are the same.
+	col_names <- colnames(filtered_SARS_mutation)[colnames(filtered_SARS_mutation) %notin% c(merge_column, "gene")]
+	is_equal <- data.frame(sapply(col_names, FUN=function(c){identical(comparisons[,paste(c[1], 1, sep="_")], comparisons[,paste(c[1], 2, sep="_")])}))
+	col_equal <- row.names(is_equal)[which(is_equal[1]=="TRUE")]
 
-# Rename one of the duplicate columns and remove the other.
-colnames(comparisons)[colnames(comparisons) %in% paste(col_equal, "1", sep="_")] <- col_equal
-comparisons[,paste(col_equal, "2", sep="_")] <- NULL
+	# Rename one of the duplicate columns and remove the other.
+	colnames(comparisons)[colnames(comparisons) %in% paste(col_equal, "1", sep="_")] <- col_equal
+	comparisons[,paste(col_equal, "2", sep="_")] <- NULL
 
-# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
-p_values <- data.frame(p_value=apply(comparisons, 1, FUN=function(c) {fisher.test(rbind(c(as.numeric(c["actual_count_1"]),as.numeric(c["rounded_expected_1"])),c(as.numeric(c["actual_count_2"]),as.numeric(c["rounded_expected_2"]))))$p.value}))
+	# Iterate through rows and compare clade 1 observed and expected values to clade 2 observed and expected values.
[…]

### 2022-12-09 expected_and_actual_counts_ratio/scripts/enriched_per_clade.R

@@ -1,49 +1,0 @@
-library(ggplot2)
-library(RColorBrewer)
-
-# Create the column to label points with.
-mutation <- paste(snakemake@wildcards[["mutation_type"]], "mutation", sep="_")
-
-# Create filename for reading in all mutation data.
-input_name <- paste("results/pvalues", "_", snakemake@wildcards[["mutation_type"]], ".csv", sep="")
-
-# Read in the data for plotting.
-combined_mutations <- read.csv(input_name, sep=",", header=TRUE)
-
-# Remove corrected p-values equal to 1.
-filtered_mutations <- combined_mutations[which(combined_mutations$p_value_corrected != 1),]
-
-# Filter for mutations enriched in each clade (positive fold-change is enriched in clade 1 and negative fold-change is enriched in clade 2).
-clade1 <- filtered_mutations[which(filtered_mutations$fold_change > 0),c("clade_1","p_value_corrected","fold_change","gene", mutation)]
-clade2 <- filtered_mutations[which(filtered_mutations$fold_change < 0),c("clade_2","p_value_corrected","fold_change","gene", mutation)]
-
-# Change the clade column name so the two data frames have the same column names.
-colnames(clade1)[1] <- "clade"
-colnames(clade2)[1] <- "clade"
-
-# Combine the clade 1 and clade 2 data frames for plotting.
-combined_data <- rbind(clade1, clade2)
-
-# Assign colors to each possible gene in the data set.
-all_colors <- data.frame(color=colorRampPalette(brewer.pal(8,"Set1"))(length(unique(filtered_mutations$gene))), gene=levels(as.factor(filtered_mutations$gene)))
-
-# Only plot data for one clade.
-plot_data <- combined_data[which(combined_data$clade == snakemake@wildcards[["clades"]]),]
-
-# Filter the assigned colors for the genes in the plotting data set.
-color <- all_colors[all_colors$gene %in% unique(plot_data$gene), "color"]
-
-# Generate plot showing the enriched mutations for each clade using all relevant comparisons.
-p <- ggplot(plot_data, aes(x=abs(fold_change), y=-log10(p_value_corrected), color=gene, label=get(mutation)))+
-  geom_point()+
-  scale_color_manual(values=color, name="Mutation")+
-  labs(x="Log2 Fold-Change", y="-log10 FDR Corrected P-Value")+
-  theme_bw()
-
-y <- layer_scales(p)$y$range$range
[…]

### 2022-12-09 expected_and_actual_counts_ratio/notebooks/combined_data_plotting.R
Create script for generating presentation plots

@@ -1,0 +1,194 @@
+library(ggplot2)
+library(dplyr)
+library(RColorBrewer)
+library(stringr)
+library(reshape)
+
+data <- read.csv("/home/kfeldman/kfeldmann/2022/SARS2-mut-rates/expected_and_actual_counts_ratio/results/pvalues_aa.csv")
+
+# Proportion of significantly enriched mutations for all clade comparisons
+data$significant <- ifelse(data$p_value_corrected<=0.05, 1, 0)
+temp <- data %>% group_by(clade_1, clade_2) %>% summarise(Proportion=sum(significant)/length(significant))
+ggplot(temp, aes(x=clade_1, y=clade_2, fill=Proportion))+
+  geom_tile()+
+  theme_bw()+
+  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5)) +
+  labs(x="Clade 1", y="Clade 2")
+
+# Proportion of significantly enriched mutations for ORF8 and S across clades
+data$significant <- ifelse(data$p_value_corrected<=0.05, 1, 0)
+data$enriched_clade <- ifelse(data$fold_change>0, data$clade_1, data$clade_2)
+temp <- as.data.frame(data %>% group_by(enriched_clade, gene) %>% summarise(Proportion=sum(significant)/length(significant)))
+
+all_colors <- data.frame(color=colorRampPalette(brewer.pal(8,"Set1"))(length(unique(temp$gene))), gene=levels(as.factor(temp$gene)))
+all_colors[,"color"] <- "black"
+all_colors[which(all_colors$gene=="S"),"color"] <- "orange"
+all_colors[which(all_colors$gene=="ORF8"),"color"] <- "skyblue"
+ggplot(temp, aes(x=enriched_clade, y=Proportion, color=gene))+
+  geom_point(size=4)+
+  scale_color_manual(values=all_colors$color, name="Gene")+
+  theme_bw()+
+  labs(x="Clade")
+
+# Proportion of significantly enriched mutations for each gene (clades combined; highlighting spike)
+data$significant <- ifelse(data$p_value_corrected<=0.05, 1, 0)
+temp <- as.data.frame(data %>% group_by(gene) %>% summarise(Proportion=sum(significant)/length(significant)))
+temp <- temp[which(temp$Proportion>0),]
+temp$gene <- factor(temp$gene, levels = temp[rev(order(temp$Proportion)),"gene"])
+
[…]

### 2023-01-03 scripts/format_altair_html.py
format legends for interactive GitHub pages plots in `./docs/` (#9)
* structure `Snakefile` to allow plot formatting

* build plots for docs from `plot_annotations.yaml`

* code to format plots for docs

* add legends for heat maps

* annotations for mut rates plot

* add clade synonyms for mut rates plot

* add legend for `effects_histogram`

* add legend for `effects_dist`

* add legend for subset and clade correlations

* change location of legend templates in `docs_plot_annotations.yaml`

* add spike DMS plot lgends

* add plot legends for Mpro DMS

* legend for terminal vs non-terminal

*

@@ -1,0 +1,163 @@
+"""Annotate ``altair`` chart with Twitter card and markdown summary.
+
+You can either use function in module or run as command-line tool.
+
+"""
+
+
+import argparse
+
+from bs4 import BeautifulSoup as bs
+
+import markdown
+
+
+def annotate_altair_chart(chart_html, annotation_md, twitter_card):
+    """
+    This function annotates an altair chart with a twitter card and markdown
+    description.
+
+
+    Parameters
+    ----------
+    chart_html: str
+        Path to an HTML file with embeded Vega spec.
+    annotation_md: str
+        Path to a text file with markdown description to appended to body.
+    twitter_card: dict
+        Site name, title, description and optionally image for Twitter card.
+
+    Returns
+    -------
+    str:
+        A string of the HTML page formatted to be human readable.
+    """
+
+    # Get the main page content
+    with open(chart_html, "r") as chart_file:
+        page = bs(chart_file, "html.parser")
+
+    # Get the annotation and convert it from markdown to HTML
+    with open(annotation_md, "r") as markdow_file:
+        annotation = bs(markdown.markdown(markdow_file.read()), "html.parser")
+
+    # Add the annotation to the bottom of the page
+    markdown_container = page.new_tag("div", attrs={"id": "markdown"})
+    page.body.append(markdown_container)
+    separator = page.new_tag("hr")
+    markdown_container.append(separator)
+    markdown_container.append(annotation)
+
+    # Make and add the twitter card
+    if not all(key in twitter_card.keys() for key in ["site", "title", "description"]):
+        raise ValueError(
+            "Missing required fields for twitter card: site, title, or description"
+        )
+    summary = page.new_tag("meta", attrs={"name": "twitter:card", "content": "summary"})
+    page.head.append(summary)
+    for name, content in twitter_card.items():
+        card_tag = page.new_tag(
+            "meta", attrs={"name": f"twitter:{name}", "content": content}
+        )
+        page.head.append(card_tag)
+
+    # Add some default styling with bootstrap
+    stylesheet = page.new_tag(
+        "link",
+        attrs={
+            "rel": "stylesheet",
+            "href": "https://cdn.jsdelivr.net/npm/bootstrap@4.3.1/dist/css/bootstrap.min.css",  # noqa: E501
+            "integrity": "sha384-ggOyR0iXCbMQv3Xipma34MD+dH/1fQ784/j6cY/iJTQUOhcWr7x9JvoRxT2MZw1T",  # noqa: E501
[…]

### 2023-01-04 notes/fitness_inference.tex
add notes

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}
\begin{document}
\title{Estimating fitness costs from mutation counts}
\author{Richard Neher}
\maketitle

With millions of SARS-CoV-2 sequences shared publicly, almost all mutations that are tolerated by the virus are observed dozens to hundreds of times.
Where on the tree and how often on the tree we observe specific mutations has information about the effect of these mutations on viral spread.
The mutation rate depends on the nucleotides involved and possibly on the sequence context and other viral determinants, but for the purpose of this note, we will assume the neutral rate $\mu$ is known.
If the mutation is neutral, the total number of times the mutation is observed on the tree is $\mu T$, where $T$ is the total length of the tree (assuming that the mutation never reached high frequency which is true for almost all mutations.).

If a mutation reduces fitness, the lineages descending from branches on which this mutation happened will spread more slowly than the those without this mutation.
As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
The number of times a mutation is observed in the tree can be readily assessed using UShER and compared to the number of times we'd expect to see the mutation if it was neutral (e.g.~same change at 4-fold synonymous sites).
To infer its effect on fitness, we need to calculate how the number of observations depends on the this fitness effect.

For a mutation to be represented in the tree, one of its descendents has to be sampled and sequenced.
If the total number of descendents is $w$ and the sampling fraction is $\epsilon$, the probability that the mutation is present in the tree is
\begin{equation}
    p = 1 - e^{-w\epsilon}
\end{equation}
$W$ is a random number that depends on the realization of the transmission process, which is commonly modeled by a branching process with birth rate $b$ and death rate $d$.
The death rate here corresponds to clearing an infection, the birth rate to onward transmission. The latter is affected by the fitness cost of the mutation.

To obtain insight how the probability of observing a lineage depends on parameters, we calculate the probability $p(w, T|t)$ that a lineage had in integrated size $w = \int_t^T k(t') \; dt'$, where $t$ is the birth time of the lineage, $T$ is the current time, and $k(t')$ is the size of the lineage at time $t'$.
To calculate $p(w,T|k)$, we generalize is slightly to $p(W, T|k,t)$, where $k$ is the number of individuals at the start time $t$. This quantity obeys the following ``first-step'' equation:
\begin{equation}
    -(\partial_t - k \partial_w) p(w,T|k,t) = -k(b+d) p(w,T|k,t) + k b p(w,T|k+1,t) + k d p(w,T| k-1, t)
\end{equation}
We will solve for the Laplace transform $\hat{p}(z,T|k,t) = \int_0^\infty dw\; e^{-wz} p(w,T|k,t) = \hat{p}^k(z,T|1,t)$.
Using the identity for the derivative of the Laplace transform
\begin{equation}
    \begin{split}
        \int_0^\infty e^{-wz} \partial_w p  \; dw& = [e^{-wz} p]_0^\infty - \int_0^\infty  p \partial_w e^{-wz} \; dw = 0 + z \int_0^\infty  p e^{-wz} = z \hat{p}
    \end{split}
\end{equation}
and using $k=1$, we have
\begin{equation}
    -\partial_t \hat{p}(z,T|t) = -(b+d+z) \hat{p}(z,T|t) + b \hat{p}^2(z,T|t) + d
\end{equation}
This simplifies further to if we substitute $\phi(z,T|t) = 1 - \hat{p}(z,T|t)$.
\begin{equation}
    \begin{split}
        \partial_t \phi(z,T|t) & = -(b+d+z) (1-\phi(z,T|t)) + b (1-\phi(z,T|t))^2 + d \\
         & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
    \end{split}
\end{equation}
where it is important to note that the derivative is with respect to the first time point and the interval $T-t$ is shrinking with increasing $t$.

\subsection*{Constant birth and death rate}
If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and $\partial_t \phi(z,T|t)$.
Further define $b=b_0 - s$ and $d=d_0$.
The steady state generating function is then
\begin{equation}
    0 = -z - (b-d-z) \phi(z) + b \phi(z)^2
\end{equation}
with solution
\begin{equation}
    \begin{split}
        \phi(z) & = -\frac{s + z}{2(b_0 - s)} \pm \frac{\sqrt{(s+z)^2 + 4z(b_0-s)}}{2(b_0-s)} \\
        &\approx -\frac{s + z}{2b_0} \pm \frac{\sqrt{(s+z)^2 + 4z b_0}}{2b_0} \\
        & \approx \begin{cases}
            \frac{z}{s+z} & (s+z)^2 \gg 4zb_0 \\
            \sqrt{\frac{z}{b_0}}\left(1+\frac{(s+z)^2}{8zb_0}\right)-\frac{s + z}{2b_0}  & (s+z)^2 \ll 4zb_0 \\
        \end{cases}
    \end{split}
\end{equation}
Since $\phi(z) = 1 - \int e^{-wz}p(w) \; dw$, $\phi(\epsilon)$ is exactly the probability that a lineage is sampled when the entire population is sampled at rate $\epsilon$.
We thus expect two regimes: if the square of the fitness effect exceeds the sampling density (typically at 1\% or less), the probability of sampling a lineage is essentially inversely proportional to the fitness effect.
The sampling probability of lineages with smaller fitness effects depends less strongly on $s$.
Their sampling mostly comes down to stochasticity independent of the fitness effect.

\subsection*{Growing populations}
In many scenarios relevant for lineages that arise during a viral outbreak, the background population isn't constant but is undergoing a rapid exponential expansion.
The background birth rate $b_0$ is bigger than $d_0$ in this case.
Since the population is growing, deleterious mutations can increase in frequency deterministically and we can not send the $t$ to infinity as before.
Instead, we need to integrate
\begin{equation}
    \begin{split}
        \partial_t \phi(z,T|t) & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
    \end{split}
\end{equation}
backwards in time starting from $\phi(z,T|T)=0$ at $t=T$.
While $\phi(z,T|t)$ is small, this is approximately solved by
\begin{equation}
\begin{split}
    \phi(z,T|t) & = z e^{\int_t^T (b-d-z)dt'} \int_t^T e^{-\int_\tau^T (b-d-z)dt'} d\tau \\
    & = z e^{(b-d-z)(T-t)} \int_t^T e^{-(b-d-z)(T-\tau)} d\tau \\
    & = z e^{(b-d-z)(T-t)}\left[1   - e^{-(b-d-z)(T-t)}\right]/(b-d-z) \\
    & = \frac{z}{b-d-z}\left[e^{(b-d-z)(T-t)} - 1\right] = \frac{z}{\gamma_0 - s - z}\left[e^{\gamma_0 (T-t) - (s+z) (T-t)} - 1\right]
\end{split}
\end{equation}
where $\gamma_0$ is the growth rate of the background population.

At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
\begin{equation}
    \begin{split}
        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
        &\approx \begin{cases}
            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
        \end{cases}
    \end{split}
\end{equation}
This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
In this case, the above simplifies to
\begin{equation}
    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
\end{equation}
This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
In growing population, the approximation fails if the product of fitness effect and the time over which the variant is grown is large, i.e., if the fitness effect is substantial on this time scale.
Outside this parameter range, there is still a dependence on $s$, but it is weaker.

\end{document}

### 2023-01-05 notes/fitness_inference.tex
add notes
Richard's notes on calculations (#10)
* add notes

* add notes

* add notes

Co-authored-by: Richard Neher <richard.neher@unibas.ch>

@@
 \title{Estimating fitness costs from mutation counts}
-\author{Richard Neher}
 \maketitle
@@
 \end{equation}
+
+In a variant that has been growing with rate $\gamma_0$ for a time $\tau=T-t$ and sampled with $z=\epsilon$, we thus expect that the number of times we observe separate mutant lineages depends on $s$ as
+\begin{equation}
+    \langle \phi \rangle \approx \frac{\epsilon}{\gamma_0(\epsilon+s)}
+\end{equation}
 This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
 In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
-In growing population, the approximation fails if the product of fitness effect and the time over which the variant is grown is large, i.e., if the fitness effect is substantial on this time scale.
+In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is large, i.e., if the fitness effect is substantial on this time scale.
 Outside this parameter range, there is still a dependence on $s$, but it is weaker.

### 2023-01-12 paper/paper.tex
outline paper

\documentclass[9pt,twocolumn,twoside]{gsajnl_modified}
% Use the documentclass option 'lineno' to view line numbers

\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
\usepackage[export]{adjustbox}
\usepackage{xurl}
\usepackage{stfloats}
\usepackage[leftcaption]{sidecap}
\sidecaptionvpos{figure}{t}

\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
    %   Parameters for TEXT pages (not float pages):
    \setcounter{topnumber}{2}
    \setcounter{bottomnumber}{2}
    \setcounter{totalnumber}{4}     % 2 may work better
    \setcounter{dbltopnumber}{2}    % for 2-column pages
    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
    %   Parameters for FLOAT pages (not text pages):
    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
	% N.B.: floatpagefraction MUST be less than topfraction !!
    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages

\newcommand\jdbcomment[1]{\textcolor{red}{[#1]}}

\title{Estimated fitness effects of amino-acid mutations to SARS-CoV-2 proteins}

\author[*]{\Large Jesse D. Bloom$^{1,2,3^*}$ and Richard A. Neher$^{4,5}$}

\affil[1]{Basic Sciences and Computational Biology, Fred Hutchinson Cancer Center

} 
\affil[2]{Department of Genome Sciences, University of Washington

}
\affil[3]{Howard Hughes Medical Institute

}
\affil[4]{Biozentrum, University of Basel

}
\affil[5]{Swiss Institute of Bioinformatics

\jdbcomment{Should we add Angie Hinrichs as co-author? She helped answer some UShER questions on GitHub. Probably just acknowledgments is fine, but what do you think?}

}

\keywords{}

\runningtitle{} % For use in the footer 
\runningauthor{}

\begin{abstract}
Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new viral variants and targeting of therapeutics to constrained sites where resistance is unlikely to emerge.
However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for the functions of many SARS-CoV-2 proteins, and for this reason systematic deep mutational scanning has been applied to only two of the virus's proteins.
Here we develop a new approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate the effects of most amino-acid mutations accessible by single-nucleotide changes.
To do this, we first calculate how many independent times each mutation is expected to occur along the SARS-CoV-2 phylogeny in the absence of selection. 
We then compare these expected counts to the actual observed counts of each mutation, and use the ratio of counts to estimate the effect of each mutation.
For the two proteins with deep mutational scanning data, our sequence-based estimates correlate with experimental measurements nearly as well as independent experiments correlate with each other.
For most genes, synonymous mutations are nearly neutral, stop-codon mutations are deleterious, and amino-acid mutations have a range of effects.
However, some viral accessory proteins are under little to no selection during SARS-CoV-2 evolution in humans.
We provide interactive visualizations of the estimated effects of individual mutations to each viral protein, enabling these data to be easily used in viral surveillance and drug development.
Overall, our work provides a framework to estimate the effects of mutations to any virus (or organism) for which the number of available sequences is sufficiently large that that each neutral mutation independently occurs many times.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
%\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{}{*\href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org} or \href{mailto:richard.neher@unibas.ch}{richard.neher@unibas.ch}}
\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page

\lettrine[lines=2]{\color{color2}B}{}ackground. It's important to understand effects of individual mutations.

Experiments have filled this role for some viral proteins (spike, Mpro), but not for majority of SARS-CoV-2 viral protein.

An alternative is sequence-based estimates, which so far have focused largely on clade growth. \jdbcomment{Richard, can you note any key papers you think should be cited as background.}

Here we describe new approach based on comparing expected versus observed counts.

\section{Results}

\section{Discussion}

{\small

\section{Methods}
\subsection{Code and data availability}
Code and data are at \url{https://github.com/jbloomlab/SARS2-mut-fitness}.

\section{Acknowledgments}
We thank Angie Hinrichs for assistance with resolving several questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{if we don't add as co-author}.
\jdbcomment{Jesse to acknowledge AVIDD, CEIRR, and R01 grants.}
JDB is an Investigator of the Howard Hughes Medical Institute.

\section{Competing interests}
JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus.
JDB receives payments as an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.

\bibliography{references}
}

\end{document}

### 2023-01-12 paper/paper.tex
draft of paper figures

@@
 
+\subsection{Theory}
+\jdbcomment{Include Richard's calculations, either in entirety or summarized with details in Methods / Appendix}
+
+\begin{figure*}
+{\bf \Large A \hspace{0.47\linewidth} B}
+
+\includegraphics[width=0.48\linewidth,valign=t]{figs/schematic/schematic.pdf}
+\hspace{0.02\linewidth}
+\includegraphics[width=0.5\linewidth,valign=t]{figs/avg_counts.pdf}
+\caption{
+Calculation of expected versus actual counts for each mutation.
+{\bf (A)}
+The number of expected counts of each type of nucleotide mutation is computed from the observed counts at four-fold degenerate sites,
+and then compared the actual observed counts of each mutation.
+Note that the counts represent independent occurrences of each mutation on the tree, not counts of mutations in the alignment of observed sequences (a single occurrence of a mutation can be present in many sequences in an alignment).
+{\bf (B)}
+Expected versus actual counts for each nucleotide mutation type averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
+Synonymous mutations are often roughly neutral, so the number of expected and actual counts are similar, but nonsynonymous and especially stop-codon mutations are often deleterious, so the number of actual counts is generally less than the number of expected counts.
+Note that due to the highly uneven mutation spectrum of SARS-CoV-2, some nucleotide mutation types are expected to occur far more often than others.
+An interactive version of panel B is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html}.
+\label{fig:expected_vs_actual}
+}
+\end{figure*}
+
+\begin{figure*}
+{\bf \Large A}
+
+\includegraphics[width=\linewidth, trim={0in 0.7in 0 0.05in}, clip=true]{figs/effects_histogram.pdf}
+
+{\bf \Large B}
+
+\includegraphics[width=\linewidth, trim={0in 0.36in 0 0in}, clip=true]{figs/effects_dist.pdf}
+\caption{
+Distribution of effects of different types of mutations.
+{\bf (A)}
+Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
+Neutral mutations have effects of zero, and deleterious mutations have negative effects.
+\jdbcomment{Maybe add dotted vertical line at zero to denote neutrality?}
+{\bf (B)}
+Distribution of effects of each type of mutation for each viral gene.
+The dark squares indicate the median effect of mutations of that type for that gene, and the lighter rectangles span the interquartile range.
+Mutation types are color-coded as in panel A.
+Both panels show only mutations with expected counts of at least 20.
+For interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene), see \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html}.
+\label{fig:effects_dist}
+}
+\end{figure*}
+
+\begin{figure*}
+\centering
+{\bf \Large A \hspace{0.32\linewidth} B \hspace{0.32\linewidth}}
+
+\includegraphics[width=0.32\linewidth]{figs/clade_corr_chart.pdf}
+\hspace{0.05\linewidth}
+\includegraphics[width=0.32\linewidth]{figs/subset_corr_chart.pdf}
+\caption{
+Correlations between amino-acid fitness estimates made using different sequence subsets.
+{\bf (A)} Estimates made using only sequences from the Omicron BA.2 clade or only the Alpha clade, or {\bf (B)} estimates made using only sequences from the USA or England.
+In the scatter plots, each point is a different amino-acid mutation, and the orange text at upper left shows the number of mutations and Pearson correlation coefficient.
+Both panels show only mutations with expected counts of at least 20.
+For interactive versions of these plots that show correlations among estimates from all clade pairs, allow adjustment of the expected-count cutoff, and allow examination of correlations for just specific genes, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html}.
+\label{fig:corr}
+}
+\end{figure*}
+
+\begin{figure*}
+{\bf \Large A}
+
+\includegraphics[width=0.9\linewidth]{figs/dms_S_corr.pdf}
+\vspace{0.05in}
+
+{\bf \Large B}
+
+\includegraphics[width=0.9\linewidth]{figs/dms_nsp5_corr.pdf}
+\caption{
+Correlation of fitness effect estimates with experimental measurements from deep mutational scanning studies for {\bf (A)} spike and {\bf (B)} Mpro (nsp5).
+Each point is an amino-acid mutation, and the oragne text in the upper left shows the number of mutations and Pearson correlation coefficient.
+Data are taken from two independent deep mutational scanning studies for spike~\citep{cite}, and two studies for Mpro~\citep{cite}.
+Each sub-panel shows a different set of mutations (those that are measured in each of the two datasets being correlated), and uses a minimum expected count cutoff of 20.
+Interactive versions of these plots that enable subsetting on just mutations shown across all datasets and adjustment of the minimum expected count cutoff are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html}.
+\label{fig:dms_corr}
+}
+\end{figure*}
+
+\begin{figure*}
+{\bf \Large A}
+
+E protein mutation effects with minimum expected counts of 20 and \textbf{x} marking wildtype identity in Wuhan-Hu-1.
+
+\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view1.pdf}
+
+{\bf \Large B}
+
+E protein mutation effects with minimum expected counts of 5 and \textbf{x} marking wildtype identity in BQ.1.
+
+\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view2.pdf}
+\caption{
+Effects of amino-acid mutations to E protein.
+{\bf (A)} and {\bf (B)} provide two views of the estimated mutation effects with different thresholds for the minimum expected counts and different references used to mark the ``wildtype'' identity, as indicated in the titles above the plots.
+For both panels, the area plot at top shows the average effects of mutations at each site, and the heatmap shows the effects of specific amino acids.
+An interactive version of this plot at \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} enables zooming, mouseovers, and exploration of the data.
+Comparable interactive plots for all other viral proteins are linked at \jdbcomment{create landing pad page with these links}.
+\label{fig:E}
+}
+\end{figure*}
+
+\begin{figure*}
+{\bf \Large A}
+
+\includegraphics[width=0.8\linewidth]{figs/clade_fixed_muts_hist.pdf}
+\vspace{0.05in}
+
+{\bf \Large B}
+
+\includegraphics[width=\linewidth]{figs/clade_fixed_muts.pdf}
+
+\caption{
+Effects of amino-acid mutations that have fixed in at least one SARS-CoV-2 clade.
+{\bf (A)}
+Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1 and all mutations that are fixed in at least one of the clades (as defined by Nextclade\jdbcomment{cite}).
+\jdbcomment{add vertical line at zero to these plots.}
+{\bf (B)}
+Effects of individual fixed mutations faceted by whether they are in spike or another protein.
+Plots show estimates of effects of mutations made across all clades (including those that have fixed the mutation) and just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
+Estimates are only shown for mutations with expected counts of at least 20.
+Interactive versions of this plot that enable easier exploration of the data are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html}.
+\label{fig:fixed}
+}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[width=0.8\linewidth, trim={0 1.1in 0 0in}, clip=true]{figs/fitness_vs_terminal.pdf}
+\caption{
+Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
+Each point is an amino-acid mutation, and the orange text in the upper right give the number of mutations and the Pearson correlation coefficient.
+This plot shows only mutations with at least 20 expected counts and 5 actual counts.
+An interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene, is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html}.
+\jdbcomment{add same color coding as in other panels for synonymous, nonsynonymous, stop types}
+\label{fig:terminal}
+}
+\end{figure*}
+
 \section{Discussion}

### 2023-01-14 scripts/docs_index.py
make a single index page for all plots on GitHub pages

@@ -1,0 +1,31 @@
+"""Build index for GitHub pages docs."""
+
+
+import markdown
+
+html_file = snakemake.output.html
+docs_url = snakemake.params.docs_url
+plot_annotations = snakemake.params.plot_annotations
+
+text = [
+    f"## {plot_annotations['index_title']}",
+    plot_annotations["index_abstract"],
+    "Interactive plots:",
+]
+
+for section, section_title in plot_annotations["sections"].items():
+    text.append(f"\n  - {section_title}")
+    for plot, plot_d in plot_annotations["plots"].items():
+        assert plot_d["section"] in plot_annotations["sections"], plot_d
+        if section == plot_d["section"]:
+            text.append(
+                f"    - [{plot_d['title']}]({docs_url}/{plot}.html)"
+            )
+
+text.append("\n" + plot_annotations["legend_suffix"])
+
+html = markdown.markdown("\n".join(text))
+
+with open(html_file, "w") as f:
+    f.write(html)
+

### 2023-01-14 paper/paper.tex
add references and small edits to paper.tex (#12)
Co-authored-by: Richard Neher <richard.neher@unibas.ch>

@@
 \newcommand\jdbcomment[1]{\textcolor{red}{[#1]}}
+\newcommand\rancomment[1]{\textcolor{blue}{[#1]}}
 
@@
 
-An alternative is sequence-based estimates, which so far have focused largely on clade growth. \jdbcomment{Richard, can you note any key papers you think should be cited as background.}
+An alternative is sequence-based estimates, which so far have focused largely on clade growth and identification of mutations that mediate immune escape or increase transmissibility \citep{obermeyer2022analysis,lee2022inferring}.
+Mutations that are repeatedly observed at the base of clades that grow faster than the background likely contribute to growth.
+Such adaptive mutations are only a small minority of all available mutations and which mutations are adaptive depends on the environment (immunity) and the genetic background.
+The great majority of mutations, however, have either small fitness effects (approximately neutral) or are deleterious to viral replication.
+Here, we describe an approach to quantify the fitness cost of these mutations based on comparing expected versus observed counts.
 
-Here we describe new approach based on comparing expected versus observed counts.
+\rancomment{older paper on flu where not counts but trajectories were used to partition sequences into good and bad \citep{strelkowa2021clonal}}
 

### 2023-01-15 paper/paper.tex
minor tweaks to abstract

--- 2023-01-12 paper/paper.tex
@@
 
-\title{Estimated fitness effects of amino-acid mutations to SARS-CoV-2 proteins}
+\title{Fitness effects of mutations to SARS-CoV-2 proteins}
 
@@
 
-\jdbcomment{Should we add Angie Hinrichs as co-author? She helped answer some UShER questions on GitHub. Probably just acknowledgments is fine, but what do you think?}
-
 }
@@
 \begin{abstract}
-Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new viral variants and targeting of therapeutics to constrained sites where resistance is unlikely to emerge.
-However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for the functions of many SARS-CoV-2 proteins, and for this reason systematic deep mutational scanning has been applied to only two of the virus's proteins.
-Here we develop a new approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate the effects of most amino-acid mutations accessible by single-nucleotide changes.
-To do this, we first calculate how many independent times each mutation is expected to occur along the SARS-CoV-2 phylogeny in the absence of selection. 
-We then compare these expected counts to the actual observed counts of each mutation, and use the ratio of counts to estimate the effect of each mutation.
-For the two proteins with deep mutational scanning data, our sequence-based estimates correlate with experimental measurements nearly as well as independent experiments correlate with each other.
+Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new viral variants and targeting of therapeutics to sites where resistance is unlikely to emerge.
+However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and systematic deep mutational scanning has been applied to only two of the virus's proteins.
+Here we develop a new approach that leverage millions of publicly available SARS-CoV-2 sequences to estimate the effects of most amino-acid mutations accessible by single-nucleotide changes.
+We first calculate how many times each mutation is expected to occur along the SARS-CoV-2 phylogeny in the absence of selection. 
+We then compare these expected counts to the actual observed counts to estimate the effect of each mutation.
+Our sequence-based estimates correlate with experimental deep mutational scanning measurements nearly as well as independent experiments correlate with each other.
 For most genes, synonymous mutations are nearly neutral, stop-codon mutations are deleterious, and amino-acid mutations have a range of effects.
-However, some viral accessory proteins are under little to no selection during SARS-CoV-2 evolution in humans.
-We provide interactive visualizations of the estimated effects of individual mutations to each viral protein, enabling these data to be easily used in viral surveillance and drug development.
-Overall, our work provides a framework to estimate the effects of mutations to any virus (or organism) for which the number of available sequences is sufficiently large that that each neutral mutation independently occurs many times.
+However, some viral accessory proteins are under little to no selection.
+To enable our results to be easily used in viral surveillance and drug development, we provide interactive visualizations of the effects of mutations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
+The framework we describe is applicable to any virus (or organism) for which the number of available sequences is sufficiently large that that each neutral mutation independently occurs many times.
 \end{abstract}
@@
 \caption{
-Calculation of expected versus actual counts for each mutation.
+Expected versus actual counts of mutations.
 {\bf (A)}
-The number of expected counts of each type of nucleotide mutation is computed from the observed counts at four-fold degenerate sites,
-and then compared the actual observed counts of each mutation.
-Note that the counts represent independent occurrences of each mutation on the tree, not counts of mutations in the alignment of observed sequences (a single occurrence of a mutation can be present in many sequences in an alignment).
+The number of expected counts of each type of nucleotide mutation is computed from four-fold degenerate sites, and then compared the actual counts of each mutation.
+The counts are of independent occurrences of each mutation on the tree, not mutations in the alignment of observed sequences (a single occurrence of a mutation can be present in many sequences in an alignment).
 {\bf (B)}
 Expected versus actual counts for each nucleotide mutation type averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
-Synonymous mutations are often roughly neutral, so the number of expected and actual counts are similar, but nonsynonymous and especially stop-codon mutations are often deleterious, so the number of actual counts is generally less than the number of expected counts.
-Note that due to the highly uneven mutation spectrum of SARS-CoV-2, some nucleotide mutation types are expected to occur far more often than others.
-An interactive version of panel B is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html}.
+Synonymous mutations are often roughly neutral, so the expected and actual counts are similar.
+Nonsynonymous and especially stop-codon mutations are often deleterious, so the actual counts are generally less than the expected counts.
+Due to the uneven mutation spectrum of SARS-CoV-2, some nucleotide mutation types are expected to occur more often than others.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html} for an interactive version of panel B.
 \label{fig:expected_vs_actual}
@@
 {\bf (B)}
-Distribution of effects of each type of mutation for each viral gene.
-The dark squares indicate the median effect of mutations of that type for that gene, and the lighter rectangles span the interquartile range.
+Effects of each type of mutation for each viral gene.
+Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
 Mutation types are color-coded as in panel A.
 Both panels show only mutations with expected counts of at least 20.
-For interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene), see \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html}.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene).
 \label{fig:effects_dist}

### 2023-01-15 paper/paper.tex
clean up figure 2

@@
 \begin{figure*}
-{\bf \Large A}
-
-\includegraphics[width=\linewidth, trim={0in 0.7in 0 0.05in}, clip=true]{figs/effects_histogram.pdf}
-
-{\bf \Large B}
-
-\includegraphics[width=\linewidth, trim={0in 0.36in 0 0in}, clip=true]{figs/effects_dist.pdf}
+\includegraphics[width=\linewidth]{figs/fig2.pdf}
 \caption{
@@
 Histograms of effects of synonymous, nonsynonymous, and stop-codon mutations across all viral genes.
-Neutral mutations have effects of zero, and deleterious mutations have negative effects.
-\jdbcomment{Maybe add dotted vertical line at zero to denote neutrality?}
+Neutral mutations have effects of zero (dashed gray vertical lines), and deleterious mutations have negative effects.
 {\bf (B)}
@@
 Mutation types are color-coded as in panel A.
-Both panels show only mutations with expected counts of at least 20.
+Both panels show only mutations with expected counts of at least 10.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene).

### 2023-01-15 paper/paper.tex
update corr figure

@@
 \begin{figure*}
-\includegraphics[width=\linewidth]{figs/fig2.pdf}
+\includegraphics[width=\linewidth]{figs/dist.pdf}
 \caption{
@@
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene).
-\label{fig:effects_dist}
+\label{fig:dist}
 }
@@
 \centering
-{\bf \Large A \hspace{0.32\linewidth} B \hspace{0.32\linewidth}}
-
-\includegraphics[width=0.32\linewidth]{figs/clade_corr_chart.pdf}
-\hspace{0.05\linewidth}
-\includegraphics[width=0.32\linewidth]{figs/subset_corr_chart.pdf}
+\includegraphics[width=\linewidth]{figs/corr.png}
 \caption{
 Correlations between amino-acid fitness estimates made using different sequence subsets.
-{\bf (A)} Estimates made using only sequences from the Omicron BA.2 clade or only the Alpha clade, or {\bf (B)} estimates made using only sequences from the USA or England.
-In the scatter plots, each point is a different amino-acid mutation, and the orange text at upper left shows the number of mutations and Pearson correlation coefficient.
-Both panels show only mutations with expected counts of at least 20.
-For interactive versions of these plots that show correlations among estimates from all clade pairs, allow adjustment of the expected-count cutoff, and allow examination of correlations for just specific genes, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html}.
+{\bf (A)} Estimates made using only sequences from the indicated viral clades, or {\bf (B)} estimates made using only sequences from the USA or England.
+Each point is a different amino-acid mutation, and the orange text at upper left shows the number of mutations and Pearson correlation coefficient.
+Both panels show only mutations with expected counts of at least 10 in each subset of sequences.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html}.For interactive versions of these plots that show correlations among additional clade pairs, allow adjustment of the expected-count cutoff, and allow examination of mutations for just specific genes.
 \label{fig:corr}

### 2023-01-15 paper/paper.tex
add dms figure

@@
 \begin{figure*}
-{\bf \Large A}
-
-\includegraphics[width=0.9\linewidth]{figs/dms_S_corr.pdf}
-\vspace{0.05in}
-
-{\bf \Large B}
-
-\includegraphics[width=0.9\linewidth]{figs/dms_nsp5_corr.pdf}
+\centering
+\includegraphics[width=0.75\linewidth]{figs/dms.png}
 \caption{
-Correlation of fitness effect estimates with experimental measurements from deep mutational scanning studies for {\bf (A)} spike and {\bf (B)} Mpro (nsp5).
-Each point is an amino-acid mutation, and the oragne text in the upper left shows the number of mutations and Pearson correlation coefficient.
+Correlation of amino-acid fitness estimates with experimental measurements from deep mutational scanning studies for {\bf (A)} spike and {\bf (B)} Mpro (nsp5).
+Each point is an amino-acid mutation, and the orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
 Data are taken from two independent deep mutational scanning studies for spike~\citep{cite}, and two studies for Mpro~\citep{cite}.
-Each sub-panel shows a different set of mutations (those that are measured in each of the two datasets being correlated), and uses a minimum expected count cutoff of 20.
-Interactive versions of these plots that enable subsetting on just mutations shown across all datasets and adjustment of the minimum expected count cutoff are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html}.
+Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment), and and only shows fitness estimates for mutations with at least 10 expected counts.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for interactive versions of these plots that enable subsetting on just mutations shared across all datasets, and adjustment of the minimum expected count cutoff .
 \label{fig:dms_corr}

### 2023-01-15 paper/paper.tex
add `E_heatmap` figure

@@
 \begin{figure*}
-{\bf \Large A}
-
-E protein mutation effects with minimum expected counts of 20 and \textbf{x} marking wildtype identity in Wuhan-Hu-1.
-
-\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view1.pdf}
-
-{\bf \Large B}
-
-E protein mutation effects with minimum expected counts of 5 and \textbf{x} marking wildtype identity in BQ.1.
-
-\includegraphics[width=\linewidth, trim={0 0 0 0.4in}, clip=true]{figs/E_view2.pdf}
+\includegraphics[width=\linewidth]{figs/E_heatmap.pdf}
 \caption{
 Effects of amino-acid mutations to E protein.
-{\bf (A)} and {\bf (B)} provide two views of the estimated mutation effects with different thresholds for the minimum expected counts and different references used to mark the ``wildtype'' identity, as indicated in the titles above the plots.
-For both panels, the area plot at top shows the average effects of mutations at each site, and the heatmap shows the effects of specific amino acids.
-An interactive version of this plot at \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} enables zooming, mouseovers, and exploration of the data.
-Comparable interactive plots for all other viral proteins are linked at \jdbcomment{create landing pad page with these links}.
-\label{fig:E}
+The area plot at top shows the average effects of mutations at each site, and the heatmap shows the effects of specific amino acids, with \textbf{x} denoting the amino-acid identity in the Wuhan-Hu-1 strain.
+Only mutations with at least 10 expected counts are shown.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} for an interactive version of this plot that enables zooming, mouseovers, and adjustment of the minimum expected count threshold.
+See the links at \url{https://jbloomlab.github.io/SARS2-mut-fitness} for comparable interactive plots for all other SARS-CoV-2 proteins.
+\label{fig:E_heatmap}
 }

### 2023-01-15 paper/paper.tex
update versions of all figures

@@
 \begin{figure*}
-{\bf \Large A}
-
-\includegraphics[width=0.8\linewidth]{figs/clade_fixed_muts_hist.pdf}
-\vspace{0.05in}
-
-{\bf \Large B}
-
-\includegraphics[width=\linewidth]{figs/clade_fixed_muts.pdf}
-
+\centering
+\includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
 \caption{
-Effects of amino-acid mutations that have fixed in at least one SARS-CoV-2 clade.
-{\bf (A)}
-Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1 and all mutations that are fixed in at least one of the clades (as defined by Nextclade\jdbcomment{cite}).
-\jdbcomment{add vertical line at zero to these plots.}
-{\bf (B)}
-Effects of individual fixed mutations faceted by whether they are in spike or another protein.
-Plots show estimates of effects of mutations made across all clades (including those that have fixed the mutation) and just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
-Estimates are only shown for mutations with expected counts of at least 20.
-Interactive versions of this plot that enable easier exploration of the data are at \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html}.
-\label{fig:fixed}
-}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=0.8\linewidth, trim={0 1.1in 0 0in}, clip=true]{figs/fitness_vs_terminal.pdf}
-\caption{
-Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
-Each point is an amino-acid mutation, and the orange text in the upper right give the number of mutations and the Pearson correlation coefficient.
-This plot shows only mutations with at least 20 expected counts and 5 actual counts.
-An interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene, is at \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html}.
-\jdbcomment{add same color coding as in other panels for synonymous, nonsynonymous, stop types}
-\label{fig:terminal}
+Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and all mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
+The vertical dashed line at zero indicates the effect of a neutral mutation.
+Estimates are only shown for mutations with expected counts of at least 10.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} for an interactive version of this plot that allows adjustment of the minimum expected count threshold.
+\label{fig:fixed_dist}
 }
@@
 
+\onecolumn
+\renewcommand{\thepage}{S\arabic{page}}
+\setcounter{page}{1}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\clearpage
+
+\section{Supplementary Material}
+
+\begin{figure*}[b]
+\includegraphics[width=\linewidth]{figs/fixed.pdf}
+\caption{
+Effects of individual mutations that fixed in at least one clade of SARS-CoV-2, faceted by whether they are in spike or another protein.
+Plots show estimates of effects of mutations made across all clades (including those that have fixed the mutation) and just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
+Estimates are only shown for mutations with expected counts of at least 10.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html} for an interactive version of this plot.
+\label{fig:fixed}
+}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\includegraphics[width=0.75\linewidth]{figs/terminal.png}
+\caption{
+Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
+Each point is an amino-acid mutation, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
+This plot shows only mutations with at least 10 expected counts and 5 actual counts.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
+\label{fig:terminal}
+}
+\end{figure*}
+
 \end{document}

### 2023-01-15 paper/paper.tex
first few paragraphs of Intro

@@
 Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new viral variants and targeting of therapeutics to sites where resistance is unlikely to emerge.
-However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and systematic deep mutational scanning has been applied to only two of the virus's proteins.
+However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two of the virus's proteins.
 Here we develop a new approach that leverage millions of publicly available SARS-CoV-2 sequences to estimate the effects of most amino-acid mutations accessible by single-nucleotide changes.
@@
 
-\lettrine[lines=2]{\color{color2}B}{}ackground. It's important to understand effects of individual mutations.
+\lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the repeated emergence of viral variants with enhanced transmissibility, escape from anti-viral therapeutics, or reduced recognition by immunity~\citep{?}.
+To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants and develop therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{?}.
+Both efforts require determining how specific mutations affect the fitness of the virus.
 
-Experiments have filled this role for some viral proteins (spike, Mpro), but not for majority of SARS-CoV-2 viral protein.
+Unfortunately, experimentally determining the effects of mutations remains challenging for most SARS-CoV-2 proteins.
+For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{?}, and enabled deep mutational scanning of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{?}.
+These experimental data have proven valuable for assessing new spike variants, and are informing development of new antibody therapeutics with greater resistance to escape~\citep{?}.
+But most other viral proteins lack comparably tractable laboratory assays, despite the fact that non-spike proteins clearly contribute to variant fitness~\citep{?} and are the targets of efforts to develop anti-viral drugs~\citep{?}.
+The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{?}, and those measurements were performed in yeast.
 

### 2023-01-15 paper/paper.tex
third Intro paragraph

@@
 
-An alternative is sequence-based estimates, which so far have focused largely on clade growth and identification of mutations that mediate immune escape or increase transmissibility \citep{obermeyer2022analysis,lee2022inferring}.
-Mutations that are repeatedly observed at the base of clades that grow faster than the background likely contribute to growth.
-Such adaptive mutations are only a small minority of all available mutations and which mutations are adaptive depends on the environment (immunity) and the genetic background.
-The great majority of mutations, however, have either small fitness effects (approximately neutral) or are deleterious to viral replication.
-Here, we describe an approach to quantify the fitness cost of these mutations based on comparing expected versus observed counts.
-
-\rancomment{older paper on flu where not counts but trajectories were used to partition sequences into good and bad \citep{strelkowa2021clonal}}
+An alternative to experiments is to estimate the effects of mutations by analyzing naturally occurring viral sequences.
+The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 viruses from millions of human infections~\citep{?}.
+So analyses of these sequences have focused largely on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
+The basic idea of these prior analyses is that mutations that repeatedly appear in clades that subsequently expand in relative frequency are likely beneficial to the virus.
+However, only a small minority of all possible mutations are beneficial, with the great majority being either nearly neutral or deleterious to viral fitness.
+For purposes such as identifying functionally constrained drug targets, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
+\jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
 

### 2023-01-15 paper/paper.tex
full draft of intro

@@
 \lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the repeated emergence of viral variants with enhanced transmissibility, escape from anti-viral therapeutics, or reduced recognition by immunity~\citep{?}.
-To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants and develop therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{?}.
+To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{?}.
 Both efforts require determining how specific mutations affect the fitness of the virus.
@@
 Unfortunately, experimentally determining the effects of mutations remains challenging for most SARS-CoV-2 proteins.
-For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{?}, and enabled deep mutational scanning of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{?}.
-These experimental data have proven valuable for assessing new spike variants, and are informing development of new antibody therapeutics with greater resistance to escape~\citep{?}.
+For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{?}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{?}.
+These experimental data are valuable for assessing new spike variants and informing development of antibody therapeutics with greater resistance to escape~\citep{?}.
 But most other viral proteins lack comparably tractable laboratory assays, despite the fact that non-spike proteins clearly contribute to variant fitness~\citep{?} and are the targets of efforts to develop anti-viral drugs~\citep{?}.
@@
 An alternative to experiments is to estimate the effects of mutations by analyzing naturally occurring viral sequences.
-The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 viruses from millions of human infections~\citep{?}.
-So analyses of these sequences have focused largely on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
-The basic idea of these prior analyses is that mutations that repeatedly appear in clades that subsequently expand in relative frequency are likely beneficial to the virus.
-However, only a small minority of all possible mutations are beneficial, with the great majority being either nearly neutral or deleterious to viral fitness.
+The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections~\citep{?}.
+So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
+The basic idea of these prior analyses is that mutations that repeatedly appear near the base of clades that expand in relative frequency are likely beneficial to the virus.
+However, only a small minority of all possible mutations are beneficial, with the great majority being either nearly neutral or deleterious.
 For purposes such as identifying functionally constrained drug targets, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
 \jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
+
+Here we develop a new approach that uses natural sequences to estimate the effects of mutations to SARS-CoV-2.
+Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur tens to hundreds of times along the observed phylogeny.
+We therefore first calculate the number of \emph{expected} occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 as estimated from four-fold degenerate synonymous sites~\citep{?}.
+We then compare these expected occurrences to the actual occurrences in the SARS-CoV-2 phylogeny to estimate the effect of each mutation.
+The resulting estimates are used to create maps of mutation-level functional constraint across each viral protein, which correlate well with existing deep mutational scanning data.
+Most viral proteins have regions that are under strong selective constraints.
+However the accessory proteins show only weak selection against amino-acid and even stop-codon mutations, with the main exception being strong selection against premature stop codons in most of ORF3a.
+Overall, our work demonstrates a new approach to determine the effects of mutations, and provides detailed maps of functional constraint across the SARS-CoV-2 proteome.
 

### 2023-01-15 paper/paper.tex
add refs to Intro

@@
 \begin{abstract}
-Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new viral variants and targeting of therapeutics to sites where resistance is unlikely to emerge.
+Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new variants and targeting of therapeutics to sites where resistance is unlikely to emerge.
 However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two of the virus's proteins.
-Here we develop a new approach that leverage millions of publicly available SARS-CoV-2 sequences to estimate the effects of most amino-acid mutations accessible by single-nucleotide changes.
+Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate the effects of mutations.
 We first calculate how many times each mutation is expected to occur along the SARS-CoV-2 phylogeny in the absence of selection. 
-We then compare these expected counts to the actual observed counts to estimate the effect of each mutation.
-Our sequence-based estimates correlate with experimental deep mutational scanning measurements nearly as well as independent experiments correlate with each other.
+We then compare these expected counts to the actual counts to estimate the effect of each mutation.
+These estimates correlate well with deep mutational scanning measurements.
 For most genes, synonymous mutations are nearly neutral, stop-codon mutations are deleterious, and amino-acid mutations have a range of effects.
 However, some viral accessory proteins are under little to no selection.
-To enable our results to be easily used in viral surveillance and drug development, we provide interactive visualizations of the effects of mutations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
-The framework we describe is applicable to any virus (or organism) for which the number of available sequences is sufficiently large that that each neutral mutation independently occurs many times.
+To enable our estimates to be easily used, we provide interactive visualizations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
+The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that each neutral mutation independently occurs many times.
 \end{abstract}
@@
 
-\lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the repeated emergence of viral variants with enhanced transmissibility, escape from anti-viral therapeutics, or reduced recognition by immunity~\citep{?}.
-To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{?}.
+\lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the repeated emergence of viral variants with enhanced transmissibility, escape from anti-viral therapeutics, or reduced recognition by immunity~\citep{harvey2021sars,abdool2021new}.
+To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants~\citep{degrace2022defining} and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{moghadasi2022,iketani2022multiple,hiscox2021shutting}.
 Both efforts require determining how specific mutations affect the fitness of the virus.
@@
 Unfortunately, experimentally determining the effects of mutations remains challenging for most SARS-CoV-2 proteins.
-For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{?}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{?}.
-These experimental data are valuable for assessing new spike variants and informing development of antibody therapeutics with greater resistance to escape~\citep{?}.
-But most other viral proteins lack comparably tractable laboratory assays, despite the fact that non-spike proteins clearly contribute to variant fitness~\citep{?} and are the targets of efforts to develop anti-viral drugs~\citep{?}.
-The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{?}, and those measurements were performed in yeast.
+For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
+These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
+But most other viral proteins lack comparably tractable laboratory assays, despite the fact that non-spike proteins clearly contribute to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and are the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
+The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{flynn2022,iketani2022functional}, and those measurements were performed in yeast.
 
 An alternative to experiments is to estimate the effects of mutations by analyzing naturally occurring viral sequences.
-The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections~\citep{?}.
+The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections.
 So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
@@
 Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur tens to hundreds of times along the observed phylogeny.
-We therefore first calculate the number of \emph{expected} occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 as estimated from four-fold degenerate synonymous sites~\citep{?}.
+We therefore first calculate the number of \emph{expected} occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 as estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
 We then compare these expected occurrences to the actual occurrences in the SARS-CoV-2 phylogeny to estimate the effect of each mutation.
-The resulting estimates are used to create maps of mutation-level functional constraint across each viral protein, which correlate well with existing deep mutational scanning data.
+The resulting estimates are used to create maps of mutation-level functional constraint across each viral protein, and correlate well with existing deep mutational scanning data.
 Most viral proteins have regions that are under strong selective constraints.

### 2023-01-16 paper/paper.tex
first two paragraphs results

@@
 Unfortunately, experimentally determining the effects of mutations remains challenging for most SARS-CoV-2 proteins.
-For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
+For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape,harvey2021sars}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
 These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
-But most other viral proteins lack comparably tractable laboratory assays, despite the fact that non-spike proteins clearly contribute to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and are the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
+But most other viral proteins lack comparably tractable laboratory assays, despite clearly contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
 The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{flynn2022,iketani2022functional}, and those measurements were performed in yeast.
 
-An alternative to experiments is to estimate the effects of mutations by analyzing naturally occurring viral sequences.
+An alternative to experiments is to estimate the effects of mutations by analyzing natural viral sequences.
 The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections.
 So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
-The basic idea of these prior analyses is that mutations that repeatedly appear near the base of clades that expand in relative frequency are likely beneficial to the virus.
+The basic idea is that mutations that repeatedly appear near the base of clades that increase in relative frequency are likely beneficial to the virus.
 However, only a small minority of all possible mutations are beneficial, with the great majority being either nearly neutral or deleterious.
 For purposes such as identifying functionally constrained drug targets, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
-\jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
+% \jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
 
-Here we develop a new approach that uses natural sequences to estimate the effects of mutations to SARS-CoV-2.
-Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur tens to hundreds of times along the observed phylogeny.
-We therefore first calculate the number of \emph{expected} occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 as estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
-We then compare these expected occurrences to the actual occurrences in the SARS-CoV-2 phylogeny to estimate the effect of each mutation.
-The resulting estimates are used to create maps of mutation-level functional constraint across each viral protein, and correlate well with existing deep mutational scanning data.
+Here we develop a new approach that uses natural sequences to estimate the effects of mutations.
+Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur many times along the observed phylogenetic tree.
+We therefore first calculate the number of expected occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
+We then compare these expected occurrences to the actual occurrences in the SARS-CoV-2 tree to estimate the effect of each mutation.
+The resulting estimates correlate well with existing deep mutational scanning data.
 Most viral proteins have regions that are under strong selective constraints.
-However the accessory proteins show only weak selection against amino-acid and even stop-codon mutations, with the main exception being strong selection against premature stop codons in most of ORF3a.
+However the accessory proteins mostly show only weak selection against amino-acid and even stop-codon mutations.
 Overall, our work demonstrates a new approach to determine the effects of mutations, and provides detailed maps of functional constraint across the SARS-CoV-2 proteome.
@@
 
-\subsection{Theory}
+\subsection{Estimating fitness from actual versus expected mutation counts}
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual};][]{bloom2022evolution}.
+Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
+The expected counts of a mutation from nucleotide $x$ to $y$ is simply the average count of this type of mutation across all four-fold degenerate sites with parental identity $x$.
+Importantly, we count independent \emph{occurrences} of each mutation along the branches of the tree, not the sequences with the mutation in the final alignment (Figure~\ref{fig:expected_vs_actual}A.
+We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\cite{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
+
+The expected counts (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
+This wide variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.
+However, as Figure~\ref{fig:expected_vs_actual}B shows, in the absence of selection we expect to observe multiple independent occurrences of all nucleotide mutations at each site.
+
 \jdbcomment{Include Richard's calculations, either in entirety or summarized with details in Methods / Appendix}
@@
 The number of expected counts of each type of nucleotide mutation is computed from four-fold degenerate sites, and then compared the actual counts of each mutation.
-The counts are of independent occurrences of each mutation on the tree, not mutations in the alignment of observed sequences (a single occurrence of a mutation can be present in many sequences in an alignment).
 {\bf (B)}
 Expected versus actual counts for each nucleotide mutation type averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
-Synonymous mutations are often roughly neutral, so the expected and actual counts are similar.
-Nonsynonymous and especially stop-codon mutations are often deleterious, so the actual counts are generally less than the expected counts.
-Due to the uneven mutation spectrum of SARS-CoV-2, some nucleotide mutation types are expected to occur more often than others.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html} for an interactive version of panel B.

### 2023-01-16 paper/paper.tex
finished draft of first results section

@@
 \subsection{Estimating fitness from actual versus expected mutation counts}
-To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual};][]{bloom2022evolution}.
-Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
-The expected counts of a mutation from nucleotide $x$ to $y$ is simply the average count of this type of mutation across all four-fold degenerate sites with parental identity $x$.
-Importantly, we count independent \emph{occurrences} of each mutation along the branches of the tree, not the sequences with the mutation in the final alignment (Figure~\ref{fig:expected_vs_actual}A.
-We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\cite{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
-
-The expected counts (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
-This wide variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.
-However, as Figure~\ref{fig:expected_vs_actual}B shows, in the absence of selection we expect to observe multiple independent occurrences of all nucleotide mutations at each site.
-
-\jdbcomment{Include Richard's calculations, either in entirety or summarized with details in Methods / Appendix}
 
@@
 \end{figure*}
+
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual};][]{bloom2022evolution}.
+Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
+The expected counts of a mutation from nucleotide $x$ to $y$ is simply the average count of this type of mutation across all four-fold degenerate sites with parental identity $x$.
+Importantly, we count independent \emph{occurrences} of each mutation along the branches of the tree, not the sequences with the mutation in the final alignment (Figure~\ref{fig:expected_vs_actual}A.
+We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\cite{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
+
+The expected counts (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
+This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.
+
+We then compared the expected counts to the actual observed counts of mutations averaged across sites (Figure~\ref{fig:expected_vs_actual}).
+For synonymous mutations, the expected and actual counts are similar.
+But for nonsynonymous and especially stop-codon mutations, the actual counts are substantially lower than the expected counts, reflecting purifying selection for protein function.
+
+The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
+The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
+Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
+In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations will scale roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
+We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
+The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends providing links to interactive plots that enable adjustment of this threshold.
 

### 2023-01-17 paper/paper.tex
a few tweaks to first results subsection

@@
 
-To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual};][]{bloom2022evolution}.
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
 Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
@@
 Importantly, we count independent \emph{occurrences} of each mutation along the branches of the tree, not the sequences with the mutation in the final alignment (Figure~\ref{fig:expected_vs_actual}A.
-We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\cite{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
+We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\citep{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
 
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\linewidth]{figs/corr.png}
+\caption{
+Correlations between amino-acid mutation fitness effect estimates made using {\bf (A)} different viral clades or {\bf (B)} sequences from only the USA or only England.
+Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html} for interactive plots that include additional viral clades, allow adjustment of the expected-count cutoff, and allow examination of just specific genes.
+\label{fig:corr}
+}
+\end{figure*}
+
 The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
@@
 Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
-In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations will scale roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
+In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations scales roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
 We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
 The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends providing links to interactive plots that enable adjustment of this threshold.
+
+The estimated effects of mutations are robust to partitioning
+
+\subsection*{new subsection}
 
@@
 \label{fig:dist}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=\linewidth]{figs/corr.png}
-\caption{
-Correlations between amino-acid fitness estimates made using different sequence subsets.
-{\bf (A)} Estimates made using only sequences from the indicated viral clades, or {\bf (B)} estimates made using only sequences from the USA or England.
-Each point is a different amino-acid mutation, and the orange text at upper left shows the number of mutations and Pearson correlation coefficient.
-Both panels show only mutations with expected counts of at least 10 in each subset of sequences.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html}.For interactive versions of these plots that show correlations among additional clade pairs, allow adjustment of the expected-count cutoff, and allow examination of mutations for just specific genes.
-\label{fig:corr}
 }

### 2023-01-19 paper/paper.tex
second results subsection

@@
 
-The expected counts (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
+The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
 This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\linewidth]{figs/corr.png}
-\caption{
-Correlations between amino-acid mutation fitness effect estimates made using {\bf (A)} different viral clades or {\bf (B)} sequences from only the USA or only England.
-Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html} for interactive plots that include additional viral clades, allow adjustment of the expected-count cutoff, and allow examination of just specific genes.
-\label{fig:corr}
-}
-\end{figure*}
-
 The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
@@
 We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
-The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends providing links to interactive plots that enable adjustment of this threshold.
+The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends linking to interactive plots that enable adjustment of this threshold.
 
-The estimated effects of mutations are robust to partitioning
+\subsection*{Mutation-effect estimates are robust to subsampling natural sequences, with some evidence of epistasis in spike}
 
-\subsection*{new subsection}
+\begin{figure*}
+\centering
+\includegraphics[width=\linewidth]{figs/corr.png}
+\caption{
+Correlations of mutation fitness effect estimates made using subsets of natural sequences.
+Correlations between estimates made {\bf (A)} just using sequences from the Delta or Omicron BA.5 clades or {\bf (B)} just from the USA or England.
+Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
+Only mutations with at least 10 expected counts are shown, which is why panels have different numbers of mutations shown (sequence subsets vary in size).
+{\bf (C)} Correlations between clade or geography subsets become higher with an increasingly large threshold for minimum expected counts.
+Spike mutations have a worse correlation when subsetting by viral clade (plot shows average correlation over all pairwise combinations of Delta, BA.1, BA.2, and BA.5), but not when subsetting by geography (USA or England).
+{\bf (D)} Correlations in estimated mutation effects decline for clades with higher protein divergence, with the effect most noticeable for spike since spike is more diverged among SARS-CoV-2 clades than other viral proteins.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_corr_chart.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/subset_corr_chart.html} for versions of A and B that include all viral clades with at least 500,000 total expected counts (summed across all mutations) and have other interactive options.
+\label{fig:corr}
+}
+\end{figure*}
+
+To assess the robustness of the estimated effects of individual amino-acid mutations, we computed the correlations among estimates made using subsets of SARS-CoV-2 sequences from different viral clades or geographic locations.
+These estimates were highly correlated, although there is modest variation in estimates across sequence subsets (Figure~\ref{fig:corr}A,B).
+
+The modest variation in mutation effects estimated from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
+It is not obvious why there should be real differences in mutation effects in geographic locations, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
+To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
+Correlations increased with this count threshold, consistent with a reduction in statistical noise when using larger mutation counts.
+But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
+The lower cross-clade correlation for spike appears to be due to epistatic shifts in mutation effects that arise as sequences diverge~\citep{starr2022shifting, pollock2012amino, shah2015contingency}, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
+However, for the rest of this paper we aggregate counts across clades, since for most purposes the benefits aggregating counts (which enables lower-noise estimates more mutations) outweighs the effect of modest epistatic shifts.
+
+\subsection*{Structural and non-structural proteins are under strong purifying selection, but many accessory genes are not}
 
@@
 
+Caveats include synonymous selection.
+
 {\small

### 2023-01-19 paper/paper.tex
add text on accessory proteins

@@
 \begin{abstract}
-Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform risk assessment of new variants and targeting of therapeutics to sites where resistance is unlikely to emerge.
+Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and basic understanding of the function of viral proteins.
 However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two of the virus's proteins.
@@
 For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape,harvey2021sars}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
-These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
+These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody,tzou2022coronavirus} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
 But most other viral proteins lack comparably tractable laboratory assays, despite clearly contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
@@
 The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections.
-So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring}.
+So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
 The basic idea is that mutations that repeatedly appear near the base of clades that increase in relative frequency are likely beneficial to the virus.
-However, only a small minority of all possible mutations are beneficial, with the great majority being either nearly neutral or deleterious.
-For purposes such as identifying functionally constrained drug targets, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
+However, only a small minority of all possible mutations are beneficial, with most being nearly neutral or deleterious.
+For purposes such as identifying functionally constrained drug targets or understanding the function of viral proteins, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
 % \jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
@@
 
-We then compared the expected counts to the actual observed counts of mutations averaged across sites (Figure~\ref{fig:expected_vs_actual}).
+We compared the expected counts to the actual observed counts of mutations averaged across sites (Figure~\ref{fig:expected_vs_actual}).
 For synonymous mutations, the expected and actual counts are similar.
@@
 We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
-The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show only mutations with $\ge10$ expected counts, with the legends linking to interactive plots that enable adjustment of this threshold.
+The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show mutations with $\ge10$ expected counts unless otherwise noted, with legends linking to interactive plots that enable adjustment of this threshold.
 
@@
 The modest variation in mutation effects estimated from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
-It is not obvious why there should be real differences in mutation effects in geographic locations, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
+It is not obvious why there should be real differences in mutation effects by geographic location, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
 To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
@@
 The lower cross-clade correlation for spike appears to be due to epistatic shifts in mutation effects that arise as sequences diverge~\citep{starr2022shifting, pollock2012amino, shah2015contingency}, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
-However, for the rest of this paper we aggregate counts across clades, since for most purposes the benefits aggregating counts (which enables lower-noise estimates more mutations) outweighs the effect of modest epistatic shifts.
+However, for the rest of this paper we aggregate counts across clades, since for most purposes the benefits aggregating counts (which enables lower-noise estimates for more mutations) outweighs the effect of modest epistatic shifts.
 
-\subsection*{Structural and non-structural proteins are under strong purifying selection, but many accessory genes are not}
+\subsection*{Structural and non-structural proteins are under strong purifying selection, but most accessory proteins are not}
 
@@
 \caption{
-Distribution of effects of different types of mutations.
+Distribution of effects of different classes of mutations.
 {\bf (A)}
@@
 {\bf (B)}
-Effects of each type of mutation for each viral gene.
+Effects of each class of mutation for each viral gene.
 Dark squares indicate the median effect, and the lighter rectangles span the interquartile range.
 Mutation types are color-coded as in panel A.
-Both panels show only mutations with expected counts of at least 10.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for interactive versions of these plots that allow adjustment of the expected-count cutoff and other options (such as separate histograms for each gene).
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
 \label{fig:dist}
@@
 \end{figure*}
+
+The overall distributions of mutation effects concurs with basic biological intuition about how different classes of mutations impact protein function.
+Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
+To investigate if the spread in the overall distributions of mutations might reflect biological differences among the viral proteins, we also computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B and interactive version of Figure~\ref{fig:dist}A linked in figure legend).
+
+The SARS-CoV-2 proteins can be grouped into three categories: the nonstructural (or nsp) proteins, the structural proteins (spike, M, N, and E), and the accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
+The nonstructural proteins are essential components of the viral replication and transcript complex, and the structural proteins are essential for physical virion formation~\citep{v2021coronavirus}.
+Consistent with their essentiality, all nonstructural and structural proteins show strong selection against stop codons, and clear although variable selection against many amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+
+However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B).
+In particular, stop-codon and amino-acid mutations to ORF7a and ORF8 are not appreciably more deleterious than synonymous mutations.
+The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\cite{su2020discovery}.
+In fact, the only accessory protein under strong purifying selection against truncation is ORF3a (Figure~\ref{fig:dist}B), for which amino-acid mutations mostly have little effect but stop codons in the first ~240 residues (but not the last ~35) have clearly deleterious effects (Figure~\ref{fig:ORF3a}).
+These observations concur with experiments showing that SARS-CoV-2 is attenuated in human-ACE2 mice by deletion of ORF3a but there is little or no effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
+Whatever important functions are performed by the accessory proteins must be relatively insensitive to their amino-acid sequences: for instance, the function of ORF3a requires the first ~85\% of the protein, but beyond that there is little constraint outside a few important amino-acids at sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
+We note that observations such as these could help guide experimental or structural studies to better understand the functions of these proteins.
+
+Synonymous constraint...
+
+\subsection*{Mutation-effect estimates correlate well with deep mutational scanning}
 
@@
 Data are taken from two independent deep mutational scanning studies for spike~\citep{cite}, and two studies for Mpro~\citep{cite}.
-Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment), and and only shows fitness estimates for mutations with at least 10 expected counts.
+Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment).
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for interactive versions of these plots that enable subsetting on just mutations shared across all datasets, and adjustment of the minimum expected count cutoff .
@@
 \end{figure*}
+
+\subsection*{Interactive exploration of amino-acid fitnesses}
 

### 2023-01-20 paper/paper.tex
add text on protein-specific constraint

@@
 
-The overall distributions of mutation effects concurs with basic biological intuition about how different classes of mutations impact protein function.
+The overall distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
-To investigate if the spread in the overall distributions of mutations might reflect biological differences among the viral proteins, we also computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B and interactive version of Figure~\ref{fig:dist}A linked in figure legend).
 
-The SARS-CoV-2 proteins can be grouped into three categories: the nonstructural (or nsp) proteins, the structural proteins (spike, M, N, and E), and the accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
-The nonstructural proteins are essential components of the viral replication and transcript complex, and the structural proteins are essential for physical virion formation~\citep{v2021coronavirus}.
-Consistent with their essentiality, all nonstructural and structural proteins show strong selection against stop codons, and clear although variable selection against many amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+To investigate possible differences among viral proteins, we computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B).
+SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
+The nonstructural proteins form the essential viral replication and transcript complex, and the structural proteins form the essential physical virion~\citep{v2021coronavirus}.
+Consistent with their essentiality, all nonstructural and structural proteins show strong selection against stop codons, and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
 
 However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B).
-In particular, stop-codon and amino-acid mutations to ORF7a and ORF8 are not appreciably more deleterious than synonymous mutations.
-The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\cite{su2020discovery}.
-In fact, the only accessory protein under strong purifying selection against truncation is ORF3a (Figure~\ref{fig:dist}B), for which amino-acid mutations mostly have little effect but stop codons in the first ~240 residues (but not the last ~35) have clearly deleterious effects (Figure~\ref{fig:ORF3a}).
-These observations concur with experiments showing that SARS-CoV-2 is attenuated in human-ACE2 mice by deletion of ORF3a but there is little or no effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
-Whatever important functions are performed by the accessory proteins must be relatively insensitive to their amino-acid sequences: for instance, the function of ORF3a requires the first ~85\% of the protein, but beyond that there is little constraint outside a few important amino-acids at sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
-We note that observations such as these could help guide experimental or structural studies to better understand the functions of these proteins.
+Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations.
+The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery}.
+The only accessory protein under strong purifying selection against premature stop codons is ORF3a (Figure~\ref{fig:dist}B), for which amino-acid mutations mostly have little effect but stop codons in the first ~240 residues are clearly deleterious effects (Figure~\ref{fig:ORF3a}).
+These observations concur with experiments showing SARS-CoV-2 is attenuated in human-ACE2 mice by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
+Even ORF3a must perform functions that are relatively insensitive to its protein sequences, since there is little constraint beyond selection against stop codons in the first ~240 residues and a few important amino-acids at sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
+Observations such as these could help guide experimental and structural studies to understand protein function.
 

### 2023-01-20 paper/paper.tex
finished draft of protein selection part

@@
 
-To assess the robustness of the estimated effects of individual amino-acid mutations, we computed the correlations among estimates made using subsets of SARS-CoV-2 sequences from different viral clades or geographic locations.
+We computed the correlations among mutation-effect estimates made using subsets of SARS-CoV-2 sequences from different viral clades or geographic locations.
 These estimates were highly correlated, although there is modest variation in estimates across sequence subsets (Figure~\ref{fig:corr}A,B).
 
-The modest variation in mutation effects estimated from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
+The modest variation in estimates from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
 It is not obvious why there should be real differences in mutation effects by geographic location, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
 To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
-Correlations increased with this count threshold, consistent with a reduction in statistical noise when using larger mutation counts.
+Correlations increased with this count threshold, consistent with a reduction in statistical noise with larger mutation counts.
 But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
@@
 
-The overall distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
+The distributions of mutation effects concur with biological intuition about how different classes of mutations impact protein function.
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
 
-To investigate possible differences among viral proteins, we computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B).
+To investigate differences among viral proteins, we computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B).
 SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
-The nonstructural proteins form the essential viral replication and transcript complex, and the structural proteins form the essential physical virion~\citep{v2021coronavirus}.
-Consistent with their essentiality, all nonstructural and structural proteins show strong selection against stop codons, and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
 
@@
 The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery}.
-The only accessory protein under strong purifying selection against premature stop codons is ORF3a (Figure~\ref{fig:dist}B), for which amino-acid mutations mostly have little effect but stop codons in the first ~240 residues are clearly deleterious effects (Figure~\ref{fig:ORF3a}).
-These observations concur with experiments showing SARS-CoV-2 is attenuated in human-ACE2 mice by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
-Even ORF3a must perform functions that are relatively insensitive to its protein sequences, since there is little constraint beyond selection against stop codons in the first ~240 residues and a few important amino-acids at sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
-Observations such as these could help guide experimental and structural studies to understand protein function.
-
-Synonymous constraint...
+The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
+These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
+However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
+Observations such as these could help guide experimental studies to better understand protein function.
 
@@
 
+\begin{figure*}[h]
+\includegraphics[width=\linewidth]{figs/ORF3a.pdf}
+\caption{
+Effects of stop-codon and amino-acid mutations across ORF3a.
+The black area plot shows the mean effect of all amino-acid mutations at each site, and the purple points show the effects of stop codon mutations.
+There is strong selection against stop codons (negative effects) for all but the C-terminus of ORF3a, but only a few positions show strong selection against amino-acid substitutions.
+This plot shows only mutations with 20 expected counts.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/ORF3a.html} for an interactive version of this plot along with zoomable heatmap of the effects of specific amino-acid substitutions.
+\label{fig:ORF3a}
+}
+\end{figure*}
+
 \begin{figure*}[b]

### 2023-01-21 paper/paper.tex
revise results up to DMS section

@@
 {\bf (B)}
-Expected versus actual counts for each nucleotide mutation type averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html} for an interactive version of panel B.
+Expected versus actual counts for each nucleotide mutation type aggregated across all viral clades and averaged across all sites where the mutation is four-fold degenerate, synonymous (including four-fold degenerate), nonsynonymous, or introduces a stop codon.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/avg_counts.html} for an interactive version of panel B that enables mouseovers to read off specific values.
 \label{fig:expected_vs_actual}
@@
 We computed the correlations among mutation-effect estimates made using subsets of SARS-CoV-2 sequences from different viral clades or geographic locations.
-These estimates were highly correlated, although there is modest variation in estimates across sequence subsets (Figure~\ref{fig:corr}A,B).
+These estimates were well correlated, with some modest variation in estimates across sequence subsets (Figure~\ref{fig:corr}A,B).
 
-The modest variation in estimates from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real differences in effects of mutations among subsets of SARS-CoV-2.
-It is not obvious why there should be real differences in mutation effects by geographic location, but there is experimental evidence that epistasis has shifted effects of some spike mutations during viral evolution~\citep{starr2022shifting,moulana2022compensatory,mccallum2021n}.
+The modest variation in estimates from different sequence subsets could have two causes: statistical noise due to finite mutation counts, or real shifts in mutation effects during SARS-CoV-2 evolution~\citep{starr2022shifting,moulana2022compensatory}.
 To test for statistical noise, we computed correlations with different thresholds for how many expected counts are required before making an estimate for a mutation (Figure~\ref{fig:corr}C).
-Correlations increased with this count threshold, consistent with a reduction in statistical noise with larger mutation counts.
+Correlations increased with this count threshold, consistent with reduced statistical noise for larger mutation counts.
 But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
-The lower cross-clade correlation for spike appears to be due to epistatic shifts in mutation effects that arise as sequences diverge~\citep{starr2022shifting, pollock2012amino, shah2015contingency}, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
-However, for the rest of this paper we aggregate counts across clades, since for most purposes the benefits aggregating counts (which enables lower-noise estimates for more mutations) outweighs the effect of modest epistatic shifts.
+The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
+
+Despite evidence for some shifts in mutation effects in spike, for the rest of this paper we aggregate counts across viral clades to make a single estimate for each amino-acid mutation.
+The reason is that the accuracy of the estimates increases with the number of counts for each mutation (Figure~\ref{fig:corr}C), and several mutation types only have enough counts for reasonable estimates when aggregating across viral clades (Figure~\ref{fig:expected_vs_actual}).
+For the purposes of this paper, we deemed it preferable to have more accurate and comprehensive pan-SARS-CoV-2 mutation effect estimates than noisier clade-specific estimates for fewer mutations.
+However, the interactive version of Figure~\ref{fig:corr}A linked in the legend enables exploration of mutations with disparate estimates among clades.
 
@@
 
-\subsection*{Mutation-effect estimates correlate well with deep mutational scanning}
+\subsection*{Mutation-effect estimates correlate with deep mutational scanning}
 
@@
 \centering
-\includegraphics[width=0.75\linewidth]{figs/dms.png}
+\includegraphics[width=0.7\linewidth]{figs/dms.png}
 \caption{
-Correlation of amino-acid fitness estimates with experimental measurements from deep mutational scanning studies for {\bf (A)} spike and {\bf (B)} Mpro (nsp5).
-Each point is an amino-acid mutation, and the orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
-Data are taken from two independent deep mutational scanning studies for spike~\citep{cite}, and two studies for Mpro~\citep{cite}.
+Correlation of amino-acid fitness estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2022pseudovirus} or its receptor binding domain (RBD)~\citep{starr2022deep}, and {\bf (B)} Mpro (nsp5)~\citep{flynn2022,iketani2022functional}.
+Each point is an amino-acid mutation, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
 Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment).
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for interactive versions of these plots that enable subsetting on just mutations shared across all datasets, and adjustment of the minimum expected count cutoff .
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that enable subsetting on just mutations shared across all datasets and other interactive options.
 \label{fig:dms_corr}

### 2023-01-22 paper/paper.tex
draft of DMS subsection of results

@@
 
-\subsection*{Mutation-effect estimates correlate with deep mutational scanning}
+\subsection*{Mutation-effect estimates correlate with experiments}
 
@@
 \caption{
-Correlation of amino-acid fitness estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2022pseudovirus} or its receptor binding domain (RBD)~\citep{starr2022deep}, and {\bf (B)} Mpro (nsp5)~\citep{flynn2022,iketani2022functional}.
+Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2022pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2022,iketani2022functional}.
 Each point is an amino-acid mutation, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
 Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment).
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that enable subsetting on just mutations shared across all datasets and other interactive options.
-\label{fig:dms_corr}
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that enable subsetting on just mutations shared across all datasets and other interactive options such as mousing over points to see mutation identities.
+\label{fig:dms}
 }
 \end{figure*}
+
+We examined how the mutation effects estimated using our approach compare with prior high-throughput deep mutational scanning measurements.
+For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD) to measure ACE2 affinity and RBD expression~\citep{starr2020deep,starr2022deep}, and pseudotyped lentiviruses to measure spike-mediated infection in cell culture~\citep{dadonaite2022pseudovirus}.
+For Mpro (also known as nsp5 or 3CLpro), two different labs have performed deep mutational scanning using the same basic methodology of assaying protease cleavage in yeast~\citep{flynn2022,iketani2022functional}.
+
+For spike, our estimates from natural sequences correlate with each set experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A).
+If we increase the minimum expected counts from 10 to 20 and subset only on mutations shared among all three data sets, then the correlations between the estimates and experiments ($r = 0.66$) become even closer to the cross-experiment correlations ($r = 0.72$; see interactive version of Figure~\ref{fig:dms}A linked in legend).
+Some of the mutations with the greatest divergence between our sequence-based estimates and the deep mutational scanning likely represent experimental artifacts.
+For instance, P527L, which is favorable in the RBD deep mutational scan but deleterious in the sequence-based estimates and full-spike scan, is at the C-terminus of the yeast-displayed RBD~\citep{starr2020deep} where it may adopt a non-native conformation.
+
+The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the two experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
+Because both Mpro experiments use a similar yeast-based methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other versus the sequence-based estimates reflects shared artifacts of the yeast-based methodology across experiments.
+In particular, some of the discrepancies appear to reflect the fact that the experiments assay Mpro outside the context of the full SARS-CoV-2 polyprotein.
+For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would of course be highly deleterious to actual virus as it would truncate the polyprotein.
+Similarly, K61N is well tolerated in the deep mutational scanning but extremely disfavorable in our estimates, possibly because in the context of the full viral polyprotein this residue helps mediate important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
 

### 2023-01-23 paper/paper.tex
draft of Discussion

@@
 These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody,tzou2022coronavirus} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
-But most other viral proteins lack comparably tractable laboratory assays, despite clearly contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
-The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{flynn2022,iketani2022functional}, and those measurements were performed in yeast.
+But most other SARS-CoV-2 proteins lack comparably tractable laboratory assays, despite clearly contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
+The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{flynn2022,iketani2022functional}.
 
@@
 The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
-This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation}.
+This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation,neher2022}.
 
@@
 Despite evidence for some shifts in mutation effects in spike, for the rest of this paper we aggregate counts across viral clades to make a single estimate for each amino-acid mutation.
-The reason is that the accuracy of the estimates increases with the number of counts for each mutation (Figure~\ref{fig:corr}C), and several mutation types only have enough counts for reasonable estimates when aggregating across viral clades (Figure~\ref{fig:expected_vs_actual}).
+The reason is that the accuracy of the estimates increases with the number of counts for each mutation (Figure~\ref{fig:corr}C), and several mutation types only have enough counts for reasonable estimates when aggregating across viral clades (Figure~\ref{fig:expected_vs_actual}B).
 For the purposes of this paper, we deemed it preferable to have more accurate and comprehensive pan-SARS-CoV-2 mutation effect estimates than noisier clade-specific estimates for fewer mutations.
@@
 We examined how the mutation effects estimated using our approach compare with prior high-throughput deep mutational scanning measurements.
-For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD) to measure ACE2 affinity and RBD expression~\citep{starr2020deep,starr2022deep}, and pseudotyped lentiviruses to measure spike-mediated infection in cell culture~\citep{dadonaite2022pseudovirus}.
+For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD)~\citep{starr2020deep,starr2022deep} and spike pseudotyped lentiviruses~\citep{dadonaite2022pseudovirus}.
 For Mpro (also known as nsp5 or 3CLpro), two different labs have performed deep mutational scanning using the same basic methodology of assaying protease cleavage in yeast~\citep{flynn2022,iketani2022functional}.
 
-For spike, our estimates from natural sequences correlate with each set experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A).
+For spike, our estimates from natural sequences correlate with each set of experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A).
 If we increase the minimum expected counts from 10 to 20 and subset only on mutations shared among all three data sets, then the correlations between the estimates and experiments ($r = 0.66$) become even closer to the cross-experiment correlations ($r = 0.72$; see interactive version of Figure~\ref{fig:dms}A linked in legend).
@@
 The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the two experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
-Because both Mpro experiments use a similar yeast-based methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other versus the sequence-based estimates reflects shared artifacts of the yeast-based methodology across experiments.
-In particular, some of the discrepancies appear to reflect the fact that the experiments assay Mpro outside the context of the full SARS-CoV-2 polyprotein.
-For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would of course be highly deleterious to actual virus as it would truncate the polyprotein.
-Similarly, K61N is well tolerated in the deep mutational scanning but extremely disfavorable in our estimates, possibly because in the context of the full viral polyprotein this residue helps mediate important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
+Because both Mpro experiments use a similar methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other versus the sequence-based estimates reflects shared artifacts of the yeast-based assay across experiments.
+For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would be highly deleterious to actual virus as it would truncate the polyprotein.
+Similarly, K61N is well tolerated in the deep mutational scanning but extremely disfavorable in our estimates, possibly because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
 
@@
 The area plot at top shows the average effects of mutations at each site, and the heatmap shows the effects of specific amino acids, with \textbf{x} denoting the amino-acid identity in the Wuhan-Hu-1 strain.
-Only mutations with at least 10 expected counts are shown.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} for an interactive version of this plot that enables zooming, mouseovers, and adjustment of the minimum expected count threshold.
-See the links at \url{https://jbloomlab.github.io/SARS2-mut-fitness} for comparable interactive plots for all other SARS-CoV-2 proteins.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/E.html} for an interactive version of this plot that enables zooming, mouseovers, adjustment of the minimum expected count threshold, and layering of stop codon effects on the site plot.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness} for comparable interactive plots for all SARS-CoV-2 proteins.
 \label{fig:E_heatmap}
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
-\caption{
-Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and all mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
-The vertical dashed line at zero indicates the effect of a neutral mutation.
-Estimates are only shown for mutations with expected counts of at least 10.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} for an interactive version of this plot that allows adjustment of the minimum expected count threshold.
-\label{fig:fixed_dist}
-}
-\end{figure*}
+To enable easy access to the mutation-effect estimates, we created an interactive plots to enable exploration of the data for each protein.
+A static view of one of these plots is in Figure~\ref{fig:E_heatmap}; see \url{https://jbloomlab.github.io/SARS2-mut-fitness} for interactive versions for all proteins.
+These plots enable both high-level inspection of functional constraint across each protein, and detailed interrogation of the effects of specific mutations.
 
@@
 
-Caveats include synonymous selection.
+We have described a new approach that estimates effects of mutations to SARS-CoV-2 by leveraging the fact that every single-nucleotide change has independently occurred many times along the evolutionary history of sequenced viruses.
+In essence, we treat natural evolution as a deep mutational scan, with the millions of publicly available SARS-CoV-2 sequences from around the world providing a readout of this experiment.
+The key is simply to calculate how many times each mutation has been ``tested'' along the history of sampled viral sequences, and compare that expectation to the actual counts of the mutation among viruses sufficiently fit to have been sequenced in actual human infections.
+
+The resulting estimates of mutational effects are robust to subsetting on specific viral clades or geographies, and correlate well with prior experiments.
+In broad strokes, the mutation effects illuminate patterns of constraint: for instance, while there is strong selection on structural and non-structural proteins, there is only limited purifying selection on the accessory proteins.
+
+However, the real value of our approach is in the detailed maps of effects of specific mutations to all viral proteins, including proteins with important but poorly understood functions that are not easily characterized in the lab.
+These maps will be of value for designing drugs that target constrained sites, interpreting the consequences of mutations observed during viral surveillance, and guiding experiments to better understand the functions of viral proteins.
+
+There are several caveats to our approach.
+First, because the number of occurrences of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach is sensitive to sequencing (or base-calling) errors as low as ~0.001\% per site.
+Here we have attempted to remove spurious mutations, but further quality control is an important area for future work.
+Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{?}.
+Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{?}. 
+Third, the exact relationship between the statistics we calculate and viral fitness depend on both the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
+Although we derive this relationship in the Methods, we do not adjust for sampling intensity and population dynamics across time and space when estimating mutation effects.
+Fourth, we make a single estimate for each mutation across all SARS-CoV-2, thereby neglecting the epistasis that has affected some mutations in at least the spike protein~\citep{?}.
+Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.
+
+Conceptually, our approach differs from prior methods that aim to identify beneficial mutations associated with viral clades that expand in relative frequency~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
+Those methods draw information primarily from what happens downstream from a mutation.
+In contrast, our approach treats all mutations equivalently regardless of whether they are on a tip node or at the base of a large clade.
+Our approach is probably better for estimating the effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
+In particular, clade size carries information beyond that contained in mutation counts alone (see Figure~\ref{fig:terminal}).
+Hopefully future work can combine the strengths of our approach and clade-growth methods for even better estimates of mutation effects.
+
+The power of the approach we have described will increase with growth in viral sequencing.
+SARS-CoV-2 is the first virus with enough sequences that every tolerated mutation is observed multiple independent times.
+As costs drop, it is easy to imagine a future with even more viral sequences.
+As this occurs, viral genomic sequencing---which has traditionally been primarily a tool to track evolution and spread---will also become a means to answer questions about the effects of mutations that have traditionally required laboratory experiments.
 
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
+\caption{
+Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and all mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
+The vertical dashed line at zero indicates the effect of a neutral mutation.
+Estimates are only shown for mutations with expected counts of at least 10.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} for an interactive version of this plot that allows adjustment of the minimum expected count threshold.
+\label{fig:fixed_dist}
+}
+\end{figure*}
+
 \begin{figure*}[b]

### 2023-01-23 paper/paper.tex
start Methods

@@
 \subsection{Code and data availability}
-Code and data are at \url{https://github.com/jbloomlab/SARS2-mut-fitness}.
+See the GitHub repository at \url{https://github.com/jbloomlab/SARS2-mut-fitness} for the computer code and processed data (eg, fitness estimates and mutation counts).
+That repository contains a README with links to specific data files as well as a description of the computational pipeline.
+The specific version of the repository used for this paper is tagged at \jdbcomment{add link to GitHub tag on final commit}.
+The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{?} with interactive plots rendered using \texttt{altair}~\citep{?}.
+
+The interactive plots are rendered at \url{https://jbloomlab.github.io/SARS2-mut-fitness} via GitHub pages.
+
+\subsection{Counting mutations along the phylogenetic tree}
+We counted occurrences of each mutation in each viral clade using the UShER pre-built mutation-annotated tree~\citep{turakhia2021ultrafast,lanfear2020} from Dec-18-2022 (\url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2/2022/12/18/public-2022-12-18.all.masked.nextclade.pangolin.pb.gz}), which contains all $\sim$6.5-million SARS-CoV-2 sequences that are available in public databases.
+To make these counts at a per-clade level, we first subsetted the mutation-annotated tree on all sequences for each Nexstrain clade~\citep{?}, and then used the matUtils program distributed with UShER to extract the nucleotide mutations on every branch of the each clade-subsetted mutation-annotated tree.
+For the analyses by geographic location (Figure~\ref{fig:corr}), we subsetted on all sequences that began with ``USA'' or ``England'' as these were the two locations with the most publicly available sequences.
+
+We then performed quality control by ignoring any branch that met any of the following criteria:
+\begin{itemize}
+\item it had more than four nucleotide mutations,
+\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence,
+\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022},
+\item it contained more than one nucleotide mutation to the same codon.
+\end{itemize}
+The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors, and the rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
+We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
+
+Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.
+
+\subsection{Calculation of expected and actual counts}
+
+\subsection{Estimation of amino-acid fitnesses}
+
+\subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
 
 \section{Acknowledgments}
-We thank Angie Hinrichs for assistance with resolving several questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{if we don't add as co-author}.
+We thank Angie Hinrichs for assistance with resolving questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{or add here as co-author}.
 \jdbcomment{Jesse to acknowledge AVIDD, CEIRR, and R01 grants.}
@@
 JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus.
-JDB receives payments as an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
+JDB receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
 

### 2023-01-24 paper/paper.tex
finish first methods section

@@
 
-To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
 Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
@@
 The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
-This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation,neher2022}.
+This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation,neher2022contributions}.
 
@@
 Here we have attempted to remove spurious mutations, but further quality control is an important area for future work.
-Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{?}.
-Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{?}. 
+Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
+Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}. 
 Third, the exact relationship between the statistics we calculate and viral fitness depend on both the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
 Although we derive this relationship in the Methods, we do not adjust for sampling intensity and population dynamics across time and space when estimating mutation effects.
-Fourth, we make a single estimate for each mutation across all SARS-CoV-2, thereby neglecting the epistasis that has affected some mutations in at least the spike protein~\citep{?}.
+Fourth, we make a single estimate for each mutation across all SARS-CoV-2, thereby neglecting the epistasis that has affected some mutations in at least the spike protein~\citep{starr2022shifting,moulana2022compensatory}.
 Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.
@@
 The specific version of the repository used for this paper is tagged at \jdbcomment{add link to GitHub tag on final commit}.
-The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{?} with interactive plots rendered using \texttt{altair}~\citep{?}.
+The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.
 
@@
 \subsection{Counting mutations along the phylogenetic tree}
-We counted occurrences of each mutation in each viral clade using the UShER pre-built mutation-annotated tree~\citep{turakhia2021ultrafast,lanfear2020} from Dec-18-2022 (\url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2/2022/12/18/public-2022-12-18.all.masked.nextclade.pangolin.pb.gz}), which contains all $\sim$6.5-million SARS-CoV-2 sequences that are available in public databases.
-To make these counts at a per-clade level, we first subsetted the mutation-annotated tree on all sequences for each Nexstrain clade~\citep{?}, and then used the matUtils program distributed with UShER to extract the nucleotide mutations on every branch of the each clade-subsetted mutation-annotated tree.
+We counted occurrences of each mutation in each viral clade using the UShER pre-built mutation-annotated tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} from Dec-18-2022 (\url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2/2022/12/18/public-2022-12-18.all.masked.nextclade.pangolin.pb.gz}), which contains all $\sim$6.5-million SARS-CoV-2 sequences that are available in public databases.
+To make these counts at a per-clade level, we first subsetted the mutation-annotated tree on all sequences for each Nexstrain clade~\citep{aksamentov2021nextclade}, retained only clades with at least $10^4$ sequences, and then used the matUtils program distributed with UShER to extract the nucleotide mutations on every branch of the each clade-subsetted mutation-annotated tree.
 For the analyses by geographic location (Figure~\ref{fig:corr}), we subsetted on all sequences that began with ``USA'' or ``England'' as these were the two locations with the most publicly available sequences.
@@
 \begin{itemize}
-\item it had more than four nucleotide mutations,
-\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence,
-\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022},
+\item it had more than four nucleotide mutations;
+\item it contained more than one nucleotide mutation that was a reversion to the Wuhan-Hu-1 reference sequence;
+\item it contained more than one nucleotide mutation that was a reversion to the founder sequence for that clade as provided at \url{https://raw.githubusercontent.com/neherlab/SC2_variant_rates/7e738194a8c6592082f1caa9a6ca70cb68289790/data/clade_gts.json} by \cite{neher2022contributions};
 \item it contained more than one nucleotide mutation to the same codon.
@@
 We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
+Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.
 
-Note also that the mutation-annotated tree does not include insertion or deletion mutations, and so we only consider (and make estimates for) point mutations.
+We then specified for exclusion certain mutations and sites that are prone to sequencing or base-calling errors.
+Specifically, we excluded
+\begin{itemize}
+\item the sites specified in Table S1 of \cite{turakhia2020stability} as being error prone;
+\item sites 5629, 6851, 7328, 28095, and 29362 since they had very high error rates in some clades;
+\item the problematic sites listed at \url{https://github.com/W-L/ProblematicSites_SARS-CoV2}, which are masked in the pre-built mutation-annotated tree;
+\item for each clade, the clade-specific sites listed in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/data/usher_masked_sites.yaml}, which are masked in the pre-built mutation-annotated tree;
+\item for each clade, any mutation that was a reversion from the clade founder to the Wuhan-Hu-1 reference, and the reverse complements of these mutations.
+\end{itemize}
+The last exclusion criteria is because some bioinformatics pipelines called low-coverage sites as reference.
+
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/mutation_counts/aggregated.csv} for the final counts of each nucleotide mutation in each clade; note that this file also contains excluded mutations.
 

### 2023-01-24 paper/paper.tex
paper draft (#13)
* minor tweaks to abstract

* clean up figure 2

* update corr figure

* add dms figure

* add `E_heatmap` figure

* update versions of all figures

* first few paragraphs of Intro

* third Intro paragraph

* full draft of intro

* add refs to Intro

* first two paragraphs results

* finished draft of first results section

* a few tweaks to first results subsection

* second results subsection

* add text on accessory proteins

* add text on protein-specific constraint

* finished draft of protein selection part

* revise results up to DMS section

* draft of DMS subsection

@@
 That repository contains a README with links to specific data files as well as a description of the computational pipeline.
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
 The specific version of the repository used for this paper is tagged at \jdbcomment{add link to GitHub tag on final commit}.
@@
 
-\subsection{Calculation of expected and actual counts}
+\subsection{Calculation of expected counts}
+To calculate the expected counts for each nucleotide mutation, we analyzed just the four-fold degenerate sites in each clade in an approach paralleling that of \cite{bloom2022evolution}.
+Specifically, we identify all non-excluded four-fold degenerate sites in each clade founder.
+We then count nucleotide mutations just at those sites in each clade, and calculate the expected per-site number of mutations from nucleotide $x$ to $y$ as the total number of $x$ to $y$ mutations at four-fold degenerate sites divided by the number of four-fold degenerate sites with $x$ as the parental identity.
+This analysis is done at the clade level for two reasons: referencing mutations to the clade founder (rather than the Wuhan-Hu-1 reference) limits problem with the approach that would arise at sites that substitute multiple times in the history of a sequence (since each clade is a relatively high-identity group multiple mutations at the same site within a clade are very rare), and because it is know that SARS-CoV-2 mutation rates vary somewhat among clades~\citep{bloom2022evolution,ruis2022mutational}.
+We only retain clades with at least 5000 mutations at four-fold degenerate sites in order to avoid inaccurate estimates of expected counts due to low sampling of mutations.
 
-\subsection{Estimation of amino-acid fitnesses}
+\subsection{Mutational effects from actual versus expected counts}
+To estimate the effects of mutations, we simply compare the expected counts of each nucleotide mutation to the actual counts in the pre-built mutation-annotated tree.
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv} for these expected versus actual counts on a per-clade basis; note that this file also includes counts at excluded sites.
+
+To estimate the effects of mutations, we first sum the counts of all non-excluded nucleotide mutations that encode each amino-acid mutation to convert the nucleotide counts to amino-acid counts.
+In doing this, we exclude any mutations that are not from the clade-founder amino-acid identity: in other words, we ignore sequences with histories that involve multiple amino-acid mutations at the same site in the same clade (this is a caveat of the approach, although because each clade is relatively high identity it does not have a major effect).
+For the overall estimates reported in this paper, we also sum these counts across all retained clades; for the analyses in Figure~\ref{fig:corr} we also make estimates without summing across clades and only for counts from sequences from specific geographic locations.
+We then compute the estimated fitness $\Delta f$ of each mutation as simply the natural logarithm of the ratio of actual to expected counts after adding a pseudocount of $P - 0.5$ to each count, namely $ \Delta f = \log \left(\frac{n_{actual} + P}{n_{expected} + P}\right)$.
+
+Note that these mutation-effect estimates will have more statistical noise the smaller the value of the expected counts for each mutation.
+Therefore, we also track the expected counts alongside the estimates.
+In this paper, we only show estimates for mutations with expected counts of at least 10 unless otherwise noted.
+However, the figures link to interactive legends that allow adjustment of this threshold: larger values (eg, 20 or more) will lead to slightly more accurate estimates but drop some mutations, lower values can be used if you need a noisier estimate for a mutation that has less then 10 expected counts.
+
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aamut_fitness_all.csv} for the estimates of amino-acid mutation effects across all clades, and see \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aamut_fitness_by_clade.csv} for the clade-specific estimates.
+The all-clade estimates of mutation effects are what are shown in Figure~\ref{fig:dist}.
+
+For the clade correlations plotting in Figure~\ref{fig:corr}, we only include clades with at least $5 \times 10^5$ expected counts across all sites, as only these clades have enough counts for reasonable per-clade estimates.
+
+\subsection{Mutation effects to amino-acid fitnesses}
+For the final estimates of amino-acid fitnesses shown in the heatmaps such as in Figure~\ref{fig:E_heatmap}, we need a single estimate for each amino acid.
+This is straightforward for sites that have the same amino-acid identity in all clade founders: the ``wildtype'' residue shared across all clades has a fitness of zero, and all other amino acids have fitnesses equal to the effect of mutating from the ``wildtype'' to that amino acid.
+However, for sites that change amino-acid identity between clade founders, things are more complicated and we need to take the extra step below.
+
+For each clade have estimated the change in fitness $\Delta f_{xy}$ caused by mutating a site from amino-acid $x$ to $y$, where $x$ is the amino acid in the clade founder sequence.
+For each such mutation, we also have $n_{xy}$ which is the number of expected mutations from the clade founder amino acid $x$ to $y$.
+These $n_{xy}$ values are important because they give some estimate of our "confidence" in the $\Delta f_{xy}$ values: if a mutation has high expected counts (large $n_{xy}$) then we can estimate the change in fitness caused by the mutation more accurately, and if $n_{xy}$ is small then the estimate will be much noisier.
+
+However, we would like to aggregate the data across multiple clades to estimate amino-acid fitness values at a site under the assumption that these are constant across clades.
+Things get complicated if not all clade founders have the same amino acid identity at a site.
+For instance, let's say at our site of interest, the clade founder amino acid is $x$ in one clade and $z$ in another clade.
+For each clade we then have a set of $\Delta f_{xy}$ and $n_{xy}$ values for the first clade (where $y$ ranges over the 20 amino acids, including stop codon, that aren't $x$), and another set of up to 20 $\Delta f_{zy}$ and $n_{zy}$ values for the second clade (where $y$ ranges over the 20 amino acids that aren't $z$).
+
+From these sets of mutation fitness changes, we'd like to estimate the fitness $f_x$ of each amino acid $x$, where the $f_x$ values satisfy $\Delta f_{xy} = f_y - f_x$ (in other words, a higher $f_x$ means higher fitness of that amino acid).
+When there are multiple clades with different founder amino acids at the site, there is no guarantee that we can find $f_x$ values that precisely satisfy the above equation since there are more $\Delta f_{xy}$ values than $f_x$ values and the $\Delta f_{xy}$ values may have noise (and is some cases even real shifts among clades due to epistasis).
+Nonetheless, we can try to find the $f_x$ values that come closest to satisfying the above equation.
+
+First, we choose one amino acid to have a fitness value of zero, since the scale of the $f_x$ values is arbitrary and there are really only 20 unique parameters among the 21 $f_x$ values (there are 21 amino acids since we consider stops, but we only measure differences among them, not absolute values).
+Typically if there was just one clade, we would set the wildtype value of $f_x = 0$ and then for mutations to all other amino acids $y$ we would simply have $f_y = \Delta f_{xy}$.
+However, when there are multple clades with different founder amino acids, there is no longer a well defined "wildtype".
+So we choose the most common non-stop parental amino-acid for the observed mutations and set that to zero.
+In other words, we find $x$ that maximizes $\sum_y n_{xy}$ and set that $f_x$ value to zero.
+
+Next, we choose the $f_x$ values that most closely match the measured mutation effects, weighting more strongly mutation effects with higher expected counts (since these should be more accurate).
+Specifically, we define a loss function as
+$$
+L = \sum_x \sum_{y \ne x} n_{xy} \left(\Delta f_{xy} - \left[f_y - f_x\right]\right)^2
+$$
+where we ignore effects of synonymous mutations (the $x \ne y$ term in second summand) because we are only examining protein-level effects.
+We then use numerical optimization to find the $f_x$ values that minimize that loss $L$.
+
+Finally, we would still like to report an equivalent of the $n_{xy}$ values for the $\Delta f_{xy}$ values that give us some sense of how accurately we have estimated the fitness $f_x$ of each amino acid.
+To do that, we tabulate $N_x = \sum_y \left(n_{xy} + n_{yx} \right)$ as the total number of mutations either from or to amino-acid $x$ as the "count" for the amino acid.
+Amino acids with larger values of $N_x$ should have more accurate estimates of $f_x$.
+
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} for these overall amino-acid fitness estimates.
+
+\subsection{Site numbering and protein naming}
+All sites are numbered according to the sequential Wuhan-Hu-1 reference numbering scheme, using the reference sequence at \url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/bigZips/wuhCor1.fa.gz}.
+The protein annotations are taken from the associated GTF at \url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/bigZips/genes/ncbiGenes.gtf.gz}.
+Those protein annotations refer to the polyproteins encoding the non-structural proteins as ORF1a and ORF1ab.
+To convert to from ORF1ab numbering/naming to the nsp-based naming (eg, nsp1, nsp2, etc) we use the conversions specified under ``orf1ab\_to\_nsps'' in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}, which are in turn taken from Theo Sanderson's annotations at \url{https://github.com/theosanderson/Codon2Nucleotide/blob/main/src/App.js}.
 
 \subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
+\jdbcomment{Jesse will add this from Richard's notes.}
 
@@
 \centering
+\includegraphics[width=0.75\linewidth]{figs/terminal.png}
+\caption{
+Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
+Each point is an amino-acid mutation, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
+This plot shows only mutations with at least 10 expected counts and 5 actual counts.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
+\label{fig:terminal}
+}
+\end{figure*}
+
+\begin{figure*}
+\centering
 \includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
 \caption{
+\jdbcomment{Right now this figure is not reference in text; should we keep it somehow?}
 Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and all mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
@@
 \caption{
+\jdbcomment{Right now this figure is not reference in text; should we keep it somehow?}
 Effects of individual mutations that fixed in at least one clade of SARS-CoV-2, faceted by whether they are in spike or another protein.
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\linewidth]{figs/terminal.png}
-\caption{
-Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
-Each point is an amino-acid mutation, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
-This plot shows only mutations with at least 10 expected counts and 5 actual counts.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
-\label{fig:terminal}
-}
-\end{figure*}
-
 \end{document}

### 2023-01-25 scripts/compare_dca_fitness.py
add rudimentary comparison of dca and fitness data

@@ -1,0 +1,40 @@
+import pandas as pd
+import numpy as np
+from collections import defaultdict
+import matplotlib.pyplot as plt
+
+remap_genes = {"Spike":"S", "ORF1a":"ORF1ab", "ORF1b":"ORF1ab",
+                "Nucleocapsid":"N", "Membrane":"M", "Envelope":"E"}
+
+def load_dca():
+    dca = pd.read_csv("data/dca_mutability.csv")
+    dca_mutability = {}
+    for i, row in dca.iterrows():
+        dca_mutability[(remap_genes.get(row["protein"], row["protein"]), row["position_protein"])] = row["mutability_score(DCA)"]
+    return dca_mutability
+
+
+
+def load_fitness():
+    return pd.read_csv("results/aa_fitness/aamut_fitness_all.csv")
+
+
+if __name__ == "__main__":
+    dca_mutability = load_dca()
+    fitness = load_fitness()
+    fitness = fitness.loc[fitness.expected_count>30]
+
+    average_fitness = fitness.groupby(["gene", "aa_site"])['delta_fitness'].mean()
+
+    linked_dca_fitness = defaultdict(dict)
+    for i, row in average_fitness.iteritems():
+        gene, aa_site = i
+        if (gene, aa_site) in dca_mutability and not pd.isna(dca_mutability[(gene, aa_site)]):
+            linked_dca_fitness[gene][aa_site] = (dca_mutability.get((gene, aa_site), np.nan), row)
+
+
+    for gene in linked_dca_fitness:
+        d = np.array(list(linked_dca_fitness[gene].values()))
+        print(gene, np.corrcoef(d.T)[0,1])
+        plt.figure()
+        plt.scatter(d[:,0], d[:,1])

### 2023-01-25 scripts/compare_dca_fitness.py
add labels to dca graphs

@@ -38,3 +38,6 @@ if __name__ == "__main__":
         print(gene, np.corrcoef(d.T)[0,1])
         plt.figure()
         plt.scatter(d[:,0], d[:,1])
+        plt.title(gene)
+        plt.xlabel('dca')
+        plt.ylabel('fitness')

### 2023-01-27 scripts/compare_dca_fitness.py
add comparison to DCA estimates (#15)
* add rudimentary comparison of dca and fitness data

* add labels to dca graphs

Co-authored-by: Richard Neher <richard.neher@unibas.ch>

@@ -1,0 +1,43 @@
+import pandas as pd
+import numpy as np
+from collections import defaultdict
+import matplotlib.pyplot as plt
+
+remap_genes = {"Spike":"S", "ORF1a":"ORF1ab", "ORF1b":"ORF1ab",
+                "Nucleocapsid":"N", "Membrane":"M", "Envelope":"E"}
+
+def load_dca():
+    dca = pd.read_csv("data/dca_mutability.csv")
+    dca_mutability = {}
+    for i, row in dca.iterrows():
+        dca_mutability[(remap_genes.get(row["protein"], row["protein"]), row["position_protein"])] = row["mutability_score(DCA)"]
+    return dca_mutability
+
+
+
+def load_fitness():
+    return pd.read_csv("results/aa_fitness/aamut_fitness_all.csv")
+
+
+if __name__ == "__main__":
+    dca_mutability = load_dca()
+    fitness = load_fitness()
+    fitness = fitness.loc[fitness.expected_count>30]
+
+    average_fitness = fitness.groupby(["gene", "aa_site"])['delta_fitness'].mean()
+
+    linked_dca_fitness = defaultdict(dict)
+    for i, row in average_fitness.iteritems():
+        gene, aa_site = i
+        if (gene, aa_site) in dca_mutability and not pd.isna(dca_mutability[(gene, aa_site)]):
+            linked_dca_fitness[gene][aa_site] = (dca_mutability.get((gene, aa_site), np.nan), row)
+
+
+    for gene in linked_dca_fitness:
+        d = np.array(list(linked_dca_fitness[gene].values()))
+        print(gene, np.corrcoef(d.T)[0,1])
+        plt.figure()
+        plt.scatter(d[:,0], d[:,1])
+        plt.title(gene)
+        plt.xlabel('dca')
+        plt.ylabel('fitness')

### 2023-01-27 scripts/compare_dca_fitness.py
add correlation plot with DCA mutabilities

@@ -7,7 +7,7 @@ remap_genes = {"Spike":"S", "ORF1a":"ORF1ab", "ORF1b":"ORF1ab",
                 "Nucleocapsid":"N", "Membrane":"M", "Envelope":"E"}
 
 def load_dca():
-    dca = pd.read_csv("data/dca_mutability.csv")
+    dca = pd.read_csv(snakemake.input.dca)
     dca_mutability = {}
     for i, row in dca.iterrows():
         dca_mutability[(remap_genes.get(row["protein"], row["protein"]), row["position_protein"])] = row["mutability_score(DCA)"]
@@ -16,7 +16,7 @@ def load_dca():
 
 
 def load_fitness():
-    return pd.read_csv("results/aa_fitness/aamut_fitness_all.csv")
+    return pd.read_csv(snakemake.input.fitness)
 
 
 if __name__ == "__main__":
@@ -32,12 +32,15 @@ if __name__ == "__main__":
         if (gene, aa_site) in dca_mutability and not pd.isna(dca_mutability[(gene, aa_site)]):
             linked_dca_fitness[gene][aa_site] = (dca_mutability.get((gene, aa_site), np.nan), row)
 
-
-    for gene in linked_dca_fitness:
+    fig, axes = plt.subplots(2, 5, figsize=(12, 5))
+    for i, gene in enumerate(linked_dca_fitness):
+        ax = axes.ravel()[i]
         d = np.array(list(linked_dca_fitness[gene].values()))
-        print(gene, np.corrcoef(d.T)[0,1])
-        plt.figure()
-        plt.scatter(d[:,0], d[:,1])
-        plt.title(gene)
-        plt.xlabel('dca')
-        plt.ylabel('fitness')
+        corr = np.corrcoef(d.T)[0,1]
+        ax.scatter(d[:,0], d[:,1], s=10, alpha=0.3)
+        ax.set_title(f"{gene} (r = {corr:.2f})")
+        ax.set_xlabel('DCA mutability')
+        ax.set_ylabel('mean fitness effect')
+    fig.tight_layout()
+    
+    fig.savefig(snakemake.output.plot)

### 2023-01-27 paper/paper.tex
fix Methods description of how mutations excluded

@@
 To estimate the effects of mutations, we first sum the counts of all non-excluded nucleotide mutations that encode each amino-acid mutation to convert the nucleotide counts to amino-acid counts.
-In doing this, we exclude any mutations that are not from the clade-founder amino-acid identity: in other words, we ignore sequences with histories that involve multiple amino-acid mutations at the same site in the same clade (this is a caveat of the approach, although because each clade is relatively high identity it does not have a major effect).
+In doing this, we exclude any mutations that are not from the clade-founder codon identity: in other words, we ignore sequences with histories that involve multiple mutations at the same codon in the same clade (this is a caveat of the approach, although because each clade is relatively high identity it does not have a major effect).
 For the overall estimates reported in this paper, we also sum these counts across all retained clades; for the analyses in Figure~\ref{fig:corr} we also make estimates without summing across clades and only for counts from sequences from specific geographic locations.

### 2023-01-27 paper/paper.tex
add mean log size to `terminal` plot

@@
 \caption{
-Relationship between fitness effects of mutations and ratio of counts on terminal (tip) branches versus non-terminal (internal) branches.
+Relationship between fitness effects of mutations and two measures of the number of descendants.
+At top is shown the log ratio of counts of the mutation on non-terminal (internal) to terminal (tip) branches; larger values indicate mutations more likely to be found in viruses that leave descendants.
+At bottom is shown the mean log number of tip descendants that share all the mutations on each branch containing the mutation of interest; larger values again indicate mutations more likely to be found in viruses that leave more descendants.
 Each point is an amino-acid mutation, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
@@
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
+The number of descendants is calculated using the ``leaves\_sharing\_mutations'' variable of the UShER mutation-annotated tree.
 \label{fig:terminal}

### 2023-01-27 paper/paper.tex
elaborate on Mpro / yeast DMS diffs

@@
 The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the two experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
-Because both Mpro experiments use a similar methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other versus the sequence-based estimates reflects shared artifacts of the yeast-based assay across experiments.
-For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would be highly deleterious to actual virus as it would truncate the polyprotein.
-Similarly, K61N is well tolerated in the deep mutational scanning but extremely disfavorable in our estimates, possibly because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
+Because both Mpro experiments use a similar yeast-based methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other than the sequence-based estimates reflects shared artifacts of the yeast experiments.
+In particular, some Mpro mutations estimated to have deleterious effects in natural sequences are well tolerated in the yeast experiments.
+This difference could arise if the yeast experiments only capture some of the constraints on Mpro in the context of actual virus.
+For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and would be highly deleterious to actual virus as it would truncate the polyprotein.
+Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, possibly because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
 

### 2023-01-27 paper/paper.tex
add Neher fitness calculations

@@
 Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
-In the Methods, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness effects of deleterious mutations scales roughly logarithmically with the ratio of actual to expected counts for mutations with effects greater than a few percent.
-We therefore estimate the effect of each mutation as the logarithm of this ratio after summing counts for all nucleotides that encode the relevant amino-acid.
-The statistical noise in the count ratios is greater for mutations with fewer expected counts: the figures in this paper show mutations with $\ge10$ expected counts unless otherwise noted, with legends linking to interactive plots that enable adjustment of this threshold.
+In the Methods and Appendix, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness cost of a deleterious mutation scales roughly inversely with the ratio of actual to expected counts for mutations with costs greater than a few percent.
+A key result from this derivation is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced then our approach would only be sensitive to highly deleterious mutations, but since the actual sampling intensity is $\sim$0.1--1\% the approach can detect mutations with more subtle effects that transmit with reduced efficiency.
+We quantify the effect of each mutation as the logarithm of the ratio of actual to expected counts after summing counts for all nucleotides that encode the relevant amino-acid.
+The statistical noise is greater for mutations with fewer expected counts: the figures in this paper show mutations with $\ge10$ expected counts unless otherwise noted, with legends linking to interactive plots that enable adjustment of this threshold.
 
@@
 \subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
-\jdbcomment{Jesse will add this from Richard's notes.}
+The ratio of actual to expected counts that we calculate in this paper is related to the probability that we observe a viral lineage containing an occurrence of a specific mutation among sequenced human SARS-CoV-2.
+This probability depends on three factors: the fitness effect of the mutation, the fraction of all SARS-CoV-2 viruses that are sequenced (sampling intensity), and the growth dynamics of the viral population.
+In the supplementary appendix, we derive the approximate relationship between this probability as a function of the fitness cost $s$ and sampling intensity $\epsilon$ for deleterious mutations for both a constant and exponentially growing viral population.
+
+We show that for a constant viral population size, the probability of observing a lineage containing a deleterious mutation with cost $s$ is roughly $\frac{\epsilon}{s + \epsilon}$ when $s^2 > \epsilon$, and more weakly dependent on $s$ for smaller fitness costs (when $s^2 < \epsilon$).
+The intuitive explanation is that the average size of a mutant lineage with fitness cost $s$ is $1/s$ and we basically ask whether we sample the lineage before it disappears.
+If we sample more intensely (larger $\epsilon$), whether a lineage gets sampled depends primarily on the stochastic dynamics and little on the fitness effect.
+With a typical sampling intensity for SARS-CoV-2 between 1/1000 and 1/100, this means our approach is sensitive to fitness effects larger than a few percent; mutations with fitness costs smaller than that will not show an appreciable difference from neutral mutations in their ratio of actual to expected accounts.
+
+In an exponentially growing population, the probability of observing a mutant lineage with fitness cost $s$ again scales as $\sim \frac{\epsilon}{\epsilon + s}$ if $sT>1$, where $T$ is the time over which the variant has expanded.
+If $T$ is $\sim$ months, that is 20 generations, which again corresponds to $s$ of at least a few percent for $sT > 1$. 
+For mutations with smaller fitness costs, the dependence scales more as $\sim \epsilon\left(1-sT\right)$.
+
+Overall, these calculations indicate that for multiple different growth dynamics of the viral population, the ratio of expected to actual counts will scale inversely with the fitness cost of deleterious mutations for mutations with costs that exceed a few percent.
+Note that the approach we use in this paper does not account for variation in sampling intensity across space or time, does not attempt to adjust for changes in viral growth dynamics over time, uses the heuristic formula of calculating the effect as the log ratio of counts, and applies this same formula to all mutations regardless of whether they are deleterious, neutral, or beneficial.
+A more complete derivation might try to calculate the fitness effects from the full distribution of lineage sizes more rigorously and incorporate information about the sampling intensity and viral growth dynamics.
+However, such a derivation (if possible at all) is beyond the scope of this study, and we also note that good empirical data is generally lacking to precisely account for sampling intensity and viral growth dynamics over the full span of time and space from which the sequences we analyze are drawn.
+The key point of the derivations for our current study is simply that our approach should be sensitive to detecting the effects of mutations with fitness costs greater than a few percent.
 
@@
 
+\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
 \onecolumn
@@
 
-\section{Supplementary Material}
+\section{Supplementary appendix deriving relationship between fitness cost and ratio of expected to actual counts}
+
+With millions of SARS-CoV-2 sequences shared publicly, almost all mutations that are tolerated by the virus are observed dozens to hundreds of times.
+Where on the tree and how often on the tree we observe specific mutations has information about the effects of these mutations on viral spread.
+The mutation rate depends on the nucleotides involved and possibly on the sequence context and other viral determinants, but for the purpose of this derivation, we will assume the neutral rate $\mu$ is known.
+If the mutation is neutral, the total number of times the mutation is observed on the tree is $\mu T$, where $T$ is the total length of the tree (assuming that the mutation never reached high frequency which is true for almost all mutations, particularly when mutations are counted on a per-clade basis relative too the clade founder as done above).
+
+If a mutation reduces fitness, the lineages descending from branches on which this mutation happened will spread more slowly than those without this mutation.
+As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
+To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on the this fitness effect.
+
+For a mutation to be represented in the tree, one of its descendants has to be sampled and sequenced.
+If the total number of descendants is $w$ and the sampling fraction is $\epsilon$, the probability that the mutation is present in the tree is
+\begin{equation}
+    P = 1 - e^{-w\epsilon}
+\end{equation}
+$W$ is a random number that depends on the realization of the transmission process, which is commonly modeled by a branching process with birth rate $b$ and death rate $d$.
+The death rate here corresponds to clearing an infection, the birth rate to onward transmission. The latter is affected by the fitness cost of the mutation.
+
+To obtain insight how the probability of observing a lineage depends on parameters, we calculate the probability $p(w, T|t)$ that a lineage had an integrated size $w = \int_t^T k(t') \; dt'$, where $t$ is the birth time of the lineage, $T$ is the current time, and $k(t')$ is the size of the lineage at time $t'$.
+To calculate $p(w,T|k)$, we generalize it slightly to $p(W, T|k,t)$, where $k$ is the number of individuals at the start time $t$. This quantity obeys the following ``first-step'' equation:
+\begin{equation}
+    -(\partial_t - k \partial_w) p(w,T|k,t) = -k(b+d) p(w,T|k,t) + k b p(w,T|k+1,t) + k d p(w,T| k-1, t)
+\end{equation}
+We will solve for the Laplace transform $\hat{p}(z,T|k,t) = \int_0^\infty dw\; e^{-wz} p(w,T|k,t) = \hat{p}^k(z,T|1,t)$.
+Using the identity for the derivative of the Laplace transform
+\begin{equation}
+    \begin{split}
+        \int_0^\infty e^{-wz} \partial_w p  \; dw& = [e^{-wz} p]_0^\infty - \int_0^\infty  p \partial_w e^{-wz} \; dw = 0 + z \int_0^\infty  p e^{-wz} = z \hat{p}
+    \end{split}
+\end{equation}
+and using $k=1$, we have
+\begin{equation}
+    -\partial_t \hat{p}(z,T|t) = -(b+d+z) \hat{p}(z,T|t) + b \hat{p}^2(z,T|t) + d
+\end{equation}
+This simplifies further to if we substitute $\phi(z,T|t) = 1 - \hat{p}(z,T|t)$.
+\begin{equation}
+    \begin{split}
+        \partial_t \phi(z,T|t) & = -(b+d+z) (1-\phi(z,T|t)) + b (1-\phi(z,T|t))^2 + d \\
+         & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
+    \end{split}
+\end{equation}
+where it is important to note that the derivative is with respect to the first time point and the interval $T-t$ is shrinking with increasing $t$.
+
+\subsection*{Constant birth and death rate}
+If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and $\partial_t \phi(z,T|t)$.
+Further define $b=b_0 - s$ and $d=d_0$ where $s$ is the fitness cost of the mutation (so larger values indicate a greater fitness cost).
+The steady state generating function is then
+\begin{equation}
+    0 = -z - (b-d-z) \phi(z) + b \phi(z)^2
+\end{equation}
+with solution
+\begin{equation}
+    \begin{split}
+        \phi(z) & = -\frac{s + z}{2(b_0 - s)} \pm \frac{\sqrt{(s+z)^2 + 4z(b_0-s)}}{2(b_0-s)} \\
+        &\approx -\frac{s + z}{2b_0} \pm \frac{\sqrt{(s+z)^2 + 4z b_0}}{2b_0} \\
+        & \approx \begin{cases}
+            \frac{z}{s+z} & (s+z)^2 \gg 4zb_0 \\
+            \sqrt{\frac{z}{b_0}}\left(1+\frac{(s+z)^2}{8zb_0}\right)-\frac{s + z}{2b_0}  & (s+z)^2 \ll 4zb_0 \\
+        \end{cases}
+    \end{split}
+\end{equation}
+Since $\phi(z) = 1 - \int e^{-wz}p(w) \; dw$, $\phi(\epsilon)$ is exactly the probability that a lineage is sampled when the entire population is sampled at rate $\epsilon$.
+We thus expect two regimes: if the square of the fitness cost exceeds the sampling intensity (typically at 1\% or less), the probability of sampling a lineage is essentially inversely proportional to the fitness cost.
+The sampling probability of lineages with smaller costs effects depends less strongly on $s$.
+Their sampling mostly comes down to stochasticity independent of the fitness cost.
+
+\subsection*{Growing populations}
+In many scenarios relevant for lineages that arise during a viral outbreak, the background population isn't constant but is undergoing a rapid exponential expansion.
+The background birth rate $b_0$ is bigger than $d_0$ in this case.
+Since the population is growing, deleterious mutations can increase in frequency deterministically and we can not send the $t$ to infinity as before.
+Instead, we need to integrate
+\begin{equation}
+    \begin{split}
+        \partial_t \phi(z,T|t) & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
+    \end{split}
+\end{equation}
+backwards in time starting from $\phi(z,T|T)=0$ at $t=T$.
+While $\phi(z,T|t)$ is small, this is approximately solved by
+\begin{equation}
+\begin{split}
+    \phi(z,T|t) & = z e^{\int_t^T (b-d-z)dt'} \int_t^T e^{-\int_\tau^T (b-d-z)dt'} d\tau \\
+    & = z e^{(b-d-z)(T-t)} \int_t^T e^{-(b-d-z)(T-\tau)} d\tau \\
+    & = z e^{(b-d-z)(T-t)}\left[1   - e^{-(b-d-z)(T-t)}\right]/(b-d-z) \\
+    & = \frac{z}{b-d-z}\left[e^{(b-d-z)(T-t)} - 1\right] = \frac{z}{\gamma_0 - s - z}\left[e^{\gamma_0 (T-t) - (s+z) (T-t)} - 1\right]
+\end{split}
+\end{equation}
+where $\gamma_0$ is the growth rate of the background population.
+
+At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
+This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
+Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
+\begin{equation}
+    \begin{split}
+        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
+        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
+        &\approx \begin{cases}
+            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
+            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
+        \end{cases}
+    \end{split}
+\end{equation}
+This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
+The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
+In this case, the above simplifies to
+\begin{equation}
+    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
+\end{equation}
+
+In a variant that has been growing with rate $\gamma_0$ for a time $\tau=T-t$ and sampled with $z=\epsilon$, we thus expect that the number of times we observe separate mutant lineages depends on $s$ as
+\begin{equation}
+    \langle \phi \rangle \approx \frac{\epsilon}{\gamma_0(\epsilon+s)}
+\end{equation}
+This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
+In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
+In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is large, i.e., if the fitness cost is substantial on this time scale.
+Outside this parameter range, there is still a dependence on $s$, but it is weaker.
+
+\clearpage
+
+\section{Supplementary figures}
 

### 2023-01-27 paper/paper.tex
add section on fixed mutations

@@
 \begin{abstract}
-Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and basic understanding of the function of viral proteins.
-However, experimentally determining the effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two of the virus's proteins.
+Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and understanding of the functions of viral proteins.
+However, experimentally determining effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two SARS-CoV-2 proteins.
 Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate the effects of mutations.
-We first calculate how many times each mutation is expected to occur along the SARS-CoV-2 phylogeny in the absence of selection. 
-We then compare these expected counts to the actual counts to estimate the effect of each mutation.
+We first calculate how many independent occurrences of each mutation are expected to be observed along the SARS-CoV-2 phylogeny in the absence of selection. 
+We then compare these expected observations to the actual observations to estimate the effect of each mutation.
 These estimates correlate well with deep mutational scanning measurements.
@@
 Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur many times along the observed phylogenetic tree.
-We therefore first calculate the number of expected occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
-We then compare these expected occurrences to the actual occurrences in the SARS-CoV-2 tree to estimate the effect of each mutation.
+We therefore first calculate the number of expected observations of independent occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
+We then compare these expected observations to the actual observations in the SARS-CoV-2 tree to estimate the effect of each mutation.
 The resulting estimates correlate well with existing deep mutational scanning data.
@@
 Correlations between estimates made {\bf (A)} just using sequences from the Delta or Omicron BA.5 clades or {\bf (B)} just from the USA or England.
-Each point is an amino-acid mutation, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
+Each point is an amino-acid mutation, the orange line is a least-squares regression, and orange text at upper left shows the number of mutations and Pearson correlation coefficient.
 Only mutations with at least 10 expected counts are shown, which is why panels have different numbers of mutations shown (sequence subsets vary in size).
@@
 However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B).
-Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations.
+Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
 The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery}.
@@
 Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2022pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2022,iketani2022functional}.
-Each point is an amino-acid mutation, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
+Each point is an amino-acid mutation, the orange line is a least-squares regression, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
 Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment).
@@
 Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, possibly because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
+
+\subsection*{Fixed mutations tend to have beneficial or neutral effects}
+Amino-acid mutations that have fixed in at least one viral clade are estimated to mostly have neutral or beneficial effects, whereas most other mutations are deleterious (Figure~\ref{fig:fixed_dist}).
+This fact is unsurprising: viral lineages that expand into new clades do so because they have acquired beneficial mutations while avoiding deleterious ones~\citep{luksza2014predictive, koelle2015effects, huddleston2020integrating}.
+But the fact that the beneficial effects of fixed mutations are correctly estimated by our approach, which simply counts mutation occurrences and does not incorporate information on lineage size, demonstrates such mutations occur independently in many viral lineages that are more successful than average.
+
+Most fixed mutations are estimated to be beneficial regardless of whether estimates are made using all viral clades, or just clades that did not fix the mutation (Figure~\ref{fig:fixed}).
+However, a few beneficial fixed mutations show epistatic entrenchment~\citep{shah2015contingency, starr2018pervasive} in the sense that they are not particularly beneficial in clades in which they did not fix (Figure~\ref{fig:fixed}).
+The most striking example is S373P in spike, which has experimentally been shown to be neutral or slightly deleterious in pre-Omicron clades, but strongly beneficial in the Omicron clades in which it fixed~\citep{starr2022deep, moulana2022compensatory}.
 
@@
 At bottom is shown the mean log number of tip descendants that share all the mutations on each branch containing the mutation of interest; larger values again indicate mutations more likely to be found in viruses that leave more descendants.
-Each point is an amino-acid mutation, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
+Each point is an amino-acid mutation, the orange line is a least-squares regression, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
 This plot shows only mutations with at least 10 expected counts and 5 actual counts.
@@
 \caption{
-\jdbcomment{Right now this figure is not reference in text; should we keep it somehow?}
-Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and all mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
+Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and just those mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
 The vertical dashed line at zero indicates the effect of a neutral mutation.
-Estimates are only shown for mutations with expected counts of at least 10.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} for an interactive version of this plot that allows adjustment of the minimum expected count threshold.
@@
 \caption{
-\jdbcomment{Right now this figure is not reference in text; should we keep it somehow?}
 Effects of individual mutations that fixed in at least one clade of SARS-CoV-2, faceted by whether they are in spike or another protein.
-Plots show estimates of effects of mutations made across all clades (including those that have fixed the mutation) and just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
-Estimates are only shown for mutations with expected counts of at least 10.
+``Mutation polarity'' indicates if the point shows the effect of the mutation estimated using all viral clades (including those that have fixed the mutation), or just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
+Some mutations are estimated to be more favorable when including clades in which they have fixed (blue circles) in addition to just clades in which it has not yet fixed (orange squares)---when this occurs, it suggests epistatic entrenchment of the mutations~\citep{shah2015contingency, starr2018pervasive}.
+Note that clades in which a mutation has already fixed contribute to estimates of its fitness via estimates of the effect of its reversion and via estimates of the effects of mutations to other amino acids at the same site.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html} for an interactive version of this plot.

### 2023-01-28 paper/paper.tex
updates to paper (#16)
* fix Methods description of how mutations excluded

* add mean log size to `terminal` plot

* elaborate on Mpro / yeast DMS diffs

* page numbers in center bottom

* add Neher fitness calculations

* remove `fitness_inference` notes as now in main paper

* add section on fixed mutations

* add DCA and Marks lab citations

@@
 For purposes such as identifying functionally constrained drug targets or understanding the function of viral proteins, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
+Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are inherently limited by the fact that mutational effects shift as proteins diverge~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep}.
 % \jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
@@
 But the correlation for spike mutations was consistently lower for cross-clade but not cross-geography comparisons (Figure~\ref{fig:corr}C).
-The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
+The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
 
@@
 \section{Acknowledgments}
-We thank Angie Hinrichs for assistance with resolving questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{or add here as co-author}.
-\jdbcomment{Jesse to acknowledge AVIDD, CEIRR, and R01 grants.}
+We thank Angie Hinrichs for assistance with questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{check with her to make sure acknowledging is OK}.
+The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.  
+\jdbcomment{Jesse needs to confirm permission of Chodera and Hensley to cite U19 and CEIRR contract.}
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2023-01-28 paper/paper.tex
edits and add Zanini reference

@@
 To enable our estimates to be easily used, we provide interactive visualizations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
-The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that each neutral mutation independently occurs many times.
+The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that many independent occurrences of each neutral mutation are observed.
 \end{abstract}
@@
 
+We previously estimated fitness costs of mutations in the HIV-1 genomes from deep sequencing data of within host diversity, where the frequency of a deleterious mutation in the viral quasi-species is determined by a balance between repeated mutation and purifying selection \citep{zanini2017vivo}.
+But in SARS-CoV-2, mutation frequency is a poor predictor of its fitness effect since moderately deleterious mutations can hitch-hike on emerging variants.
 Here we develop a new approach that uses natural sequences to estimate the effects of mutations.
@@
 In the Methods and Appendix, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness cost of a deleterious mutation scales roughly inversely with the ratio of actual to expected counts for mutations with costs greater than a few percent.
-A key result from this derivation is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced then our approach would only be sensitive to highly deleterious mutations, but since the actual sampling intensity is $\sim$0.1--1\% the approach can detect mutations with more subtle effects that transmit with reduced efficiency.
+A key result from this derivation is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced even very deleterious mutations would have a high chance of being sampled and we would need to study the subsequent spread of the mutations to assess their fitness.
+But since the actual sampling intensity is $\sim$0.1--1\% the number of times mutation is observed sensitively probes more subtle effects that transmit with reduced efficiency.
 We quantify the effect of each mutation as the logarithm of the ratio of actual to expected counts after summing counts for all nucleotides that encode the relevant amino-acid.
@@
 Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
-The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery}.
+The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
 The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
@@
 If we sample more intensely (larger $\epsilon$), whether a lineage gets sampled depends primarily on the stochastic dynamics and little on the fitness effect.
-With a typical sampling intensity for SARS-CoV-2 between 1/1000 and 1/100, this means our approach is sensitive to fitness effects larger than a few percent; mutations with fitness costs smaller than that will not show an appreciable difference from neutral mutations in their ratio of actual to expected accounts.
+With a typical sampling intensity for SARS-CoV-2 between 1/1000 and 1/100, this means our approach is sensitive to fitness effects larger than a few percent per serial interval; mutations with fitness costs smaller than that will not show an appreciable difference from neutral mutations in their ratio of actual to expected accounts.
 
@@
 As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
-To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on the this fitness effect.
+To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on this fitness effect.
 
@@
 We will solve for the Laplace transform $\hat{p}(z,T|k,t) = \int_0^\infty dw\; e^{-wz} p(w,T|k,t) = \hat{p}^k(z,T|1,t)$.
-Using the identity for the derivative of the Laplace transform
+Using the following identity for the derivative of the Laplace transform
 \begin{equation}
@@
 \end{equation}
-and using $k=1$, we have
+and setting $k=1$, we have
 \begin{equation}
@@
 \subsection*{Constant birth and death rate}
-If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and $\partial_t \phi(z,T|t)$.
+If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and the long time asymptotic $\partial_t \phi(z,T|t)=0$.
 Further define $b=b_0 - s$ and $d=d_0$ where $s$ is the fitness cost of the mutation (so larger values indicate a greater fitness cost).
@@
 backwards in time starting from $\phi(z,T|T)=0$ at $t=T$.
-While $\phi(z,T|t)$ is small, this is approximately solved by
+While $\phi(z,T|t)$ is small and the quadratic term can be neglected, this is approximately solved by
 \begin{equation}
@@
 This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
-The most relevant fitness effects will be those with $s(T-t)>1$ but $\gamma_0>s$ such that the mutant lineage has been amplified but its fitness difference to the background is a perturbation.
+The most relevant fitness effects will be those with $s(T-t)>1$, that is the fitness effect has strong effect on variant frequency, but $s<\gamma_0$ such that the variant is still spreading and can give rise to large lineages in an expanding variant.
 In this case, the above simplifies to
@@
 This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
+The same basic dependency is observed at steady state in a quasi-species \citep{zanini2017vivo}.
 In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
-In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is large, i.e., if the fitness cost is substantial on this time scale.
-Outside this parameter range, there is still a dependence on $s$, but it is weaker.
+In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is small, i.e., if the fitness cost does not affect variant frequency strongly.
+In these cases, there is still a dependence on $s$, but it is weaker.
 

### 2023-01-28 paper/paper.tex
modify section on DCA papers

@@
 For purposes such as identifying functionally constrained drug targets or understanding the function of viral proteins, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
-Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are inherently limited by the fact that mutational effects shift as proteins diverge~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep}.
-% \jdbcomment{Richard, I could not quite figure out how to fit in \citep{strelkowa2021clonal}, let me know if you have ideas.}
+Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are inherently limited by the sparse sampling of related coronaviruses.
 

### 2023-01-29 paper/paper.tex
describe per-phenotype DMS correlation

@@
 Knowledge of the fitness effects of mutations to SARS-CoV-2 can inform assessment of new variants, design of therapeutics resistant to escape, and understanding of the functions of viral proteins.
-However, experimentally determining effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two SARS-CoV-2 proteins.
-Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate the effects of mutations.
+However, experimentally measuring effects of mutations is challenging: we lack tractable lab assays for many SARS-CoV-2 proteins, and comprehensive deep mutational scanning has been applied to only two SARS-CoV-2 proteins.
+Here we develop an approach that leverages millions of publicly available SARS-CoV-2 sequences to estimate effects of mutations.
 We first calculate how many independent occurrences of each mutation are expected to be observed along the SARS-CoV-2 phylogeny in the absence of selection.
@@
 
-Unfortunately, experimentally determining the effects of mutations remains challenging for most SARS-CoV-2 proteins.
+Unfortunately, experimentally measuring the effects of mutations remains challenging for most SARS-CoV-2 proteins.
 For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape,harvey2021sars}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
@@
 For purposes such as identifying functionally constrained drug targets or understanding the function of viral proteins, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
-Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are inherently limited by the sparse sampling of related coronaviruses.
+Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are limited by the sparse sampling of SARS-related coronaviruses.
 
@@
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that enable subsetting on just mutations shared across all datasets and other interactive options such as mousing over points to see mutation identities.
+The experiments in \cite{starr2022deep, flynn2022} measure multiple phenotypes and these plots show the effect of each mutation averaged across these phenotypes; see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html} for plots that show each phenotype separately.
 \label{fig:dms}
@@
 
+\subsection{Comparison to deep mutational scanning}
+Deep mutational scanning data were taken from published studies~\citep{dadonaite2022pseudovirus,starr2022deep,flynn2022,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
+For the spike deep mutational scanning~\citep{dadonaite2022pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
+The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
+In both cases, Figure~\ref{fig:dms} shows the effect averaged across all phenotypes measured by the study.
+For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.
+
 \subsection{Derivation of relationship between actual to expected count ratio and viral fitness}
@@
 \section{Acknowledgments}
-We thank Angie Hinrichs for assistance with questions related to use of the UShER package and its pre-built mutation-annotated tree \jdbcomment{check with her to make sure acknowledging is OK}.
+We thank Angie Hinrichs for assistance with questions related to use of the UShER package and its pre-built mutation-annotated tree.
 The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
-\jdbcomment{Jesse needs to confirm permission of Chodera and Hensley to cite U19 and CEIRR contract.}
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2023-01-29 paper/paper.tex
some final fine-tuning and a bit of shortening of redundant text

@@
 
-\lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the repeated emergence of viral variants with enhanced transmissibility, escape from anti-viral therapeutics, or reduced recognition by immunity~\citep{harvey2021sars,abdool2021new}.
-To anticipate and mitigate this evolution, the scientific community has launched major efforts to assess the risk of new viral variants~\citep{degrace2022defining} and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{moghadasi2022,iketani2022multiple,hiscox2021shutting}.
-Both efforts require determining how specific mutations affect the fitness of the virus.
+\lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the emergence of viral variants with enhanced transmissibility, escape from therapeutics, or reduced recognition by immunity~\citep{harvey2021sars,abdool2021new}.
+To anticipate and mitigate this evolution, the scientific community has launched efforts to assess the risk of new viral variants~\citep{degrace2022defining} and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{moghadasi2022,iketani2022multiple,hiscox2021shutting}.
+Both efforts require determining how specific mutations affect viral fitness.
 
-Unfortunately, experimentally measuring the effects of mutations remains challenging for most SARS-CoV-2 proteins.
-For spike, tractable laboratory assays have identified key functional and antigenic mutations~\citep{weisblum2020escape,harvey2021sars}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
-These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody,tzou2022coronavirus} and informing development of antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
-But most other SARS-CoV-2 proteins lack comparably tractable laboratory assays, despite clearly contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being the targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
-The only non-spike SARS-CoV-2 protein with large-scale deep mutational scanning data is Mpro~\citep{flynn2022,iketani2022functional}.
+Unfortunately, experimentally measuring the effects of mutations is challenging for most SARS-CoV-2 proteins.
+For spike, tractable lab assays have identified key functional and antigenic mutations~\citep{harvey2021sars,weisblum2020escape}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
+These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody,tzou2022coronavirus} and designing antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
+But most SARS-CoV-2 proteins lack tractable lab assays, despite contributing to viral fitness~\citep{thorne2022evolution,syed2021rapid,mcgrath2022sars} and being targets of efforts to develop anti-viral drugs~\citep{tao2021sars}.
+The only non-spike SARS-CoV-2 protein with large-scale experimental measurements of mutation effects is Mpro~\citep{flynn2022,iketani2022functional}.
 
-An alternative to experiments is to estimate the effects of mutations by analyzing natural viral sequences.
-The amount of data available for such analyses has increased dramatically over the last few years with the full-genome sequencing of SARS-CoV-2 from millions of human infections.
+An alternative to experiments is to estimate effects of mutations by analyzing natural viral sequences.
+The amount of data available for such analyses has increased dramatically over the last few years with the sequencing of SARS-CoV-2 from millions of human infections.
 So far analyses of these sequences have focused on analyzing expanding viral clades to identify mutations that mediate immune escape or increase transmissibility~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
@@
 However, only a small minority of all possible mutations are beneficial, with most being nearly neutral or deleterious.
-For purposes such as identifying functionally constrained drug targets or understanding the function of viral proteins, it is important to accurately estimate the effects of neutral or deleterious mutations as well as beneficial ones.
-Other studies have analyzed broader alignments that include coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are limited by the sparse sampling of SARS-related coronaviruses.
+For purposes such as identifying constrained drug targets or understanding the function of viral proteins, it is important to estimate the effects of neutral or deleterious mutations as well as beneficial ones.
+Other studies have analyzed broader alignments of coronaviruses substantially diverged from SARS-CoV-2~\citep{rodriguez2022epistatic, thadani2022learning}, but the resulting estimates are limited by sparse sampling and possible changes in the impacts of some mutations across divergent viruses.
 
-We previously estimated fitness costs of mutations in the HIV-1 genomes from deep sequencing data of within host diversity, where the frequency of a deleterious mutation in the viral quasi-species is determined by a balance between repeated mutation and purifying selection \citep{zanini2017vivo}.
-But in SARS-CoV-2, mutation frequency is a poor predictor of its fitness effect since moderately deleterious mutations can hitch-hike on emerging variants.
 Here we develop a new approach that uses natural sequences to estimate the effects of mutations.
 Our basic insight is that there are now so many SARS-CoV-2 sequences that all non-deleterious single-nucleotide mutations are expected to independently occur many times along the observed phylogenetic tree.
-We therefore first calculate the number of expected observations of independent occurrences of each mutation based on the neutral mutation spectrum of SARS-CoV-2 estimated from four-fold degenerate synonymous sites~\citep{bloom2022evolution}.
+We therefore first calculate the number of expected observations of independent occurrences of each mutation based on the neutral mutation rate of SARS-CoV-2.
 We then compare these expected observations to the actual observations in the SARS-CoV-2 tree to estimate the effect of each mutation.
 The resulting estimates correlate well with existing deep mutational scanning data.
-Most viral proteins have regions that are under strong selective constraints.
+Most viral proteins have regions under strong selective constraints.
 However the accessory proteins mostly show only weak selection against amino-acid and even stop-codon mutations.
@@
 
-\subsection{Estimating fitness from actual versus expected mutation counts}
+\subsection{Mutation effects from actual versus expected counts}
 
@@
 The ratio of actual to expected counts for each mutation is related to its effect on viral fitness.
-The intuition is straightforward: mutations arise at all sites during viral replication, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
+The intuition is straightforward: mutations arise at all sites, but viruses with deleterious mutations are less likely to transmit and be observed in sequencing of human SARS-CoV-2.
 Therefore, the ratio of actual to expected counts will be one for neutral mutations, and less than one for deleterious mutations.
 In the Methods and Appendix, we show that under plausible assumptions about SARS-CoV-2 evolution and sampling intensity (fraction of viruses sequenced), the fitness cost of a deleterious mutation scales roughly inversely with the ratio of actual to expected counts for mutations with costs greater than a few percent.
-A key result from this derivation is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced even very deleterious mutations would have a high chance of being sampled and we would need to study the subsequent spread of the mutations to assess their fitness.
-But since the actual sampling intensity is $\sim$0.1--1\% the number of times mutation is observed sensitively probes more subtle effects that transmit with reduced efficiency.
+A key result is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced even deleterious mutations would have a high chance of being sampled and we would need to study the subsequent spread of the mutations to assess their fitness.
+But since the actual sampling intensity is $\sim$0.1--1\% the number of times a mutation is observed reflects more subtle reductions in transmission efficiency.
 We quantify the effect of each mutation as the logarithm of the ratio of actual to expected counts after summing counts for all nucleotides that encode the relevant amino-acid.
@@
 Despite evidence for some shifts in mutation effects in spike, for the rest of this paper we aggregate counts across viral clades to make a single estimate for each amino-acid mutation.
-The reason is that the accuracy of the estimates increases with the number of counts for each mutation (Figure~\ref{fig:corr}C), and several mutation types only have enough counts for reasonable estimates when aggregating across viral clades (Figure~\ref{fig:expected_vs_actual}B).
-For the purposes of this paper, we deemed it preferable to have more accurate and comprehensive pan-SARS-CoV-2 mutation effect estimates than noisier clade-specific estimates for fewer mutations.
+The reason is that the accuracy of the estimates increases with the number of counts (Figure~\ref{fig:corr}C), and several mutation types only have enough counts for reasonable estimates when aggregating across clades (Figure~\ref{fig:expected_vs_actual}B).
+For the purposes of this paper, we deemed it preferable to have more accurate and comprehensive pan-SARS-CoV-2 estimates than noisier clade-specific estimates for fewer mutations.
 However, the interactive version of Figure~\ref{fig:corr}A linked in the legend enables exploration of mutations with disparate estimates among clades.
@@
 
-We have described a new approach that estimates effects of mutations to SARS-CoV-2 by leveraging the fact that every single-nucleotide change has independently occurred many times along the evolutionary history of sequenced viruses.
-In essence, we treat natural evolution as a deep mutational scan, with the millions of publicly available SARS-CoV-2 sequences from around the world providing a readout of this experiment.
-The key is simply to calculate how many times each mutation has been ``tested'' along the history of sampled viral sequences, and compare that expectation to the actual counts of the mutation among viruses sufficiently fit to have been sequenced in actual human infections.
+Enough SARS-CoV-2 viruses have now been sequenced that many independent occurrences of every tolerated single-nucleotide mutation have been observed along the viral phylogeny.
+Here we have described a new approach that leverages this fact to estimate the effects of these mutations.
+In essence, we treat natural evolution as a deep mutational scan, with the millions of publicly available SARS-CoV-2 sequences providing a readout of this experiment.
+The key is simply to calculate how many times each mutation has been ``tested'' along the history of sampled viral sequences, and compare that expectation to the actual observations of the mutation among viruses sufficiently fit to have been sequenced in actual human infections.
 
-The resulting estimates of mutational effects are robust to subsetting on specific viral clades or geographies, and correlate well with prior experiments.
-In broad strokes, the mutation effects illuminate patterns of constraint: for instance, while there is strong selection on structural and non-structural proteins, there is only limited purifying selection on the accessory proteins.
+The resulting estimates of mutational effects are robust to subsetting on specific viral clades or geographies, and correlate well with experimental measurements.
+In broad strokes, the mutation effects illuminate patterns of constraint: for instance, there is strong selection on structural and non-structural proteins, but only limited purifying selection on the accessory proteins.
 
-However, the real value of our approach is in the detailed maps of effects of specific mutations to all viral proteins, including proteins with important but poorly understood functions that are not easily characterized in the lab.
-These maps will be of value for designing drugs that target constrained sites, interpreting the consequences of mutations observed during viral surveillance, and guiding experiments to better understand the functions of viral proteins.
+However, the real value of our approach is in the detailed maps of effects of specific mutations to all viral proteins, including proteins with poorly understood functions not easily characterized in the lab.
+These maps will be of value for designing drugs that target constrained sites, interpreting the consequences of mutations observed during viral surveillance, and guiding experiments to mechanistically characterize protein function.
 
 There are several caveats to our approach.
-First, because the number of occurrences of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach is sensitive to sequencing (or base-calling) errors as low as ~0.001\% per site.
-Here we have attempted to remove spurious mutations, but further quality control is an important area for future work.
+First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control of publicly available sequences to remove those affected by sequencing errors.
 Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
 Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}.
-Third, the exact relationship between the statistics we calculate and viral fitness depend on both the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
-Although we derive this relationship in the Methods, we do not adjust for sampling intensity and population dynamics across time and space when estimating mutation effects.
-Fourth, we make a single estimate for each mutation across all SARS-CoV-2, thereby neglecting the epistasis that has affected some mutations in at least the spike protein~\citep{starr2022shifting,moulana2022compensatory}.
+Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
+Although we derive this relationship, we do not adjust for sampling intensity and population dynamics when estimating mutation effects.
+Fourth, we make a single estimate for each mutation across all SARS-CoV-2, neglecting the epistasis that can affect some mutations~\citep{starr2022shifting,moulana2022compensatory}.
 Finally there are a few technical caveats to how we count mutations that are discussed in the Methods section.
 
-Conceptually, our approach differs from prior methods that aim to identify beneficial mutations associated with viral clades that expand in relative frequency~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
-Those methods draw information primarily from what happens downstream from a mutation.
-In contrast, our approach treats all mutations equivalently regardless of whether they are on a tip node or at the base of a large clade.
-Our approach is probably better for estimating the effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
-In particular, clade size carries information beyond that contained in mutation counts alone (see Figure~\ref{fig:terminal}).
-Hopefully future work can combine the strengths of our approach and clade-growth methods for even better estimates of mutation effects.
+Conceptually, our approach differs from prior methods that aim to identify beneficial SARS-CoV-2 mutations associated with viral clades that increase in frequency~\citep{obermeyer2022analysis,lee2022inferring,maher2022predicting}.
+Those methods draw information primarily from what happens downstream of a mutation.
+In contrast, we treat all mutations equivalently regardless of whether they are on a tip node or at the base of a large clade.
+Our approach is better for estimating effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
+In particular, clade size carries information beyond that contained in mutation counts alone (Figure~\ref{fig:terminal}).
+Hopefully future work can combine mutation-counting and clade-growth methods for even better estimates of SARS-CoV-2 mutation effects.
+Note our approach is conceptually similar to estimating fitness costs of HIV or polio mutations from mutation-selection balance in deep sequencing of intra-population viral quasispecies~\citep{zanini2017vivo, acevedo2014mutational}, except we analyze mutation occurrences rather than frequencies to account for the phylogenetic structure and genetic hitchhiking that characterize global SARS-CoV-2 evolution.
 
-The power of the approach we have described will increase with growth in viral sequencing.
+The power of the approach we have described will increase with more viral sequencing.
 SARS-CoV-2 is the first virus with enough sequences that every tolerated mutation is observed multiple independent times.
 As costs drop, it is easy to imagine a future with even more viral sequences.
-As this occurs, viral genomic sequencing---which has traditionally been primarily a tool to track evolution and spread---will also become a means to answer questions about the effects of mutations that have traditionally required laboratory experiments.
+As this occurs, viral genomic sequencing---which has traditionally been used primarily to track evolution and spread---will also become an increasingly precise tool to determine the effects of specific mutations.
 
@@
 See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
-The specific version of the repository used for this paper is tagged at \jdbcomment{add link to GitHub tag on final commit}.
+The specific version of the repository used for this paper is tagged as ``bioRxiv-v1'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v1})
 The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.
@@
 \centering
-\includegraphics[width=0.75\linewidth]{figs/terminal.png}
-\caption{
-Relationship between fitness effects of mutations and two measures of the number of descendants.
-At top is shown the log ratio of counts of the mutation on non-terminal (internal) to terminal (tip) branches; larger values indicate mutations more likely to be found in viruses that leave descendants.
-At bottom is shown the mean log number of tip descendants that share all the mutations on each branch containing the mutation of interest; larger values again indicate mutations more likely to be found in viruses that leave more descendants.
-Each point is an amino-acid mutation, the orange line is a least-squares regression, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
-This plot shows only mutations with at least 10 expected counts and 5 actual counts.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
-The number of descendants is calculated using the ``leaves\_sharing\_mutations'' variable of the UShER mutation-annotated tree.
-\label{fig:terminal}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
 \includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=0.75\linewidth]{figs/terminal.png}
+\caption{
+Relationship between fitness effects of mutations and two measures of the number of descendants.
+At top is shown the log ratio of counts of the mutation on non-terminal (internal) to terminal (tip) branches; larger values indicate mutations more likely to be found in viruses that leave descendants.
+At bottom is shown the mean log number of tip descendants that share all the mutations on each branch containing the mutation of interest; larger values again indicate mutations more likely to be found in viruses that leave more descendants.
+Each point is an amino-acid mutation, the orange line is a least-squares regression, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
+This plot shows only mutations with at least 10 expected counts and 5 actual counts.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
+The number of descendants is calculated using the ``leaves\_sharing\_mutations'' variable of the UShER mutation-annotated tree.
+\label{fig:terminal}
+}
+\end{figure*}
+
 \end{document}

### 2023-01-31 paper/paper.tex

@@
 However, some viral accessory proteins are under little to no selection.
-To enable our estimates to be easily used, we provide interactive visualizations (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
+We provide interactive visualizations of effects of mutations to all SARS-CoV-2 proteins (\url{https://jbloomlab.github.io/SARS2-mut-fitness/}).
 The framework we describe is applicable to any virus for which the number of available sequences is sufficiently large that many independent occurrences of each neutral mutation are observed.
@@
 \section{Acknowledgments}
-We thank Angie Hinrichs for assistance with questions related to use of the UShER package and its pre-built mutation-annotated tree.
+We thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built mutation-annotated tree.
 The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.

### 2023-05-18 paper/paper.tex
better plot of synonymous constraint (#29)
* add orf10 to reference.gb

* clean-up script for non-coding constraint

* add axis label

* add rule to snakefile

* add supp figure to paper

* add flag to include/exclude the additional conserved regions in E and M

@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=0.75\linewidth]{figs/synonymous_figure.pdf}
+\caption{
+{\bf Non-coding constraint is concentrated in a few isolated regions in SARS-CoV-2 genome.}
+Panels A and B show the average fitness cost of mutations that do not change amino acids in ORF1ab, S, E, M, N, and ORF3a, non-coding regions between ORFs, and all positions in ORF6, ORF7a, ORF7b, ORF8, ORF10 (as these ORFs that tolerate stop codons) with an expected count exceeding 20.
+Panel A shows the entire genome, panel B zooms into the 3' end of the genome.
+Fitness effects are averaged in sliding windows of size 25 and 50.
+With the exception of a small number of well defined regions, in most of the genome these average fluctuate around 0.
+Regions for which we infer large fitness costs correspond to the ribosomal slippage site at the end of ORF1a, the transcription regulatory sites of S, ORF3a, E, M, and N, as well as two regions in the center of E and the end of M (see red bars in the top part of each panel) \citep{markov2023evolution}.
+Notably, TRS sites of ORF6, ORF7a and ORF8 don't show strong signal of conservation, consistent with their tolerance of stop codons.
+Panels C and D show the distribution of inferred fitness cost of four-fold degenerate sites at the a priori defined slippage and TRS sites.
+In total, 25 positions in these regions are four-fold degenerate compared to 4844 outside these regions.
+Panel C shows that a large fraction of mutations in these a priori known regions are deleterious, while the distribution at the remaining sites is centered around 0 between -2 and +1.
+Furthermore, strongly deleterious mutations with fitness estimate below -3 predominantly fall in these 25 sites.
+If the two additional conserved regions in E and M are included, all but one mutation outside these regions has a fitness cost estimate above -3.
+\label{fig:synonymous}
+}
+\end{figure*}
+
 \end{document}

### 2023-05-18 scripts/known_conserved_noncoding.py

@@ -2,7 +2,7 @@ from Bio import SeqIO
 import numpy as np
 import matplotlib.pyplot as plt
 
-def get_conserved_regions():
+def get_conserved_regions(only_known=False):
     TRS_motif = "ACGAAC"
     margin = 0
     ref = SeqIO.read('data/reference.gb', 'genbank')
@@ -16,22 +16,27 @@ def get_conserved_regions():
             pos += 1
 
 
-    conserved_regions = []
-    for mpos in motifs:
-        conserved_regions.append((mpos-margin,mpos+len(TRS_motif)+margin))
+    conserved_regions = {}
+    for mi, mpos in enumerate(motifs):
+        conserved_regions[f'TRS {mi+1}'] = (mpos-margin,mpos+len(TRS_motif)+margin)
 
     ## attenuator hairpin
-    conserved_regions.append((ref.seq.find("ATGCTTCA"), ref.seq.find("CGTTTTT")))
+    conserved_regions['hairpin']=(ref.seq.find("ATGCTTCA"), ref.seq.find("CGTTTTT"))
+    ## attenuator hairpin
+    conserved_regions['slippery seq']=(ref.seq.find("TTTAAACG"), ref.seq.find("TTTAAACG")+7)
     ## 3 stem pseudoknot
-    conserved_regions.append((ref.seq.find("GCGGTGT"), ref.seq.find("TTTTGA", 13474)))
-    ## center of E
-    conserved_regions.append((26330,26360))
+    conserved_regions['3-stem-pseudoknot']=(ref.seq.find("GCGGTGT"), ref.seq.find("TTTTGA", 13474))
+    if not only_known:
+        ## center of E
+        conserved_regions['E-center']=(26330,26360)
+        ## end of M
+        conserved_regions['M-end']=(27170,27200)
 
     conserved_vector = np.zeros(len(ref.seq), dtype=bool)
-    for r in conserved_regions:
+    for r in conserved_regions.values():
         conserved_vector[r[0]:r[1]] = True
 
-    return conserved_vector
+    return conserved_vector, conserved_regions
 
 
 if __name__=="__main__":
@@ -40,8 +45,8 @@ if __name__=="__main__":
     fname = 'results/nt_fitness/ntmut_fitness_all.csv'
     min_expected_count = 20
     fitness = pd.read_csv(fname, sep=',')
-    known_conserved_regions = get_conserved_regions()
[…]

### 2023-05-18 scripts/noncoding_constraints.py

@@ -1,8 +1,10 @@
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
+from matplotlib.gridspec import GridSpec
 import argparse
 from scipy.interpolate import interp1d
+from known_conserved_noncoding import get_conserved_regions
 
 def convolve_by_valid_entry(position, values, ws, mode='valid'):
     window = np.ones(ws)/ws
@@ -10,11 +12,13 @@ def convolve_by_valid_entry(position, values, ws, mode='valid'):
 
 def convolve_by_position(position, values, ind, ws, mode='valid'):
     pos = position
+    tmp_values = np.copy(values)
+    tmp_values[~ind]=0
     window = np.ones(ws)
     return np.convolve(pos*ind, window, mode)/np.convolve(ind, window, mode), \
-           np.convolve(values*ind, window, mode)/np.convolve(ind, window, mode)
+           np.convolve(tmp_values*ind, window, mode)/np.convolve(ind, window, mode)
 
-def add_genes(ax, vmin, vmax, both=False):
+def add_genes(ax, vmin, vmax, both=False, additional_regions = None, xscale=30000, region=None):
     import matplotlib.patches as patches
     from Bio import SeqIO
     ref = SeqIO.read('data/reference.gb', 'genbank')
@@ -26,11 +30,50 @@ def add_genes(ax, vmin, vmax, both=False):
             f_stop = f.location.end.position
             rect = patches.Rectangle((f_start, vmin+offset), f_stop - f_start, vmax-vmin, linewidth=1, edgecolor=f'C{fi}', facecolor=f'C{fi}')
             ax.add_patch(rect)
-            ax.text(f_start, vmin + offset, f.qualifiers['gene'][0])
-            if both==True:
-                ax.text(f_stop, vmin + offset, f.qualifiers['gene'][0], ha='right')
+            if region is None or (f_start>=region[0] and f_start<region[1]):
+                ax.text(f_start, vmin + offset, f.qualifiers['gene'][0])
+                if both==True:
+                    ax.text(f_stop, vmin + offset, f.qualifiers['gene'][0], ha='right')
+            fi += 1
+    if additional_regions is not None:
+        for label, (f_start, f_stop) in sorted(additional_regions.items(), key=lambda x:x[1][0]):
[…]

### 2023-05-18 scripts/noncoding_constraints.py
integration synonymous plot script into pipeline better
add `min_expected_count` as parameter to synonymous constraint script

integrated into pipeline with a per mat output

@@ -79,6 +79,7 @@ if __name__=="__main__":
     parser = argparse.ArgumentParser()
     parser.add_argument("--fitness", type=str, help="fitness file")
     parser.add_argument("--output", type=str, help="output file")
+    parser.add_argument("--min_expected_count", type=int, help="min expected count")
     args = parser.parse_args()
 
     # read in tabular fitness file
@@ -97,7 +98,7 @@ if __name__=="__main__":
 
     fitness_noORFs = erase_genes(fitness, ['ORF6', 'ORF7a', 'ORF7b','ORF8', 'ORF9b', 'ORF10'])
 
-    expected_count_cutoff = 20
+    expected_count_cutoff = args.min_expected_count
     fitness_measure = "delta_fitness"
     fs = 16
 
@@ -190,5 +191,5 @@ if __name__=="__main__":
     ax4.legend()
     ax4.text(-0.1,0.9, 'D', fontsize=fs*1.5, transform=ax4.transAxes)
 
-    plt.savefig(args.output)
+    plt.savefig(args.output, metadata={"creationDate": None})
 

### 2023-05-31 scripts/compare_dca_fitness.py
add comparison to other studies estimating mutational effects (#34)
* get comparator data

* parse comparator CSVs to data frames

* merge comparator, DMS, and fitness into one data frame

* make `analyze_comparator_studies` plot

* add comparator plot to docs

* rerun pipeline

@@ -1,46 +1,0 @@
-import pandas as pd
-import numpy as np
-from collections import defaultdict
-import matplotlib.pyplot as plt
-
-remap_genes = {"Spike":"S", "ORF1a":"ORF1ab", "ORF1b":"ORF1ab",
-                "Nucleocapsid":"N", "Membrane":"M", "Envelope":"E"}
-
-def load_dca():
-    dca = pd.read_csv(snakemake.input.dca)
-    dca_mutability = {}
-    for i, row in dca.iterrows():
-        dca_mutability[(remap_genes.get(row["protein"], row["protein"]), row["position_protein"])] = row["mutability_score(DCA)"]
-    return dca_mutability
-
-
-
-def load_fitness():
-    return pd.read_csv(snakemake.input.fitness)
-
-
-if __name__ == "__main__":
-    dca_mutability = load_dca()
-    fitness = load_fitness()
-    fitness = fitness.loc[fitness.expected_count>30]
-
-    average_fitness = fitness.groupby(["gene", "aa_site"])['delta_fitness'].mean()
-
-    linked_dca_fitness = defaultdict(dict)
-    for i, row in average_fitness.iteritems():
-        gene, aa_site = i
-        if (gene, aa_site) in dca_mutability and not pd.isna(dca_mutability[(gene, aa_site)]):
-            linked_dca_fitness[gene][aa_site] = (dca_mutability.get((gene, aa_site), np.nan), row)
-
-    fig, axes = plt.subplots(2, 5, figsize=(12, 5))
-    for i, gene in enumerate(linked_dca_fitness):
-        ax = axes.ravel()[i]
-        d = np.array(list(linked_dca_fitness[gene].values()))
-        corr = np.corrcoef(d.T)[0,1]
-        ax.scatter(d[:,0], d[:,1], s=10, alpha=0.3)
-        ax.set_title(f"{gene} (r = {corr:.2f})")
-        ax.set_xlabel('DCA mutability')
-        ax.set_ylabel('mean fitness effect')
-    fig.tight_layout()
-    
-    fig.savefig(snakemake.output.plot)

### 2023-06-01 paper/paper.tex
comment out supp figure so LaTex compiles
update all existing figures (new figs and text revisions not yet added)

@@
 
-To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} of $\sim$6.5-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} of $\sim$7-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
 Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
@@
 
-The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$500 for \texttt{C$\rightarrow$T} to only $\sim$8 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
+The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$565 for \texttt{C$\rightarrow$T} to only $\sim$9 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
 This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation,neher2022contributions}.
@@
 The experiments in \cite{starr2022deep, flynn2022} measure multiple phenotypes and these plots show the effect of each mutation averaged across these phenotypes; see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html} for plots that show each phenotype separately.
+{\color{red} Update citations and min counts}
 \label{fig:dms}
@@
 \subsection{Counting mutations along the phylogenetic tree}
-We counted occurrences of each mutation in each viral clade using the UShER pre-built mutation-annotated tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} from Dec-18-2022 (\url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2/2022/12/18/public-2022-12-18.all.masked.nextclade.pangolin.pb.gz}), which contains all $\sim$6.5-million SARS-CoV-2 sequences that are available in public databases.
+We counted occurrences of each mutation in each viral clade using the UShER pre-built mutation-annotated tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} from May-11-2023 (\url{http://hgdownload.soe.ucsc.edu/goldenPath/wuhCor1/UShER_SARS-CoV-2/2023/05/11/public-2023-05-11.all.masked.nextclade.pangolin.pb.gz}), which contains all $\sim$7-million SARS-CoV-2 sequences that are available in public databases.
 To make these counts at a per-clade level, we first subsetted the mutation-annotated tree on all sequences for each Nexstrain clade~\citep{aksamentov2021nextclade}, retained only clades with at least $10^4$ sequences, and then used the matUtils program distributed with UShER to extract the nucleotide mutations on every branch of the each clade-subsetted mutation-annotated tree.
@@
 There is strong selection against stop codons (negative effects) for all but the C-terminus of ORF3a, but only a few positions show strong selection against amino-acid substitutions.
-This plot shows only mutations with 20 expected counts.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/ORF3a.html} for an interactive version of this plot along with zoomable heatmap of the effects of specific amino-acid substitutions.
@@
 \centering
-\includegraphics[width=0.75\linewidth]{figs/synonymous_figure.pdf}
+%\includegraphics[width=0.75\linewidth]{figs/synonymous_figure.pdf}
 \caption{

### 2023-06-02 paper/paper.tex
add new figures (not linked to text / legends yet)
revise first two results sections

@@
 \lettrine[lines=2]{\color{color2}T}{}he rapid evolution of SARS-CoV-2 has led to the emergence of viral variants with enhanced transmissibility, escape from therapeutics, or reduced recognition by immunity~\citep{harvey2021sars,abdool2021new}.
-To anticipate and mitigate this evolution, the scientific community has launched efforts to assess the risk of new viral variants~\citep{degrace2022defining} and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{moghadasi2022,iketani2022multiple,hiscox2021shutting}.
+To anticipate and mitigate this evolution, the scientific community has launched efforts to assess the risk of new viral variants~\citep{degrace2022defining} and create therapeutics that target constrained regions of the virus where resistance is less likely to evolve~\citep{moghadasi2023,iketani2022multiple,hiscox2021shutting}.
 Both efforts require determining how specific mutations affect viral fitness.
@@
 Unfortunately, experimentally measuring the effects of mutations is challenging for most SARS-CoV-2 proteins.
-For spike, tractable lab assays have identified key functional and antigenic mutations~\citep{harvey2021sars,weisblum2020escape}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2022pseudovirus,greaney2021complete,cao2022imprinted}.
+For spike, tractable lab assays have identified key functional and antigenic mutations~\citep{harvey2021sars,weisblum2020escape}, and enabled deep mutational scanning measurements of how most mutations affect receptor binding, cellular infection, and antibody recognition~\citep{starr2020deep,dadonaite2023pseudovirus,greaney2021complete,cao2022imprinted}.
 These experimental data are valuable for assessing new spike variants~\citep{degrace2022defining,greaney2022antibody,tzou2022coronavirus} and designing antibody therapeutics with greater resistance to escape~\citep{starr2021sars,rappazzo2021broad,cao2022rational}.
@@
 Most viral proteins have regions under strong selective constraints.
-However the accessory proteins mostly show only weak selection against amino-acid and even stop-codon mutations.
-Overall, our work demonstrates a new approach to determine the effects of mutations, and provides detailed maps of functional constraint across the SARS-CoV-2 proteome.
+However some accessory proteins show only weak selection against amino-acid and even stop-codon mutations.
+Overall, our work demonstrates a new approach to determine the effects of mutations, and provides detailed maps of mutational effects across the SARS-CoV-2 proteome.
 
@@
 
-To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} of $\sim$7-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2022evolution}.
-Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see Discussion for caveats about nucleotide-level selection).
+To determine how many times each mutation is expected to be observed, we used the pre-built UShER tree~\citep{mcbroome2021daily,turakhia2021ultrafast,lanfear2020} of $\sim$7-million public SARS-CoV-2 sequences to count nucleotide mutations at four-fold degenerate sites~\citep[Figure~\ref{fig:expected_vs_actual}A;][]{bloom2023evolution}.
+Because mutations at such sites never alter the amino-acid sequence, these counts reflect the mutation process in the absence of protein-level selection (see below for caveats about nucleotide-level selection).
 The expected counts of a mutation from nucleotide $x$ to $y$ is simply the average count of this type of mutation across all four-fold degenerate sites with parental identity $x$.
 Importantly, we count independent \emph{occurrences} of each mutation along the branches of the tree, not the sequences with the mutation in the final alignment (Figure~\ref{fig:expected_vs_actual}A.
-We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\citep{bloom2022evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
+We also compute expected counts separately for each SARS-CoV-2 clade to account for shifts in mutation spectrum~\citep{bloom2023evolution,ruis2022mutational}, and apply quality-control steps to remove spurious mutations (see Methods).
 
-The expected counts per mutation (summed across all viral clades) vary with mutation type, ranging from $\sim$565 for \texttt{C$\rightarrow$T} to only $\sim$9 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
-This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2022evolution,ruis2022mutational,de2021mutation,neher2022contributions}.
+The expected counts per mutation vary with mutation type, ranging from $\sim$565 for \texttt{C$\rightarrow$T} to only $\sim$9 for \texttt{T$\rightarrow$G} mutations (Figure~\ref{fig:expected_vs_actual}B).
+This variation is because the SARS-CoV-2 mutation spectrum is highly biased towards specific mutation types~\citep{bloom2023evolution,ruis2022mutational,de2021mutation,neher2022contributions}.
+However, because there are so many SARS-CoV-2 sequences we are able to estimate the rate of even the rarest mutation types with high accuracy~\cite{bloom2023evolution}.
+For instance, there are $\sim 1.9\times 10^4$ observed occurrences of \texttt{T$\rightarrow$G} mutations across all $\sim$2,100 four-fold degenerate sites with a parental identity of \texttt{T}, which is enough to estimate the \texttt{T$\rightarrow$G} mutation rate (and therefore the expected counts of each mutation) with high accuracy.
 
@@
 A key result is the dependence on sampling intensity: if all human SARS-CoV-2 were sequenced even deleterious mutations would have a high chance of being sampled and we would need to study the subsequent spread of the mutations to assess their fitness.
-But since the actual sampling intensity is $\sim$0.1--1\% the number of times a mutation is observed reflects more subtle reductions in transmission efficiency.
+But the actual sampling intensity is $\sim$0.1\%, since there are $\sim$7-million publicly available SARS-CoV-2 sequences and the total number of human infections is now probably roughly on par with the total global population of $\sim$8-billion people.
+At this sampling intensity, the number of times a mutation is observed reflects more subtle reductions in transmission efficiency.
 We quantify the effect of each mutation as the logarithm of the ratio of actual to expected counts after summing counts for all nucleotides that encode the relevant amino-acid.
@@
 
-\subsection*{Mutation-effect estimates are robust to subsampling natural sequences, with some evidence of epistasis in spike}
+\subsection*{Mutation-effect estimates are robust to subsampling and phylogeny, with some evidence of epistasis in spike}
 
@@
 However, the interactive version of Figure~\ref{fig:corr}A linked in the legend enables exploration of mutations with disparate estimates among clades.
+
+An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations even after aggregating across clades.
+To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID as of March-29-2023.
+There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
+This concordance indicates that the $\sim$7-million public sequence set is large enough that statistical sampling noise is low, and so throughout this paper we use that sequence set due to our preference for publicly available data.
+We also note that the fact that two trees (which have independently inferred topologies) give nearly identical estimates indicates that uncertainty in tree topology has minimal effect on our fitness estimates.
+The insensitivity of our fitness estimates to uncertainty in tree topology is expected: the main uncertainty in densely sampled viral phylogenies involves polytomies of closely related sequences, but our estimates only use the counts of mutations along tree branches, which are unaffected by uncertainty in the resolution of polytomies of closely related sequences.
 
@@
 \caption{
-Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2022pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2022,iketani2022functional}.
+Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2023pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2022,iketani2022functional}.
 Each point is an amino-acid mutation, the orange line is a least-squares regression, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
@@
 We examined how the mutation effects estimated using our approach compare with prior high-throughput deep mutational scanning measurements.
-For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD)~\citep{starr2020deep,starr2022deep} and spike pseudotyped lentiviruses~\citep{dadonaite2022pseudovirus}.
+For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD)~\citep{starr2020deep,starr2022deep} and spike pseudotyped lentiviruses~\citep{dadonaite2023pseudovirus}.
 For Mpro (also known as nsp5 or 3CLpro), two different labs have performed deep mutational scanning using the same basic methodology of assaying protease cleavage in yeast~\citep{flynn2022,iketani2022functional}.
@@
 \subsection{Calculation of expected counts}
-To calculate the expected counts for each nucleotide mutation, we analyzed just the four-fold degenerate sites in each clade in an approach paralleling that of \cite{bloom2022evolution}.
+To calculate the expected counts for each nucleotide mutation, we analyzed just the four-fold degenerate sites in each clade in an approach paralleling that of \cite{bloom2023evolution}.
 Specifically, we identify all non-excluded four-fold degenerate sites in each clade founder.
 We then count nucleotide mutations just at those sites in each clade, and calculate the expected per-site number of mutations from nucleotide $x$ to $y$ as the total number of $x$ to $y$ mutations at four-fold degenerate sites divided by the number of four-fold degenerate sites with $x$ as the parental identity.
-This analysis is done at the clade level for two reasons: referencing mutations to the clade founder (rather than the Wuhan-Hu-1 reference) limits problem with the approach that would arise at sites that substitute multiple times in the history of a sequence (since each clade is a relatively high-identity group multiple mutations at the same site within a clade are very rare), and because it is know that SARS-CoV-2 mutation rates vary somewhat among clades~\citep{bloom2022evolution,ruis2022mutational}.
+This analysis is done at the clade level for two reasons: referencing mutations to the clade founder (rather than the Wuhan-Hu-1 reference) limits problem with the approach that would arise at sites that substitute multiple times in the history of a sequence (since each clade is a relatively high-identity group multiple mutations at the same site within a clade are very rare), and because it is know that SARS-CoV-2 mutation rates vary somewhat among clades~\citep{bloom2023evolution,ruis2022mutational}.
 We only retain clades with at least 5000 mutations at four-fold degenerate sites in order to avoid inaccurate estimates of expected counts due to low sampling of mutations.
@@
 \subsection{Comparison to deep mutational scanning}
-Deep mutational scanning data were taken from published studies~\citep{dadonaite2022pseudovirus,starr2022deep,flynn2022,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
-For the spike deep mutational scanning~\citep{dadonaite2022pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
+Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
+For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
 The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
@@
 \begin{figure*}[h]
+\centering
+\includegraphics[width=0.4\linewidth]{figs/gisaid_corr.png}
+\caption{
+Correlation between amino-acid fitness estimates from the $\sim$7-million sequence set of all publicly available sequences as of May-11-2023, or the set of all $\sim$14-million sequences in GISAID as of March-29-2023.
+The orange text in the upper-left corner of the plot shows the Pearson correlation.
+The high correlation indicates that our estimates are not substantially limited by noise related to statistical sampling of mutation counts, since using another tree with twice as many sequences does not substantially alter the estimates.
+Note that the main text figures all use the smaller $\sim$7-million publicly available sequence set.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/mat_aa_fitness_correlations.html} for an interactive version of this plot that also shows the correlations for other sets of publicly available sequences.
+\label{fig:gisaid_corr}
+}
+\end{figure*}
+
+\begin{figure*}[h]
+\centering
+\includegraphics[width=\linewidth]{figs/dnds.png}
+\caption{
+\jdbcomment{write}
+\url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html}
+\label{fig:dnds}
+}
+\end{figure*}
+
+\begin{figure*}[h]
+\centering
+\includegraphics[width=0.75\linewidth]{figs/comparator_corr.png}
+\caption{
+\jdbcomment{write}
+\url{https://jbloomlab.github.io/SARS2-mut-fitness/comparator_corr.html}
+\label{fig:comparator}
+}
+\end{figure*}
+
+\begin{figure*}[h]
 \includegraphics[width=\linewidth]{figs/ORF3a.pdf}
@@
 \centering
-%\includegraphics[width=0.75\linewidth]{figs/synonymous_figure.pdf}
+\includegraphics[width=\linewidth]{figs/synonymous_constraint_figure.pdf}
 \caption{

### 2023-06-03 paper/paper.tex
add dn/ds and comparator predictions

@@
 An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations even after aggregating across clades.
-To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID as of March-29-2023.
+To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
 There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
-This concordance indicates that the $\sim$7-million public sequence set is large enough that statistical sampling noise is low, and so throughout this paper we use that sequence set due to our preference for publicly available data.
-We also note that the fact that two trees (which have independently inferred topologies) give nearly identical estimates indicates that uncertainty in tree topology has minimal effect on our fitness estimates.
-The insensitivity of our fitness estimates to uncertainty in tree topology is expected: the main uncertainty in densely sampled viral phylogenies involves polytomies of closely related sequences, but our estimates only use the counts of mutations along tree branches, which are unaffected by uncertainty in the resolution of polytomies of closely related sequences.
+This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
+%We also note that the fact that two trees (which have independently inferred topologies) give nearly identical fitness estimates indicates that uncertainty in tree topology has minimal effect.
+% The insensitivity of our fitness estimates to uncertainty in tree topology is expected: the main uncertainty in densely sampled viral phylogenies involves polytomies of closely related sequences, but our estimates only use the counts of mutations along tree branches, which are unaffected by uncertainty in the resolution of polytomies of closely related sequences.
 
@@
 Mutation types are color-coded as in panel A.
+The apparent constraint on synonymous mutation in ORF9b is probably because this gene is encoded in an overlapping reading frame with N~\cite{jungreis2021conflicting}.
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
@@
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
+The handful of synonymous mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
 
-To investigate differences among viral proteins, we computed the distributions of effects separately for each gene (Figure~\ref{fig:dist}B).
+To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B).
 SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
@@
 \caption{
-Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2023pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2022,iketani2022functional}.
+Correlation of mutation-effect estimates with experimental deep mutational scanning measurements for {\bf (A)} the full spike~\citep{dadonaite2023pseudovirus} or its RBD~\citep{starr2022deep}, and {\bf (B)} Mpro~\citep{flynn2023,iketani2022functional}.
 Each point is an amino-acid mutation, the orange line is a least-squares regression, and orange text in the upper left shows the number of mutations and Pearson correlation coefficient.
 Each sub-panel shows a different set of mutations (depending on which mutations were measured in that experiment).
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that enable subsetting on just mutations shared across all datasets and other interactive options such as mousing over points to see mutation identities.
-The experiments in \cite{starr2022deep, flynn2022} measure multiple phenotypes and these plots show the effect of each mutation averaged across these phenotypes; see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html} for plots that show each phenotype separately.
-{\color{red} Update citations and min counts}
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_corr.html} for plots that also show the Mpro dataset from~\citep{flynn2022} and have various interactive options.
+The plots in this figure show the average of the multiple phenotypes measured in the deep mutational scanning of \cite{starr2022deep}; see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} for each phenotype separately.
+This figure only shows mutations with at least 20 expected counts, which is higher than the threshold of 10 used in most of the rest of this paper (this threshold can be adjusted in the interactive plots).
 \label{fig:dms}
@@
 For spike, two distinct experimental methodologies have been used to characterize large numbers of mutations: yeast display of the receptor-binding domain (RBD)~\citep{starr2020deep,starr2022deep} and spike pseudotyped lentiviruses~\citep{dadonaite2023pseudovirus}.
-For Mpro (also known as nsp5 or 3CLpro), two different labs have performed deep mutational scanning using the same basic methodology of assaying protease cleavage in yeast~\citep{flynn2022,iketani2022functional}.
+For Mpro (also known as nsp5 or 3CLpro), two different labs have performed deep mutational scanning using the same basic methodology of assaying protease cleavage in yeast~\citep{flynn2022,flynn2023,iketani2022functional}.
 
-For spike, our estimates from natural sequences correlate with each set of experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A).
-If we increase the minimum expected counts from 10 to 20 and subset only on mutations shared among all three data sets, then the correlations between the estimates and experiments ($r = 0.66$) become even closer to the cross-experiment correlations ($r = 0.72$; see interactive version of Figure~\ref{fig:dms}A linked in legend).
-Some of the mutations with the greatest divergence between our sequence-based estimates and the deep mutational scanning likely represent experimental artifacts.
+For spike, our estimates from natural sequences correlate with the experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A), with Pearson correlations of 0.66 between the fitness estimates and experiments versus 0.72 between the two types of experiments.
+Neither set of experimental measurements fully captures how mutations affect viral fitness, since both RBD yeast display and lentiviral pseudotyping are imperfect proxies for spike function during actual human infections.
+Therefore, it is unclear how much the differences between the fitness estimates and experiments are due to noise in the fitness estimates versus limitations of the experiments.
+However, the fact that the fitness estimates correlate with the experiments almost as well as experimental methodologies correlate with each (Figure~\ref{fig:dms}A) suggests estimates are of comparable quality to experimental measurements.
+At least some of the mutations with the greatest divergence between our fitness estimates and the deep mutational scanning likely represent experimental artifacts.
 For instance, P527L, which is favorable in the RBD deep mutational scan but deleterious in the sequence-based estimates and full-spike scan, is at the C-terminus of the yeast-displayed RBD~\citep{starr2020deep} where it may adopt a non-native conformation.
 
-The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the two experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
-Because both Mpro experiments use a similar yeast-based methodology~\citep{flynn2022,iketani2022functional} it is possible that the higher correlation of the experiments with each other than the sequence-based estimates reflects shared artifacts of the yeast experiments.
-In particular, some Mpro mutations estimated to have deleterious effects in natural sequences are well tolerated in the yeast experiments.
-This difference could arise if the yeast experiments only capture some of the constraints on Mpro in the context of actual virus.
-For instance, a stop codon at Q306 is well tolerated in both deep mutational scans but extremely disfavorable in our sequence-based estimates, and would be highly deleterious to actual virus as it would truncate the polyprotein.
-Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, possibly because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
+The sequence-based fitness estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
+However, the Mpro experiments all use a similar yeast-based methodology~\citep{flynn2022,flynn2023,iketani2022functional} that probably fails to capture significant aspects of Mpro's actual function during human infections.
+For instance, a stop codon at Q306 is well tolerated in the yeast-based Mpro deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would clearly be highly deleterious to actual virus as it would truncate the polyprotein.
+Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, probably because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
+
+\subsection*{Mutation-effect estimates better capture functional constraint than dN/dS ratios or predictions from other methods}
+A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
+These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
+A known limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
+Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
+The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
+The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio. 
+For instance, the ACE2-affinity enhancing spike mutation N501Y arose in several SARS-CoV-2 variants early in the pandemic, and has since remained fixed due its importance for receptor binding~\citep{starr2022deep}.
+Our mutation-effect estimates correctly reflect that site 501 is strongly prefers tyrosine, but the site has a high dN/dS ratio due to the early convergent evolution of this site to that preferred amino acid.
+
+Our mutation-effect estimates also correlate substantially better with deep mutational scanning experiments than predictions from two methods trained to learn mutation effects and epistatic interactions from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
+Our estimates also correlate better with experiments than predictions of a machine-learning method that integrates a wide range of sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
+These results suggest that our relatively straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.
 
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=\linewidth]{figs/synonymous_constraint_figure.pdf}
+\caption{
+{\bf Non-coding constraint is concentrated in a few isolated regions in SARS-CoV-2 genome.}
+{\bf (A), (B)} The average fitness cost of synonymous mutations (do not change amino acids) in ORF1ab, S, E, M, N, and ORF3a, non-coding regions between ORFs, and all positions in ORF6, ORF7a, ORF7b, ORF8, ORF10 (as these ORFs tolerate stop codons, see Figure~\ref{fig:dist}B).
+Panel A shows the entire genome, panel B zooms into the 3' end of the genome.
+Fitness effects are averaged in sliding windows of size 25 and 50.
+With the exception of a small number of well defined regions, in most of the genome these averages fluctuate around zero.
+Regions with large fitness costs for non-coding mutations correspond to the ribosomal slippage site at the end of ORF1a, the transcription regulatory sites of S, ORF3a, E, M, and N, as well as two regions in the center of E and the end of M (see red bars in the top part of each panel) \citep{markov2023evolution}.
+Notably, TRS sites of ORF6, ORF7a and ORF8 don't show strong signal of conservation, consistent with their tolerance of stop codons.
+{\bf (C), (D)} The distribution of inferred fitness cost of four-fold degenerate sites at regions of known non-coding constraint within the protein coding genes (the ribosomal slippage and TRS sites) versus all other four-fold degenerate sites.
+In total, 25 four-fold degenerate sites are in these regions of known constraint, whereas 4844 four-fold degenerate sites are outside these regions.
+Panel C shows that a large fraction of mutations in these known regions of constraint are deleterious, while the distribution for the remaining four-fold degenerate sites is centered around 0 between -2 and +1.
+Furthermore, strongly deleterious mutations with fitness estimate below -3 predominantly fall in the 25 sites of known non-coding constraint.
+If the two additional conserved regions in E and M are included, all but one mutation outside these regions has a fitness cost estimate above -3.
+\label{fig:synonymous}
+}
+\end{figure*}
+
 \begin{figure*}[h]
@@
 \caption{
-\jdbcomment{write}
-\url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html}
+The site-mean of the estimated fitness effects outperforms dN/dS ratios as an indicator of functional constraint as assessed by correlation with deep mutational scanning measurements.
+{\bf (A)}
+Pearson correlation of log dN/dS ratio at each site with the mean estimated fitness of all amino acid mutations at each site from the current study.
+For most genes, the log dN/dS ratio is correlated with the mean fitness effect, but the correlations are only modest (always substantially less than one).
+The exception is the accessory proteins which are mostly under minimal amino-acid level selection (Figure~\ref{fig:dist}B) and only show a weak correlation. 
+{\bf (B)}
+Correlation of the log dN/dS ratio or the mean estimated fitness effect at each site with the mean impact of amino-acid mutations at that site as experimentally measured by deep mutational scanning.
+The numbers in the upper right of each panel give the Pearson correlation.
+In every case, the mean estimated fitness effect (top row) is much better correlated with the experimental measurements than the log dN/dS (bottom row).
+The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version of this plot.
 \label{fig:dnds}
@@
 \caption{
-\jdbcomment{write}
-\url{https://jbloomlab.github.io/SARS2-mut-fitness/comparator_corr.html}
+The mutation effect estimates from the current study are substantially better correlated with deep mutational scanning measurements for spike than predicted mutation effects made using three other approaches.
+Each row shows the correlation of the fitness estimates from the current study and full-spike deep mutational scanning measurements versus values from another study.
+The Maher et al (2022) and Thadani et (2023) studies make predictions for individual amino-acids, so those correlations are with the mutation-level fitness estimates and deep mutational scanning measurements.
+The Rodriguez-Rivas et al (2022) study only reports site-level predictions, so those correlations are with the site mean fitness estimates and deep mutational scanning measurements.
+Each row only shows mutations or sites with a reported value from all three approaches.
+The numbers in the upper right give the Pearson correlation.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/comparator_corr.html} for an interactive version of this plot.
 \label{fig:comparator}
@@
 
-\begin{figure*}
-\centering
-\includegraphics[width=\linewidth]{figs/synonymous_constraint_figure.pdf}
-\caption{
-{\bf Non-coding constraint is concentrated in a few isolated regions in SARS-CoV-2 genome.}
-Panels A and B show the average fitness cost of mutations that do not change amino acids in ORF1ab, S, E, M, N, and ORF3a, non-coding regions between ORFs, and all positions in ORF6, ORF7a, ORF7b, ORF8, ORF10 (as these ORFs that tolerate stop codons) with an expected count exceeding 20.
-Panel A shows the entire genome, panel B zooms into the 3' end of the genome.
-Fitness effects are averaged in sliding windows of size 25 and 50.
-With the exception of a small number of well defined regions, in most of the genome these average fluctuate around 0.
-Regions for which we infer large fitness costs correspond to the ribosomal slippage site at the end of ORF1a, the transcription regulatory sites of S, ORF3a, E, M, and N, as well as two regions in the center of E and the end of M (see red bars in the top part of each panel) \citep{markov2023evolution}.
-Notably, TRS sites of ORF6, ORF7a and ORF8 don't show strong signal of conservation, consistent with their tolerance of stop codons.
-Panels C and D show the distribution of inferred fitness cost of four-fold degenerate sites at the a priori defined slippage and TRS sites.
-In total, 25 positions in these regions are four-fold degenerate compared to 4844 outside these regions.
-Panel C shows that a large fraction of mutations in these a priori known regions are deleterious, while the distribution at the remaining sites is centered around 0 between -2 and +1.
-Furthermore, strongly deleterious mutations with fitness estimate below -3 predominantly fall in these 25 sites.
-If the two additional conserved regions in E and M are included, all but one mutation outside these regions has a fitness cost estimate above -3.
-\label{fig:synonymous}
-}
-\end{figure*}
-
 \end{document}

### 2023-06-03 paper/paper.tex
revision of paper

@@
 
-For spike, our estimates from natural sequences correlate with the experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A), with Pearson correlations of 0.66 between the fitness estimates and experiments versus 0.72 between the two types of experiments.
-Neither set of experimental measurements fully captures how mutations affect viral fitness, since both RBD yeast display and lentiviral pseudotyping are imperfect proxies for spike function during actual human infections.
-Therefore, it is unclear how much the differences between the fitness estimates and experiments are due to noise in the fitness estimates versus limitations of the experiments.
-However, the fact that the fitness estimates correlate with the experiments almost as well as experimental methodologies correlate with each (Figure~\ref{fig:dms}A) suggests estimates are of comparable quality to experimental measurements.
-At least some of the mutations with the greatest divergence between our fitness estimates and the deep mutational scanning likely represent experimental artifacts.
+For spike, our estimates from natural sequences correlate with the experiments almost as well as the two experimental methodologies correlate with each other (Figure~\ref{fig:dms}A), with Pearson correlations of 0.66 between the estimates and experiments versus 0.72 between the two experiments.
+Neither experiment fully captures how mutations affect viral fitness, since both RBD yeast display and lentiviral pseudotyping are imperfect proxies for spike function during actual human infections.
+Therefore, it is unclear how much the differences between the mutation-effect estimates and experiments are due to noise in the estimates versus limitations of the experiments.
+However, the fact that the estimates correlate with the experiments almost as well as the experiments correlate with each (Figure~\ref{fig:dms}A) suggests the estimates are of comparable quality to experimental measurements.
+At least some of the mutations with the greatest divergence between our estimates and the deep mutational scanning likely represent experimental artifacts.
 For instance, P527L, which is favorable in the RBD deep mutational scan but deleterious in the sequence-based estimates and full-spike scan, is at the C-terminus of the yeast-displayed RBD~\citep{starr2020deep} where it may adopt a non-native conformation.
 
-The sequence-based fitness estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
-However, the Mpro experiments all use a similar yeast-based methodology~\citep{flynn2022,flynn2023,iketani2022functional} that probably fails to capture significant aspects of Mpro's actual function during human infections.
-For instance, a stop codon at Q306 is well tolerated in the yeast-based Mpro deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would clearly be highly deleterious to actual virus as it would truncate the polyprotein.
+The sequence-based estimates for Mpro also correlate with the deep mutational scans for that protein, although in this case the experiments correlate substantially better with each other than with our estimates (Figure~\ref{fig:dms}B).
+However, the Mpro experiments all use a similar yeast-based methodology~\citep{flynn2022,flynn2023,iketani2022functional} that fails to capture significant aspects of Mpro's function during human infections.
+For instance, a stop codon at Q306 is well tolerated in the deep mutational scans but extremely disfavorable in our sequence-based estimates, and such a mutation would clearly be highly deleterious to actual virus as it would truncate the polyprotein.
 Similarly, K61N is well tolerated in the deep mutational scans but extremely disfavorable in our estimates, probably because in the full viral polyprotein this residue mediates important interactions between Mpro and nsp7-10~\citep{yadav2022biochemical}
@@
 These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
-A known limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
+A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
 Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
@@
 
-Our mutation-effect estimates also correlate substantially better with deep mutational scanning experiments than predictions from two methods trained to learn mutation effects and epistatic interactions from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
-Our estimates also correlate better with experiments than predictions of a machine-learning method that integrates a wide range of sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
-These results suggest that our relatively straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.
+Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
+Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
+These results suggest that our straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.
 
@@
 
-To enable easy access to the mutation-effect estimates, we created an interactive plots to enable exploration of the data for each protein.
+To enable easy access to the mutation-effect estimates, we created interactive plots to enable exploration of the data for each protein.
 A static view of one of these plots is in Figure~\ref{fig:E_heatmap}; see \url{https://jbloomlab.github.io/SARS2-mut-fitness} for interactive versions for all proteins.
@@
 There are several caveats to our approach.
-First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control of publicly available sequences to remove those affected by sequencing errors.
+First, because the number of observations of any given mutation is small compared to the millions of SARS-CoV-2 sequences being analyzed, our approach requires careful quality control to remove sequencing errors.
 Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
-Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}.
+Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~\ref{fig:synonymous}).
 Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
@@
 See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
-The specific version of the repository used for this paper is tagged as ``bioRxiv-v1'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v1})
+The specific version of the repository used for this paper is tagged as ``bioRxiv-v2'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v2})
 The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.
@@
 The interactive plots are rendered at \url{https://jbloomlab.github.io/SARS2-mut-fitness} via GitHub pages.
+
+\subsection{Versioning of analyses of different sequence sets}
+The figures in this manuscript show analyses of the set of all publicly available sequences as of May-11-2023.
+However, the pipeline can be run on different sequence sets.
+The sequence sets on which the analysis is currently run include all sets listed under the ``mat\_trees'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}; these include the sets of all public sequences from several earlier dates (such as those available for the first version of this analysis), as well as the set of all sequences in GISAID as of March-29-2023.
+A version of the results for each sequence set is provided in the GitHub repository (\url{https://github.com/jbloomlab/SARS2-mut-fitness} ) in subdirectories with names like ``results\_public\_2023-05-11'', and the index page for the interactive plots (\url{https://jbloomlab.github.io/SARS2-mut-fitness}) links at the bottom to plots for each sequence set.
+The ``current\_mat'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml} specifies which sequence set is used to generate the main results that in the ``results'' subdirectory in the GitHub repository and are shown by default in the interactive plats; we anticipate periodically updating this to newer sequence sets as more sequences become available.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/mat_aa_fitness_correlations.html} for the correlations among mutation effects estimated from the different sequence sets.
+
+For the GISAID sequence set, we acknowledge the submitters of the sequences listed at the following URLs: \url{https://doi.org/10.55876/gis8.230403ab}, \url{https://doi.org/10.55876/gis8.230403hg}, \url{https://doi.org/10.55876/gis8.230403ht}, and \url{https://doi.org/10.55876/gis8.230403tg}.
 
@@
 \end{itemize}
-The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors, and the rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
+The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors of viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
+Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
+The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
 We ignore branches with multiple nucleotide mutations to the same codon (this is very rare) because as detailed below our method is only designed to make estimates for mutations that represent single-nucleotide changes from the clade founder.
@@
 Typically if there was just one clade, we would set the wildtype value of $f_x = 0$ and then for mutations to all other amino acids $y$ we would simply have $f_y = \Delta f_{xy}$.
-However, when there are multple clades with different founder amino acids, there is no longer a well defined "wildtype".
+However, when there are multiple clades with different founder amino acids, there is no longer a well defined "wildtype".
 So we choose the most common non-stop parental amino-acid for the observed mutations and set that to zero.
@@
 \subsection{Comparison to deep mutational scanning}
-Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
+Deep mutational scanning data were taken from published studies~\citep{dadonaite2023pseudovirus,starr2022deep,flynn2022,flynn2023,iketani2022functional}, using the data at the links specified under the ``dms\_datasets'' key in \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/config.yaml}.
 For the spike deep mutational scanning~\citep{dadonaite2023pseudovirus} we only included mutations with ``times seen'' values of at least three in the deep mutational scanning.
 The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
-In both cases, Figure~\ref{fig:dms} shows the effect averaged across all phenotypes measured by the study.
+Figures~\ref{fig:dms}, \ref{fig:dnds}, and \ref{fig:comparator} shows the effect averaged across all phenotypes measured by each of these studies.
 For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.
+
+\subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
+For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
+The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
+We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
+The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.
+
+We also compared both our mutation-effect estimates and the spike deep mutational scanning measurements~\citep{dadonaite2023pseudovirus} to predictions from three other algorithms:
+\begin{itemize}
+\item the EpiScores reported by Maher et al (2022)~\cite{maher2022predicting},
+\item the DCA mutability scores reported by Rodriguez-Rivas et al (2022)~\cite{rodriguez2022epistatic}, and
+\item the EVE scores reported by Thadani et al (2023)~\cite{thadani2022learning}.
+\end{itemize}
+These comparisons are shown in Figure~\ref{fig:comparator}.
+The Maher et al (2022) and Thadani et al (2023) studies report mutation-level predictions and so are compared directly to the deep mutational scanning our mutation-effect estimates; Rodriguez-Rivas et al (2023) report only site-level metrics and so are compared to site-summary metrics as for the dN/dS analysis.
 
@@
 \section{Acknowledgments}
-We thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built mutation-annotated tree.
+We thank Angie Hinrichs for providing the pre-built mutation-annotated trees and thank the UShER team for promptly answering and addressing GitHub issues related to use of this package and its pre-built trees.
+We thank the sequence submitters to GISAID, who are listed in the tables cited in the Methods section.
 The work of JDB was supported in part by the NIH/NIAID under grants U19AI171399 and R01AI141707, and contract 75N93021C00015.
@@
 \section{Competing interests}
-JDB is on the scientific advisory boards of Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, and Oncorus.
+JDB consults Apriori Bio, Aerium Therapeutics, Invivyd, the Vaccine Company, GSK, and Pfizer on topics related to viral evolution.
 JDB receives royalty payments as an inventor on Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
@@
 
+\begin{figure*}
+\centering
+\includegraphics[width=0.7\linewidth]{figs/nmut_counts.pdf}
+\caption{
+Number of mutations per branch on all branches of the SARS-CoV-2 tree with at least one mutation.
+The analysis used here only considers mutations on branches with four or fewer mutations and so excludes the small fraction of highly mutated branches.
+\label{fig:nmut_counts}
+}
+\end{figure*}
+
 \end{document}

### 2023-06-03 paper/paper.tex
revision of paper based on initial reviews (#36)
* update all existing figures (new figs and text revisions not yet added)

* add new figures (not linked to text / legends yet)

* revise first two results sections

* add dn/ds and comparator predictions

* revision of paper

* generate track-changes manuscript

* finish revising paper

* updated track changes manuscript

@@
 
-\subsection*{Mutation-effect estimates are robust to subsampling and phylogeny, with some evidence of epistasis in spike}
+\subsection*{Mutation-effect estimates are robust to subsampling, with some evidence of epistasis in spike}
 
@@
 The lower cross-clade correlation for spike appears due to epistatic shifts in mutation effects~\citep{starr2022shifting, moulana2022compensatory, pollock2012amino, shah2015contingency, lee2018deep} or changes in the selective landscape~\citep{sun2023rapidly} between SARS-CoV-2 clades, since the correlation is lower between clades with higher spike divergence (Figure~\ref{fig:corr}D).
+In particular, the interactive version of Figure~\ref{fig:corr}A shows that mutations that are more beneficial in Omicron BA.5 than Delta are often antibody-escape mutations (eg, K444N or G446S in spike~\citep{greaney2022antibody})---a result that makes sense, since newer variants like Omicron BA.5 are evolving under increased immune selection compared to earlier variants like Delta that circulated in a more immunologically naive population.
 
@@
 This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
-%We also note that the fact that two trees (which have independently inferred topologies) give nearly identical fitness estimates indicates that uncertainty in tree topology has minimal effect.
-% The insensitivity of our fitness estimates to uncertainty in tree topology is expected: the main uncertainty in densely sampled viral phylogenies involves polytomies of closely related sequences, but our estimates only use the counts of mutations along tree branches, which are unaffected by uncertainty in the resolution of polytomies of closely related sequences.
 
@@
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
+See Figure~\ref{fig:effects_dist_position_order} for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
 \label{fig:dist}
@@
 
-To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B).
+To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
 SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
-The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).
 
-However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B).
+However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}).
 Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
 The lack of deleterious mutations to ORF8 is consistent with the fact that viruses with deletions in this gene have spread in humans~\citep{su2020discovery} and that major variants had stop codons early in ORF8.
+Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}. 
 The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
@@
 A longstanding approach for analyzing protein constraint is to compare rates of nonsynonymous (dN) and synonymous (dS) substitutions at each site~\citep{nielsen1998likelihood,kosakovsky2005not}.
-These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not}.
+These dN/dS ratios can be calculated by counting mutations or using phylogenetic substitution models~\citep{kosakovsky2005not,yang2000estimating}.
 A limitation of dN/dS ratios is they cannot be interpreted in terms of the fitness effects, since they simply represent the relative rate of amino-acid substitution rather than the effects of specific mutations~\citep{spielman2015relationship,kryazhimskiy2008population}.
@@
 
+\begin{figure*}
+\includegraphics[width=\linewidth]{figs/effects_dist_position_order.pdf}
+\caption{
+A version of Figure~\ref{fig:dist}B but with genes ordered by position in the genome rather than extent of constraint on nonsynonymous mutations.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist_position_order.html} for an interactive version of this plot.
+\label{fig:effects_dist_position_order}
+}
+\end{figure*}
+
+\begin{figure*}[h]
+\includegraphics[width=\linewidth]{figs/ORF3a.pdf}
+\caption{
+Effects of stop-codon and amino-acid mutations across ORF3a.
+The black area plot shows the mean effect of all amino-acid mutations at each site, and the purple points show the effects of stop codon mutations.
+There is strong selection against stop codons (negative effects) for all but the C-terminus of ORF3a, but only a few positions show strong selection against amino-acid substitutions.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/ORF3a.html} for an interactive version of this plot along with zoomable heatmap of the effects of specific amino-acid substitutions.
+\label{fig:ORF3a}
+}
+\end{figure*}
+
 \begin{figure*}[h]
@@
 \label{fig:comparator}
-}
-\end{figure*}
-
-\begin{figure*}[h]
-\includegraphics[width=\linewidth]{figs/ORF3a.pdf}
-\caption{
-Effects of stop-codon and amino-acid mutations across ORF3a.
-The black area plot shows the mean effect of all amino-acid mutations at each site, and the purple points show the effects of stop codon mutations.
-There is strong selection against stop codons (negative effects) for all but the C-terminus of ORF3a, but only a few positions show strong selection against amino-acid substitutions.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/ORF3a.html} for an interactive version of this plot along with zoomable heatmap of the effects of specific amino-acid substitutions.
-\label{fig:ORF3a}
 }

### 2023-06-04 paper/paper.tex
fix figure reference

@@
 SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
-The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).
 
-However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_positon_order}).
+However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
 Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).

### 2023-06-05 paper/paper.tex
Richard Neher final edits (#37)
* fix figure reference

* add note on regression line and seq/bioinfo errors

---------

Co-authored-by: Richard Neher <richard.neher@unibas.ch>

@@
 Only mutations with at least 10 expected counts are shown, which is why panels have different numbers of mutations shown (sequence subsets vary in size).
+Different subset size are also the reason why the regression line in (A) deviates from the identity $x=y$.
 {\bf (C)} Correlations between clade or geography subsets become higher with an increasingly large threshold for minimum expected counts.
@@
 
-An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations even after aggregating across clades.
+An important question is whether the mutation fitness estimates are affected by noise from limited statistical sampling of mutations or whether sequencing errors and bioinformatic artefacts distort the estimates.
 To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
@@
 This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
+Furthermore, fitness estimates using sequences from specific countries (USA and England, Figure~\ref{fig:corr}B) are also highly concordant, suggesting that sequencing and bioinformatic workflows are not driving the signal.
+Lastly, positions known to be under strong constraint {\it a priori} (e.g.~start codons, the ribosomal slippage site) typically have no or only few mutations, suggesting that sequencing errors in consensus sequences are rare.
 

### 2023-08-07 paper/paper.tex
fixed editor comment 1 (typo)

@@
 That repository contains a README with links to specific data files as well as a description of the computational pipeline.
-See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
-The specific version of the repository used for this paper is tagged as ``bioRxiv-v2'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v2})
+See \url{https://github.com/jbloomlab/SARS2-mut-fitness/blob/main/results/aa_fitness/aa_fitness.csv} for final estimates of amino-acid fitnesses across all clades; other intermediate data files are also provided in the GitHub repository.
+The specific version of the repository used for this paper is tagged as ``bioRxiv-v2'' on GitHub (\url{https://github.com/jbloomlab/SARS2-mut-fitness/tree/bioRxiv-v2}).
 The pipeline is fully reproducible, and is run using \texttt{snakemake}~\citep{molder2021sustainable} with interactive plots rendered using \texttt{altair}~\citep{vanderplas2018altair}.
@@
 \end{itemize}
-The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors of viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
+The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
 Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).

### 2023-08-07 paper/paper.tex
add discussion of decreased selection in last 1/3 of genome (editor point 2)

@@
 Observations such as these could help guide experimental studies to better understand protein function.
+
+The accessory proteins are all encoded in the last third of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the nucleocapsid gene) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
+This finding is consistent with observations that the genes encoded in the last third of the genome vary across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
+This pattern of decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.
 

### 2023-08-08 paper/paper.tex
Reviewer 1 comment 2: y-axis limits in Fig S5
paper revisions for Virus Evolution (#40)
* remove old track changes paper

* fixed editor comment 1 (typo)

* add discussion of decreased selection in last 1/3 of genome (editor point 2)

* Reviewer 1 comment 2: y-axis limits in Fig S5

* reviewer 1 comment 2: dN/dS

@@
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
-The handful of synonymous mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
+The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
 
@@
 For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
+Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
+We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
+In order to ensure comparability in the sequence sets used, for the comparisons in Figure~\ref{fig:dnds}, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
+We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.
+
 The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
@@
 The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version of this plot.
+Note that the y-axis in each panel is extended so that it fully contains the trend line over the entire x-axis domain, explaining why the y-axis limits vary across the rows even though the range of the data are the same on the y-axis for each row.
+Both the fitness estimates and dN/dS values shown here are calculated using only sequences available as of Jan-31-2022 (see Methods).
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/public_2022-01-31/dnds_corr.html} for an interactive version of this plot, and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version that uses fitness estimates from all SARS-CoV-2 sequences including more recent ones.
 \label{fig:dnds}

### 2023-08-08 paper/paper.tex
edit: change last third to last 5kb

@@
 
-The accessory proteins are all encoded in the last third of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the nucleocapsid gene) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
-This finding is consistent with observations that the genes encoded in the last third of the genome vary across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
+The accessory proteins are all encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
+This finding is consistent with observations that the genes encoded in the last part of the genome vary across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
 This pattern of decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.

### 2023-08-08 paper/paper.tex
made paragraph on last 5 kb of genome a bit more succinct

--- 2023-08-08 paper/paper.tex
@@
 
-The accessory proteins are all encoded in the last third of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the nucleocapsid gene) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
-This finding is consistent with observations that the genes encoded in the last third of the genome vary across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
-This pattern of decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.
+The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
+Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
+Decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.
 

### 2023-08-11 paper/final_submission/paper.tex
made version of `paper.tex` with refs in file and supp excluded for final submission

--- 2023-08-08 paper/paper.tex
@@
 To assess if this is the case, we repeated the entire fitness estimation using an even larger pre-built UShER mutation-annotated tree of all $\sim$14-million SARS-CoV-2 sequences in GISAID~\cite{shu2017gisaid} as of March-29-2023.
-There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~\ref{fig:gisaid_corr}).
+There is an extremely high correlation between fitness estimates made using the $\sim$7-million publicly available sequences and the larger GISAID tree (Figure~S1).
 This concordance indicates that the set of $\sim$7-million public sequences is large enough that doubling the data does not appreciably shift the estimates, and so throughout this paper we use that sequence set due to our preference for publicly available data.
@@
 See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_histogram.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist.html} for plots that allow adjustment of the expected-count cutoff and other interactive options (such as separate histograms for each gene).
-See Figure~\ref{fig:effects_dist_position_order} for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
+See Figure~S3 for a version of panel B with genes ordered by genomic position rather than constraint on nonsynonymous mutations.
 \label{fig:dist}
@@
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
-The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
+The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~S2).
 
-To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
+To investigate differences in functional constraint among viral proteins, we computed the distributions of mutation effects separately for each gene (Figures~\ref{fig:dist}B and S3).
 SARS-CoV-2 proteins are grouped into three categories: nonstructural (or nsp) proteins, structural proteins (spike, M, N, and E), and accessory proteins (names prefixed with ``ORF'')~\citep{v2021coronavirus}.
-The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}; e.g., nsp13 is under stronger protein-level constraint than nsp1).
+The nonstructural and structural proteins are essential, and these proteins show strong selection against stop codons and clear although variable purifying selection against amino-acid mutations (Figures~\ref{fig:dist}B and S3; e.g., nsp13 is under stronger protein-level constraint than nsp1).
 
-However, most accessory proteins are under little constraint (Figure~\ref{fig:dist}B and \ref{fig:effects_dist_position_order}).
+However, most accessory proteins are under little constraint (Figures~\ref{fig:dist}B and S3).
 Stop-codon and amino-acid mutations to ORF7a and ORF8 are not more deleterious than synonymous mutations (although recall that our estimates are only sensitive to fitness costs greater than a few percent).
@@
 Indeed, the loss of accessory proteins such as ORF8 appears to occur with some regularity during the early evolution of non-human viruses in humans~\citep{rochman2022molecular}.
-The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~\ref{fig:ORF3a}).
+The only accessory protein under strong purifying selection against stop codons is ORF3a (Figure~\ref{fig:dist}B), for which stop codons in the first ~240 residues are clearly deleterious (Figure~S4).
 These observations concur with experiments showing SARS-CoV-2 is attenuated by deletion of ORF3a but there is little effect of deleting ORF6, ORF7a, or ORF8~\citep{mcgrath2022sars,silvas2021contribution,liu2022live}.
-However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~\ref{fig:ORF3a}).
+However, ORF3a's function must be relatively insensitive to its protein sequence, since other than selection against stop codons there is only amino-acid level constraint at a few sites like 135 and 138 (Figure~S4).
 Observations such as these could help guide experimental studies to better understand protein function.
 
-The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
+The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~S3).
 Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
@@
 Nonetheless, we can compare dN/dS ratios to our mutation-effect estimates as measures of the average constraint at each site.
-The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~\ref{fig:dnds}).
+The mutation-effect estimates greatly outperform dN/dS ratios as a measure of site-level constraint as assessed by correlation with deep mutational scanning experiments (Figure~S5).
 The reason is in part because some aspects of functional constraint cannot be captured by a dN/dS ratio.
@@
 
-Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~\ref{fig:comparator}).
-Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~\ref{fig:comparator}).
+Our mutation-effect estimates also correlate better with deep mutational scanning experiments than predictions from two algorithms trained to learn epistatic models of mutation effects from phylogenetically broader but more sparsely sampled sequence data~\cite{thadani2022learning,rodriguez2022epistatic} (Figure~S6).
+Our estimates also correlate better with experiments than predictions by a machine-learning algorithm that integrates sequence and epidemiological data~\cite{maher2022predicting} (Figure~S6).
 These results suggest that our straightforward approach of directly reading out the effects of mutations from their actual versus expected counts can outperform much more complex models when millions of sequences are available.
@@
 \subsection*{Fixed mutations tend to have beneficial or neutral effects}
-Amino-acid mutations that have fixed in at least one viral clade are estimated to mostly have neutral or beneficial effects, whereas most other mutations are deleterious (Figure~\ref{fig:fixed_dist}).
+Amino-acid mutations that have fixed in at least one viral clade are estimated to mostly have neutral or beneficial effects, whereas most other mutations are deleterious (Figure~S7).
 This fact is unsurprising: viral lineages that expand into new clades do so because they have acquired beneficial mutations while avoiding deleterious ones~\citep{luksza2014predictive, koelle2015effects, huddleston2020integrating}.
@@
 
-Most fixed mutations are estimated to be beneficial regardless of whether estimates are made using all viral clades, or just clades that did not fix the mutation (Figure~\ref{fig:fixed}).
-However, a few beneficial fixed mutations show epistatic entrenchment~\citep{shah2015contingency, starr2018pervasive} in the sense that they are not particularly beneficial in clades in which they did not fix (Figure~\ref{fig:fixed}).
+Most fixed mutations are estimated to be beneficial regardless of whether estimates are made using all viral clades, or just clades that did not fix the mutation (Figure~S8).
+However, a few beneficial fixed mutations show epistatic entrenchment~\citep{shah2015contingency, starr2018pervasive} in the sense that they are not particularly beneficial in clades in which they did not fix (Figure~S8).
 The most striking example is S373P in spike, which has experimentally been shown to be neutral or slightly deleterious in pre-Omicron clades, but strongly beneficial in the Omicron clades in which it fixed~\citep{starr2022deep, moulana2022compensatory}.
@@
 Second, we assume the rate of each type of nucleotide mutation is uniform across the viral genome, and neglect higher-order context that may influence mutation rate~\citep{sadykov2021short,beale2004comparison}.
-Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~\ref{fig:synonymous}).
+Likewise, we neglect constraint on nucleotide identity beyond the encoded protein sequence~\citep{huston2021comprehensive,kuo2013functional}---although this probably has only a minor effect, since our analyses show just a handful of synonymous sites are under strong selection (Figure~S2).
 Third, the exact relationship between the statistics we calculate and viral fitness depend on the fraction of all infections that are sequenced (sampling intensity) and viral population dynamics.
@@
 Our approach is better for estimating effects of deleterious or nearly neutral mutations, but clade-growth methods may be better for beneficial mutations.
-In particular, clade size carries information beyond that contained in mutation counts alone (Figure~\ref{fig:terminal}).
+In particular, clade size carries information beyond that contained in mutation counts alone (Figure~S9).
 Hopefully future work can combine mutation-counting and clade-growth methods for even better estimates of SARS-CoV-2 mutation effects.
@@
 The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
-Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
+Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~S10).
 The rationale for the second and third exclusions is that excess reversions can arise from base-calling pipelines that erroneously call low-coverage sites as reference.
@@
 The RBD data~\citep{starr2022deep} include measurements for two phenotypes (ACE2 affinity and RBD expression), and one of the Mpro studies~\citep{flynn2022} includes measurements for three different phenotypes in yeast (growth, FRET, and transcription factor activity).
-Figures~\ref{fig:dms}, \ref{fig:dnds}, and \ref{fig:comparator} shows the effect averaged across all phenotypes measured by each of these studies.
+Figures~\ref{fig:dms}, S5, and S6 shows the effect averaged across all phenotypes measured by each of these studies.
 For plots that break the correlations out by phenotype, see \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_S_all_corr.html} and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dms_nsp5_all_corr.html}.
@@
 \subsection{Comparison to dN/dS and other mutation-effect prediction algorithms}
-For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
+For the comparison to the dN/dS approaches shown in Figure~S5, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
 Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
 We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
-In order to ensure comparability in the sequence sets used, for the comparisons in Figure~\ref{fig:dnds}, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
+In order to ensure comparability in the sequence sets used, for the comparisons in Figure~S5, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
 We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.
@@
 We therefore computed site-summary metrics of the mutation-effect estimates and the deep mutational scanning as the average effect of all measured amino-acid mutations at each site, excluding stop codons.
-The correlations in Figure~\ref{fig:dnds} are with those site-summary metrics.
+The correlations in Figure~S5 are with those site-summary metrics.
 
@@
 \end{itemize}
-These comparisons are shown in Figure~\ref{fig:comparator}.
+These comparisons are shown in Figure~S6.
 The Maher et al (2022) and Thadani et al (2023) studies report mutation-level predictions and so are compared directly to the deep mutational scanning our mutation-effect estimates; Rodriguez-Rivas et al (2023) report only site-level metrics and so are compared to site-summary metrics as for the dN/dS analysis.
@@
 
-\bibliography{references}
+\begin{thebibliography}{76}
+
+\bibitem[{Abdool~Karim {\rm and} de~Oliveira}(2021)]{abdool2021new}
+{Abdool~Karim, S.~S. {\rm and} T.~de~Oliveira}, 2021 New {SARS-CoV-2}
+  variants—clinical, public health, and vaccine implications. New England
+  Journal of Medicine {\bf 384}: 1866--1868.
+
+\bibitem[{Acevedo {\em et~al.\/}}(2014){Acevedo, Brodsky, {\rm and}
+  Andino}]{acevedo2014mutational}
+{Acevedo, A., L.~Brodsky, {\rm and} R.~Andino}, 2014 Mutational and fitness
+  landscapes of an {RNA} virus revealed through population sequencing. Nature
+  {\bf 505}: 686--690.
+
+\bibitem[{Aksamentov {\em et~al.\/}}(2021){Aksamentov, Roemer, Hodcroft, {\rm
+  and} Neher}]{aksamentov2021nextclade}
+{Aksamentov, I., C.~Roemer, E.~B. Hodcroft, {\rm and} R.~A. Neher}, 2021
+  Nextclade: clade assignment, mutation calling and quality control for viral
+  genomes. Journal of Open Source Software {\bf 6}: 3773.
+
+\bibitem[{Beale {\em et~al.\/}}(2004){Beale, Petersen-Mahrt, Watt, Harris,
+  Rada, {\rm and} Neuberger}]{beale2004comparison}
+{Beale, R.~C., S.~K. Petersen-Mahrt, I.~N. Watt, R.~S. Harris, C.~Rada, {\em
+  et~al.\/}}, 2004 {Comparison of the differential context-dependence of DNA
+  deamination by APOBEC enzymes: correlation with mutation spectra in vivo}.
+  Journal of Molecular Biology {\bf 337}: 585--596.
+
+\bibitem[{Bhatt {\em et~al.\/}}(2021){Bhatt, Scaiola, Loughran, Leibundgut,
+  Kratzel, Meurs, Dreos, O’Connor, McMillan, Bode, {\em
+  et~al.\/}}]{bhatt2021structural}
+{Bhatt, P.~R., A.~Scaiola, G.~Loughran, M.~Leibundgut, A.~Kratzel, {\em
+  et~al.\/}}, 2021 Structural basis of ribosomal frameshifting during
+  translation of the {SARS-CoV-2 RNA} genome. Science {\bf 372}: 1306--1313.
+
+\bibitem[{Bloom {\em et~al.\/}}(2023){Bloom, Beichman, Neher, {\rm and}
+  Harris}]{bloom2023evolution}
+{Bloom, J.~D., A.~C. Beichman, R.~A. Neher, {\rm and} K.~Harris}, 2023
+  Evolution of the {SARS-CoV-2} mutational spectrum. Molecular Biology and
+  Evolution {\bf 40}: msad085.
+
+\bibitem[{Cao {\em et~al.\/}}(2022{a}){Cao, Jian, Wang, Yu, Song, Yisimayi, An,
+  Chen, Zhang, Wang, {\em et~al.\/}}]{cao2022imprinted}
+{Cao, Y., F.~Jian, J.~Wang, Y.~Yu, W.~Song, {\em et~al.\/}}, 2022{a} {Imprinted
+  SARS-CoV-2 humoral immunity induces convergent Omicron RBD evolution}. Nature
+  {\bf DOI 10.1038/s41586-022-05644-7}.
+
+\bibitem[{Cao {\em et~al.\/}}(2022{b}){Cao, Jian, Zhang, Yisimayi, Hao, Bao,
+  Yuan, Yu, Du, Wang, {\em et~al.\/}}]{cao2022rational}
+{Cao, Y., F.~Jian, Z.~Zhang, A.~Yisimayi, X.~Hao, {\em et~al.\/}}, 2022{b}
+  Rational identification of potent and broad sarbecovirus-neutralizing
+  antibody cocktails from {SARS} convalescents. Cell reports {\bf 41}: 111845.
+
+\bibitem[{Dadonaite {\em et~al.\/}}(2023){Dadonaite, Crawford, Radford,
+  Farrell, Timothy, Hannon, Zhou, Andrabi, Burton, Liu, {\em
+  et~al.\/}}]{dadonaite2023pseudovirus}
+{Dadonaite, B., K.~H. Crawford, C.~E. Radford, A.~G. Farrell, C.~Y. Timothy,
+  {\em et~al.\/}}, 2023 {A pseudovirus system enables deep mutational scanning
+  of the full SARS-CoV-2 spike}. Cell {\bf 186}: 1263--1278.
+
+\bibitem[{Dawson {\rm and} Folimonova}(2013)]{dawson2013virus}
+{Dawson, W.~O. {\rm and} S.~Y. Folimonova}, 2013 Virus-based transient
+  expression vectors for woody crops: a new frontier for vector design and use.
+  Annual Review of Phytopathology {\bf 51}: 321--337.
+
+\bibitem[{De~Maio {\em et~al.\/}}(2021){De~Maio, Walker, Turakhia, Lanfear,
+  Corbett-Detig, {\rm and} Goldman}]{de2021mutation}
+{De~Maio, N., C.~R. Walker, Y.~Turakhia, R.~Lanfear, R.~Corbett-Detig, {\em
+  et~al.\/}}, 2021 Mutation rates and selection on synonymous mutations in
+  {SARS-CoV-2}. Genome Biology and Evolution {\bf 13}: evab087.
+
+\bibitem[{DeGrace {\em et~al.\/}}(2022){DeGrace, Ghedin, Frieman, Krammer,
+  Grifoni, Alisoltani, Alter, Amara, Baric, Barouch, {\em
+  et~al.\/}}]{degrace2022defining}
+{DeGrace, M.~M., E.~Ghedin, M.~B. Frieman, F.~Krammer, A.~Grifoni, {\em
+  et~al.\/}}, 2022 Defining the risk of {SARS-CoV-2} variants on immune
+  protection. Nature {\bf 605}: 640--652.
+
+\bibitem[{Flynn {\em et~al.\/}}(2023){Flynn, Huang, Zvornicanin,
+  Schneider-Nachum, Shaqra, Kurt~Yilmaz, Moquin, Dovala, Schiffer, {\rm and}
+  Bolon}]{flynn2023}
+{Flynn, J.~M., Q.~Y.~M. Huang, S.~N. Zvornicanin, G.~Schneider-Nachum, A.~M.
+  Shaqra, {\em et~al.\/}}, 2023 Systematic analyses of the resistance potential
+  of drugs targeting {SARS-CoV-2} main protease. bioRxiv {\bf DOI
+  10.1101/2023.03.02.530652}.
+
+\bibitem[{Flynn {\em et~al.\/}}(2022){Flynn, Samant, Schneider-Nachum, Barkan,
+  Yilmaz, Schiffer, Moquin, Dovala, {\rm and} Bolon}]{flynn2022}
+{Flynn, J.~M., N.~Samant, G.~Schneider-Nachum, D.~T. Barkan, N.~K. Yilmaz, {\em
+  et~al.\/}}, 2022 Comprehensive fitness landscape of {SARS-CoV-2 Mpro} reveals
+  insights into viral resistance mechanisms. eLife {\bf 11}: e77433.
+
+\bibitem[{Greaney {\em et~al.\/}}(2022){Greaney, Starr, {\rm and}
+  Bloom}]{greaney2022antibody}
+{Greaney, A.~J., T.~N. Starr, {\rm and} J.~D. Bloom}, 2022 An antibody-escape
+  estimator for mutations to the {SARS-CoV-2} receptor-binding domain. Virus
+  Evolution {\bf 8}: veac021.
+
+\bibitem[{Greaney {\em et~al.\/}}(2021){Greaney, Starr, Gilchuk, Zost,
+  Binshtein, Loes, Hilton, Huddleston, Eguia, Crawford, {\em
+  et~al.\/}}]{greaney2021complete}
+{Greaney, A.~J., T.~N. Starr, P.~Gilchuk, S.~J. Zost, E.~Binshtein, {\em
+  et~al.\/}}, 2021 Complete mapping of mutations to the {SARS-CoV-2} spike
+  receptor-binding domain that escape antibody recognition. Cell Host \&
+  Microbe {\bf 29}: 44--57.
+
+\bibitem[{Harvey {\em et~al.\/}}(2021){Harvey, Carabelli, Jackson, Gupta,
+  Thomson, Harrison, Ludden, Reeve, Rambaut, Peacock, {\em
+  et~al.\/}}]{harvey2021sars}
+{Harvey, W.~T., A.~M. Carabelli, B.~Jackson, R.~K. Gupta, E.~C. Thomson, {\em
+  et~al.\/}}, 2021 {SARS-CoV-2} variants, spike mutations and immune escape.
+  Nature Reviews Microbiology {\bf 19}: 409--424.
+
+\bibitem[{Hiscox {\em et~al.\/}}(2021){Hiscox, Khoo, Stewart, {\rm and}
+  Owen}]{hiscox2021shutting}
+{Hiscox, J.~A., S.~H. Khoo, J.~P. Stewart, {\rm and} A.~Owen}, 2021 Shutting
+  the gate before the horse has bolted: is it time for a conversation about
+  {SARS-CoV-2} and antiviral drug resistance? Journal of Antimicrobial
+  Chemotherapy {\bf 76}: 2230--2233.
+
+\bibitem[{Huddleston {\em et~al.\/}}(2020){Huddleston, Barnes, Rowe, Xu,
+  Kondor, Wentworth, Whittaker, Ermetal, Daniels, McCauley, {\em
+  et~al.\/}}]{huddleston2020integrating}
+{Huddleston, J., J.~R. Barnes, T.~Rowe, X.~Xu, R.~Kondor, {\em et~al.\/}}, 2020
+  Integrating genotypes and phenotypes improves long-term forecasts of seasonal
+  influenza {A/H3N2} evolution. Elife {\bf 9}: e60067.
+
+\bibitem[{Huston {\em et~al.\/}}(2021){Huston, Wan, Strine, Tavares, Wilen,
+  {\rm and} Pyle}]{huston2021comprehensive}
+{Huston, N.~C., H.~Wan, M.~S. Strine, R.~d. C.~A. Tavares, C.~B. Wilen, {\em
+  et~al.\/}}, 2021 Comprehensive in vivo secondary structure of the
+  {SARS-CoV-2} genome reveals novel regulatory motifs and mechanisms. Molecular
+  Cell {\bf 81}: 584--598.
+
+\bibitem[{Iketani {\em et~al.\/}}(2022{a}){Iketani, Hong, Sheng, Bahari,
+  Culbertson, Atanaki, Aditham, Kratz, Luck, Tian, {\em
+  et~al.\/}}]{iketani2022functional}
+{Iketani, S., S.~J. Hong, J.~Sheng, F.~Bahari, B.~Culbertson, {\em et~al.\/}},
+  2022{a} {Functional map of SARS-CoV-2 3CL protease reveals tolerant and
+  immutable sites}. Cell Host \& Microbe {\bf 30}: 1354--1362.
+
+\bibitem[{Iketani {\em et~al.\/}}(2022{b}){Iketani, Mohri, Culbertson, Hong,
+  Duan, Luck, Annavajhala, Guo, Sheng, Uhlemann, {\em
+  et~al.\/}}]{iketani2022multiple}
+{Iketani, S., H.~Mohri, B.~Culbertson, S.~J. Hong, Y.~Duan, {\em et~al.\/}},
+  2022{b} Multiple pathways for {SARS-CoV-2} resistance to nirmatrelvir. Nature
+  {\bf DOI 10.1038/s41586-022-05514-2}.
+
+\bibitem[{Jungreis {\em et~al.\/}}(2021){Jungreis, Nelson, Ardern, Finkel,
+  Krogan, Sato, Ziebuhr, Stern-Ginossar, Pavesi, Firth, {\em
+  et~al.\/}}]{jungreis2021conflicting}
+{Jungreis, I., C.~W. Nelson, Z.~Ardern, Y.~Finkel, N.~J. Krogan, {\em
+  et~al.\/}}, 2021 {Conflicting and ambiguous names of overlapping ORFs in the
+  SARS-CoV-2 genome: A homology-based resolution}. Virology {\bf 558}:
+  145--151.
+
+\bibitem[{Koelle {\rm and} Rasmussen}(2015)]{koelle2015effects}
+{Koelle, K. {\rm and} D.~A. Rasmussen}, 2015 The effects of a deleterious
+  mutation load on patterns of influenza {A/H3N2's} antigenic evolution in
+  humans. Elife {\bf 4}: e07361.
+
+\bibitem[{Kosakovsky~Pond {\rm and} Frost}(2005)]{kosakovsky2005not}
+{Kosakovsky~Pond, S.~L. {\rm and} S.~D. Frost}, 2005 Not so different after
+  all: a comparison of methods for detecting amino acid sites under selection.
+  Molecular biology and evolution {\bf 22}: 1208--1222.
+
+\bibitem[{Kryazhimskiy {\rm and} Plotkin}(2008)]{kryazhimskiy2008population}
+{Kryazhimskiy, S. {\rm and} J.~B. Plotkin}, 2008 The population genetics of
+  {dN/dS}. PLoS Genetics {\bf 4}: e1000304.
+
+\bibitem[{Kuo {\rm and} Masters}(2013)]{kuo2013functional}
+{Kuo, L. {\rm and} P.~S. Masters}, 2013 Functional analysis of the murine
+  coronavirus genomic {RNA} packaging signal. Journal of Virology {\bf 87}:
+  5182--5192.
+
+\bibitem[{Lanfear}(2020)]{lanfear2020}
+{Lanfear, R.}, 2020 {A global phylogeny of SARS-CoV-2 sequences from GISAID}.
+  Zenodo {\bf DOI 10.5281/zenodo.3958883}.
+
+\bibitem[{Lee {\em et~al.\/}}(2022){Lee, Sohail, Finney, Ahmed, Quadeer, McKay,
+  {\rm and} Barton}]{lee2022inferring}
+{Lee, B., M.~S. Sohail, E.~Finney, S.~F. Ahmed, A.~A. Quadeer, {\em et~al.\/}},
+  2022 Inferring effects of mutations on {SARS}-{CoV}-2 transmission from
+  genomic surveillance data. medRxiv {\bf 10.1101/2021.12.31.21268591v1}:
+  2021.12.31.21268591.
+
+\bibitem[{Lee {\em et~al.\/}}(2018){Lee, Huddleston, Doud, Hooper, Wu, Bedford,
+  {\rm and} Bloom}]{lee2018deep}
+{Lee, J.~M., J.~Huddleston, M.~B. Doud, K.~A. Hooper, N.~C. Wu, {\em
+  et~al.\/}}, 2018 Deep mutational scanning of hemagglutinin helps predict
+  evolutionary fates of human {H3N2} influenza variants. Proceedings of the
+  National Academy of Sciences {\bf 115}: E8276--E8285.
+
+\bibitem[{Liu {\em et~al.\/}}(2014){Liu, Fung, Chong, Shukla, {\rm and}
+  Hilgenfeld}]{liu2014accessory}
+{Liu, D.~X., T.~S. Fung, K.~K.-L. Chong, A.~Shukla, {\rm and} R.~Hilgenfeld},
+  2014 Accessory proteins of {SARS-CoV} and other coronaviruses. Antiviral
+  Research {\bf 109}: 97--109.
+
+\bibitem[{Liu {\em et~al.\/}}(2022){Liu, Zhang, Liu, Xia, Zou, Muruato,
+  Periasamy, Kurhade, Plante, Bopp, {\em et~al.\/}}]{liu2022live}
+{Liu, Y., X.~Zhang, J.~Liu, H.~Xia, J.~Zou, {\em et~al.\/}}, 2022 A
+  live-attenuated {SARS-CoV-2} vaccine candidate with accessory protein
+  deletions. Nature Communications {\bf 13}: 1--14.
+
+\bibitem[{Llanes {\em et~al.\/}}(2020){Llanes, Restrepo, Caballero, Rajeev,
+  Kennedy, {\rm and} Lleonart}]{llanes2020betacoronavirus}
+{Llanes, A., C.~M. Restrepo, Z.~Caballero, S.~Rajeev, M.~A. Kennedy, {\em
+  et~al.\/}}, 2020 Betacoronavirus genomes: how genomic information has been
+  used to deal with past outbreaks and the {COVID-19} pandemic. International
+  Journal of Molecular Sciences {\bf 21}: 4546.
+
+\bibitem[{{\L}uksza {\rm and} L{\"a}ssig}(2014)]{luksza2014predictive}
+{{\L}uksza, M. {\rm and} M.~L{\"a}ssig}, 2014 A predictive fitness model for
+  influenza. Nature {\bf 507}: 57--61.
+
+\bibitem[{Maher {\em et~al.\/}}(2022){Maher, Bartha, Weaver, Di~Iulio, Ferri,
+  Soriaga, Lempp, Hie, Bryson, Berger, {\em et~al.\/}}]{maher2022predicting}
+{Maher, M.~C., I.~Bartha, S.~Weaver, J.~Di~Iulio, E.~Ferri, {\em et~al.\/}},
+  2022 Predicting the mutational drivers of future {SARS-CoV-2} variants of
+  concern. Science translational medicine {\bf 14}: eabk3445.
+
+\bibitem[{Markov {\em et~al.\/}}(2023){Markov, Ghafari, Beer, Lythgoe,
+  Simmonds, Stilianakis, {\rm and} Katzourakis}]{markov2023evolution}
+{Markov, P.~V., M.~Ghafari, M.~Beer, K.~Lythgoe, P.~Simmonds, {\em et~al.\/}},
+  2023 The evolution of {SARS}-{CoV}-2. Nature Reviews Microbiology pp. 1--19,
+  Publisher: Nature Publishing Group.
+
+\bibitem[{Martin {\em et~al.\/}}(2021){Martin, Weaver, Tegally, San, Shank,
+  Wilkinson, Lucaci, Giandhari, Naidoo, Pillay, {\em
+  et~al.\/}}]{martin2021emergence}
+{Martin, D.~P., S.~Weaver, H.~Tegally, J.~E. San, S.~D. Shank, {\em et~al.\/}},
+  2021 {The emergence and ongoing convergent evolution of the SARS-CoV-2 N501Y
+  lineages}. Cell {\bf 184}: 5189--5200.
+
+\bibitem[{McBroome {\em et~al.\/}}(2021){McBroome, Thornlow, Hinrichs, Kramer,
+  De~Maio, Goldman, Haussler, Corbett-Detig, {\rm and}
+  Turakhia}]{mcbroome2021daily}
+{McBroome, J., B.~Thornlow, A.~S. Hinrichs, A.~Kramer, N.~De~Maio, {\em
+  et~al.\/}}, 2021 A daily-updated database and tools for comprehensive
+  {SARS-CoV-2} mutation-annotated trees. Molecular Biology and Evolution {\bf
+  38}: 5819--5824.
+
+\bibitem[{McGrath {\em et~al.\/}}(2022){McGrath, Xue, Dillen, Oldfield,
+  Assad-Garcia, Zaveri, Singh, Baracco, Taylor, Vashee, {\em
+  et~al.\/}}]{mcgrath2022sars}
+{McGrath, M., Y.~Xue, C.~Dillen, L.~Oldfield, N.~Assad-Garcia, {\em et~al.\/}},
+  2022 {SARS-CoV-2} variant spike and accessory gene mutations alter
+  pathogenesis. Proceedings National Academy of Sciences USA {\bf 119}:
+  e2204717119.
+
+\bibitem[{Moghadasi {\em et~al.\/}}(2023){Moghadasi, Heilmann, Khalil,
+  Nnabuife, Kearns, Ye, Moraes, Costacurta, Esler, Aihara, Laer,
+  Martinez-Sobrido, Palzkill, Amaro, {\rm and} Harris}]{moghadasi2023}
+{Moghadasi, S.~A., E.~Heilmann, A.~M. Khalil, C.~Nnabuife, F.~L. Kearns, {\em
+  et~al.\/}}, 2023 Transmissible {SARS-CoV-2} variants with resistance to
+  clinical protease inhibitors. Science Advances {\bf 9}: eade8778.
+
+\bibitem[{M{\"o}lder {\em et~al.\/}}(2021){M{\"o}lder, Jablonski, Letcher,
+  Hall, Tomkins-Tinch, Sochat, Forster, Lee, Twardziok, Kanitz, {\em
+  et~al.\/}}]{molder2021sustainable}
+{M{\"o}lder, F., K.~P. Jablonski, B.~Letcher, M.~B. Hall, C.~H. Tomkins-Tinch,
+  {\em et~al.\/}}, 2021 Sustainable data analysis with snakemake. F1000Research
+  {\bf 10}.
+
+\bibitem[{Moulana {\em et~al.\/}}(2022){Moulana, Dupic, Phillips, Chang,
+  Nieves, Roffler, Greaney, Starr, Bloom, {\rm and}
+  Desai}]{moulana2022compensatory}
+{Moulana, A., T.~Dupic, A.~M. Phillips, J.~Chang, S.~Nieves, {\em et~al.\/}},
+  2022 {Compensatory epistasis maintains ACE2 affinity in SARS-CoV-2 Omicron
+  BA. 1}. Nature Communications {\bf 13}: 7011.
+
+\bibitem[{Neher}(2022)]{neher2022contributions}
+{Neher, R.~A.}, 2022 Contributions of adaptation and purifying selection to
+  {SARS-CoV-2} evolution. Virus Evolution {\bf 8}: veac113.
+
+\bibitem[{Nielsen {\rm and} Yang}(1998)]{nielsen1998likelihood}
+{Nielsen, R. {\rm and} Z.~Yang}, 1998 Likelihood models for detecting
+  positively selected amino acid sites and applications to the {HIV-1} envelope
+  gene. Genetics {\bf 148}: 929--936.
+
+\bibitem[{Obermeyer {\em et~al.\/}}(2022){Obermeyer, Jankowiak, Barkas,
+  Schaffner, Pyle, Yurkovetskiy, Bosso, Park, Babadi, MacInnis, Luban, Sabeti,
+  {\rm and} Lemieux}]{obermeyer2022analysis}
+{Obermeyer, F., M.~Jankowiak, N.~Barkas, S.~F. Schaffner, J.~D. Pyle, {\em
+  et~al.\/}}, 2022 Analysis of 6.4 million {SARS}-{CoV}-2 genomes identifies
+  mutations associated with fitness. Science {\bf 376}: 1327--1332.
+
+\bibitem[{Pollock {\em et~al.\/}}(2012){Pollock, Thiltgen, {\rm and}
+  Goldstein}]{pollock2012amino}
+{Pollock, D.~D., G.~Thiltgen, {\rm and} R.~A. Goldstein}, 2012 Amino acid
+  coevolution induces an evolutionary {Stokes} shift. Proceedings of the
+  National Academy of Sciences {\bf 109}: E1352--E1359.
+
+\bibitem[{Rappazzo {\em et~al.\/}}(2021){Rappazzo, Tse, Kaku, Wrapp, Sakharkar,
+  Huang, Deveau, Yockachonis, Herbert, Battles, {\em
+  et~al.\/}}]{rappazzo2021broad}
+{Rappazzo, C.~G., L.~V. Tse, C.~I. Kaku, D.~Wrapp, M.~Sakharkar, {\em
+  et~al.\/}}, 2021 Broad and potent activity against {SARS-like} viruses by an
+  engineered human monoclonal antibody. Science {\bf 371}: 823--829.
+
+\bibitem[{Rochman {\em et~al.\/}}(2022){Rochman, Wolf, {\rm and}
+  Koonin}]{rochman2022molecular}
+{Rochman, N.~D., Y.~I. Wolf, {\rm and} E.~V. Koonin}, 2022 Molecular
+  adaptations during viral epidemics. EMBO reports {\bf 23}: e55393.
+
+\bibitem[{Rodriguez-Rivas {\em et~al.\/}}(2022){Rodriguez-Rivas, Croce, Muscat,
+  {\rm and} Weigt}]{rodriguez2022epistatic}
+{Rodriguez-Rivas, J., G.~Croce, M.~Muscat, {\rm and} M.~Weigt}, 2022 Epistatic
+  models predict mutable sites in {SARS-CoV-2} proteins and epitopes.
+  Proceedings of the National Academy of Sciences {\bf 119}: e2113118119.
+
+\bibitem[{Ruis {\em et~al.\/}}(2022){Ruis, Peacock, Polo, Masone, Alvarez,
+  Hinrichs, Turakhia, Ye, McBroome, Corbett-Detig, {\em
+  et~al.\/}}]{ruis2022mutational}
+{Ruis, C., T.~P. Peacock, L.~M. Polo, D.~Masone, M.~S. Alvarez, {\em
+  et~al.\/}}, 2022 Mutational spectra distinguish {SARS-CoV-2} replication
+  niches. bioRxiv {\bf DOI 10.1101/2022.09.27.509649}.
+
+\bibitem[{Sadykov {\em et~al.\/}}(2021){Sadykov, Mourier, Guan, {\rm and}
+  Pain}]{sadykov2021short}
+{Sadykov, M., T.~Mourier, Q.~Guan, {\rm and} A.~Pain}, 2021 {Short sequence
+  motif dynamics in the SARS-CoV-2 genome suggest a role for cytosine
+  deamination in CpG reduction}. Journal of Molecular Cell Biology {\bf 13}:
+  225--227.
+
+\bibitem[{Shah {\em et~al.\/}}(2015){Shah, McCandlish, {\rm and}
+  Plotkin}]{shah2015contingency}
+{Shah, P., D.~M. McCandlish, {\rm and} J.~B. Plotkin}, 2015 Contingency and
+  entrenchment in protein evolution under purifying selection. Proceedings of
+  the National Academy of Sciences {\bf 112}: E3226--E3235.
+
+\bibitem[{Shu {\rm and} McCauley}(2017)]{shu2017gisaid}
+{Shu, Y. {\rm and} J.~McCauley}, 2017 {GISAID}: Global initiative on sharing
+  all influenza data--from vision to reality. Eurosurveillance {\bf 22}: 30494.
+
+\bibitem[{Silvas {\em et~al.\/}}(2021){Silvas, Vasquez, Park, Chiem,
+  Allu{\'e}-Guardia, Garcia-Vilanova, Platt, Miorin, Kehrer, Cupic, {\em
+  et~al.\/}}]{silvas2021contribution}
+{Silvas, J.~A., D.~M. Vasquez, J.-G. Park, K.~Chiem, A.~Allu{\'e}-Guardia, {\em
+  et~al.\/}}, 2021 {Contribution of SARS-CoV-2 accessory proteins to viral
+  pathogenicity in K18 human ACE2 transgenic mice}. Journal of Virology {\bf
+  95}: e00402--21.
+
+\bibitem[{Spielman {\rm and} Wilke}(2015)]{spielman2015relationship}
+{Spielman, S.~J. {\rm and} C.~O. Wilke}, 2015 The relationship between {dN/dS}
+  and scaled selection coefficients. Molecular biology and evolution {\bf 32}:
+  1097--1108.
+
+\bibitem[{Starr {\em et~al.\/}}(2021){Starr, Czudnochowski, Liu, Zatta, Park,
+  Addetia, Pinto, Beltramello, Hernandez, Greaney, {\em
+  et~al.\/}}]{starr2021sars}
+{Starr, T.~N., N.~Czudnochowski, Z.~Liu, F.~Zatta, Y.-J. Park, {\em et~al.\/}},
+  2021 {SARS-CoV-2 RBD} antibodies that maximize breadth and resistance to
+  escape. Nature {\bf 597}: 97--102.
+
+\bibitem[{Starr {\em et~al.\/}}(2018){Starr, Flynn, Mishra, Bolon, {\rm and}
+  Thornton}]{starr2018pervasive}
+{Starr, T.~N., J.~M. Flynn, P.~Mishra, D.~N. Bolon, {\rm and} J.~W. Thornton},
+  2018 Pervasive contingency and entrenchment in a billion years of {Hsp90}
+  evolution. Proceedings of the National Academy of Sciences {\bf 115}:
+  4453--4458.
+
+\bibitem[{Starr {\em et~al.\/}}(2022{a}){Starr, Greaney, Hannon, Loes, Hauser,
+  Dillen, Ferri, Farrell, Dadonaite, McCallum, {\em
+  et~al.\/}}]{starr2022shifting}
+{Starr, T.~N., A.~J. Greaney, W.~W. Hannon, A.~N. Loes, K.~Hauser, {\em
+  et~al.\/}}, 2022{a} Shifting mutational constraints in the {SARS-CoV-2}
+  receptor-binding domain during viral evolution. Science {\bf 377}: 420--424.
+
+\bibitem[{Starr {\em et~al.\/}}(2020){Starr, Greaney, Hilton, Ellis, Crawford,
+  Dingens, Navarro, Bowen, Tortorici, Walls, {\em et~al.\/}}]{starr2020deep}
+{Starr, T.~N., A.~J. Greaney, S.~K. Hilton, D.~Ellis, K.~H. Crawford, {\em
+  et~al.\/}}, 2020 {Deep mutational scanning of SARS-CoV-2 receptor binding
+  domain reveals constraints on folding and ACE2 binding}. Cell {\bf 182}:
+  1295--1310.
+
+\bibitem[{Starr {\em et~al.\/}}(2022{b}){Starr, Greaney, Stewart, Walls,
+  Hannon, Veesler, {\rm and} Bloom}]{starr2022deep}
+{Starr, T.~N., A.~J. Greaney, C.~M. Stewart, A.~C. Walls, W.~W. Hannon, {\em
+  et~al.\/}}, 2022{b} {Deep mutational scans for ACE2 binding, RBD expression,
+  and antibody escape in the SARS-CoV-2 Omicron BA. 1 and BA. 2
+  receptor-binding domains}. PLoS Pathogens {\bf 18}: e1010951.
+
+\bibitem[{Su {\em et~al.\/}}(2020){Su, Anderson, Young, Linster, Zhu,
+  Jayakumar, Zhuang, Kalimuddin, Low, Tan, {\em et~al.\/}}]{su2020discovery}
+{Su, Y.~C., D.~E. Anderson, B.~E. Young, M.~Linster, F.~Zhu, {\em et~al.\/}},
+  2020 {Discovery and genomic characterization of a 382-nucleotide deletion in
+  ORF7b and ORF8 during the early evolution of SARS-CoV-2}. mBio {\bf 11}:
+  e01610--20.
+
+\bibitem[{Sun {\em et~al.\/}}(2023){Sun, Tempia, Kleynhans, von Gottberg,
+  McMorrow, Wolter, Bhiman, Moyes, Carrim, Martinson, {\em
+  et~al.\/}}]{sun2023rapidly}
+{Sun, K., S.~Tempia, J.~Kleynhans, A.~von Gottberg, M.~L. McMorrow, {\em
+  et~al.\/}}, 2023 {Rapidly shifting immunologic landscape and severity of
+  SARS-CoV-2 in the Omicron era in South Africa}. Nature Communications {\bf
+  14}: 246.
+
+\bibitem[{Syed {\em et~al.\/}}(2021){Syed, Taha, Tabata, Chen, Ciling, Khalid,
+  Sreekumar, Chen, Hayashi, Soczek, {\em et~al.\/}}]{syed2021rapid}
+{Syed, A.~M., T.~Y. Taha, T.~Tabata, I.~P. Chen, A.~Ciling, {\em et~al.\/}},
+  2021 Rapid assessment of {SARS-CoV-2}--evolved variants using virus-like
+  particles. Science {\bf 374}: 1626--1632.
+
+\bibitem[{Tao {\em et~al.\/}}(2021){Tao, Tzou, Nouhin, Bonilla, Jagannathan,
+  {\rm and} Shafer}]{tao2021sars}
+{Tao, K., P.~L. Tzou, J.~Nouhin, H.~Bonilla, P.~Jagannathan, {\em et~al.\/}},
+  2021 {SARS-CoV-2} antiviral therapy. Clinical microbiology reviews {\bf 34}:
+  e00109--21.
+
+\bibitem[{Thadani {\em et~al.\/}}(2022){Thadani, Gurev, Notin, Youssef,
+  Rollins, Sander, Gal, {\rm and} Marks}]{thadani2022learning}
+{Thadani, N.~N., S.~Gurev, P.~Notin, N.~Youssef, N.~J. Rollins, {\em
+  et~al.\/}}, 2022 Learning from pre-pandemic data to forecast viral antibody
+  escape. bioRxiv {\bf DOI 10.1101/2022.07.21.501023}.
+
+\bibitem[{Thorne {\em et~al.\/}}(2022){Thorne, Bouhaddou, Reuschl,
+  Zuliani-Alvarez, Polacco, Pelin, Batra, Whelan, Hosmillo, Fossati, {\em
+  et~al.\/}}]{thorne2022evolution}
+{Thorne, L.~G., M.~Bouhaddou, A.-K. Reuschl, L.~Zuliani-Alvarez, B.~Polacco,
+  {\em et~al.\/}}, 2022 Evolution of enhanced innate immune evasion by
+  {SARS-CoV-2}. Nature {\bf 602}: 487--495.
+
+\bibitem[{Turakhia {\em et~al.\/}}(2020){Turakhia, De~Maio, Thornlow, Gozashti,
+  Lanfear, Walker, Hinrichs, Fernandes, Borges, Slodkowicz, {\em
+  et~al.\/}}]{turakhia2020stability}
+{Turakhia, Y., N.~De~Maio, B.~Thornlow, L.~Gozashti, R.~Lanfear, {\em
+  et~al.\/}}, 2020 Stability of {SARS-CoV-2} phylogenies. PLoS Genetics {\bf
+  16}: e1009175.
+
+\bibitem[{Turakhia {\em et~al.\/}}(2021){Turakhia, Thornlow, Hinrichs, De~Maio,
+  Gozashti, Lanfear, Haussler, {\rm and} Corbett-Detig}]{turakhia2021ultrafast}
+{Turakhia, Y., B.~Thornlow, A.~S. Hinrichs, N.~De~Maio, L.~Gozashti, {\em
+  et~al.\/}}, 2021 {Ultrafast Sample placement on Existing tRees (UShER)
+  enables real-time phylogenetics for the SARS-CoV-2 pandemic}. Nature Genetics
+  {\bf 53}: 809--816.
+
+\bibitem[{Tzou {\em et~al.\/}}(2022){Tzou, Tao, Pond, {\rm and}
+  Shafer}]{tzou2022coronavirus}
+{Tzou, P.~L., K.~Tao, S.~L.~K. Pond, {\rm and} R.~W. Shafer}, 2022 {Coronavirus
+  Resistance Database (CoV-RDB): SARS-CoV-2 susceptibility to monoclonal
+  antibodies, convalescent plasma, and plasma from vaccinated persons}. Plos
+  one {\bf 17}: e0261045.
+
+\bibitem[{VanderPlas {\em et~al.\/}}(2018){VanderPlas, Granger, Heer, Moritz,
+  Wongsuphasawat, Satyanarayan, Lees, Timofeev, Welsh, {\rm and}
+  Sievert}]{vanderplas2018altair}
+{VanderPlas, J., B.~Granger, J.~Heer, D.~Moritz, K.~Wongsuphasawat, {\em
+  et~al.\/}}, 2018 Altair: interactive statistical visualizations for {Python}.
+  Journal of Open Source Software {\bf 3}: 1057.
+
+\bibitem[{V’kovski {\em et~al.\/}}(2021){V’kovski, Kratzel, Steiner,
+  Stalder, {\rm and} Thiel}]{v2021coronavirus}
+{V’kovski, P., A.~Kratzel, S.~Steiner, H.~Stalder, {\rm and} V.~Thiel}, 2021
+  Coronavirus biology and replication: implications for {SARS-CoV-2}. Nature
+  Reviews Microbiology {\bf 19}: 155--170.
+
+\bibitem[{Wang {\em et~al.\/}}(2011){Wang, Sharma, Duffy, {\rm and}
+  Almeida}]{wang2011genetic}
+{Wang, J., A.~M. Sharma, S.~Duffy, {\rm and} R.~P. Almeida}, 2011 Genetic
+  diversity in the 3' terminal 4.7-kb region of grapevine leafroll-associated
+  virus 3. Phytopathology {\bf 101}: 445--450.
+
+\bibitem[{Weisblum {\em et~al.\/}}(2020){Weisblum, Schmidt, Zhang, DaSilva,
+  Poston, Lorenzi, Muecksch, Rutkowska, Hoffmann, Michailidis, {\em
+  et~al.\/}}]{weisblum2020escape}
+{Weisblum, Y., F.~Schmidt, F.~Zhang, J.~DaSilva, D.~Poston, {\em et~al.\/}},
+  2020 Escape from neutralizing antibodies by {SARS-CoV-2} spike protein
+  variants. eLife {\bf 9}: e61312.
+
+\bibitem[{Yadav {\em et~al.\/}}(2022){Yadav, Courouble, Dey, Harrison, Timm,
+  Hopkins, Slack, Sarafianos, Ruiz, Griffin, {\em
+  et~al.\/}}]{yadav2022biochemical}
+{Yadav, R., V.~V. Courouble, S.~K. Dey, J.~J.~E. Harrison, J.~Timm, {\em
+  et~al.\/}}, 2022 {Biochemical and structural insights into SARS-CoV-2
+  polyprotein processing by Mpro}. Science Advances {\bf 8}: eadd2191.
+
+\bibitem[{Yang {\rm and} Nielsen}(2000)]{yang2000estimating}
+{Yang, Z. {\rm and} R.~Nielsen}, 2000 Estimating synonymous and nonsynonymous
+  substitution rates under realistic evolutionary models. Molecular biology and
+  evolution {\bf 17}: 32--43.
+
+\bibitem[{Zanini {\em et~al.\/}}(2017){Zanini, Puller, Brodin, Albert, {\rm
+  and} Neher}]{zanini2017vivo}
+{Zanini, F., V.~Puller, J.~Brodin, J.~Albert, {\rm and} R.~A. Neher}, 2017 In
+  vivo mutation rates and the landscape of fitness costs of {HIV}-1. Virus
+  Evolution {\bf 3}: vex003.
+
+\end{thebibliography}
+
 }
 
-\newgeometry{left=1in, right=1in, top=48pt, bottom=60pt}
-\onecolumn
-\renewcommand{\thepage}{S\arabic{page}}
-\setcounter{page}{1}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-
-\clearpage
-
-\section{Supplementary appendix deriving relationship between fitness cost and ratio of expected to actual counts}
-
-With millions of SARS-CoV-2 sequences shared publicly, almost all mutations that are tolerated by the virus are observed dozens to hundreds of times.
-Where on the tree and how often on the tree we observe specific mutations has information about the effects of these mutations on viral spread.
-The mutation rate depends on the nucleotides involved and possibly on the sequence context and other viral determinants, but for the purpose of this derivation, we will assume the neutral rate $\mu$ is known.
-If the mutation is neutral, the total number of times the mutation is observed on the tree is $\mu T$, where $T$ is the total length of the tree (assuming that the mutation never reached high frequency which is true for almost all mutations, particularly when mutations are counted on a per-clade basis relative too the clade founder as done above).
-
-If a mutation reduces fitness, the lineages descending from branches on which this mutation happened will spread more slowly than those without this mutation.
-As a result, the down-stream subclades are smaller and more short lived, which in turn means that they will be less likely to be sampled and represented in the tree.
-To infer a mutation's effect on fitness, we need to calculate how the probability of observation depends on this fitness effect.
-
-For a mutation to be represented in the tree, one of its descendants has to be sampled and sequenced.
-If the total number of descendants is $w$ and the sampling fraction is $\epsilon$, the probability that the mutation is present in the tree is
-\begin{equation}
-    P = 1 - e^{-w\epsilon}
-\end{equation}
-$W$ is a random number that depends on the realization of the transmission process, which is commonly modeled by a branching process with birth rate $b$ and death rate $d$.
-The death rate here corresponds to clearing an infection, the birth rate to onward transmission. The latter is affected by the fitness cost of the mutation.
-
-To obtain insight how the probability of observing a lineage depends on parameters, we calculate the probability $p(w, T|t)$ that a lineage had an integrated size $w = \int_t^T k(t') \; dt'$, where $t$ is the birth time of the lineage, $T$ is the current time, and $k(t')$ is the size of the lineage at time $t'$.
-To calculate $p(w,T|k)$, we generalize it slightly to $p(W, T|k,t)$, where $k$ is the number of individuals at the start time $t$. This quantity obeys the following ``first-step'' equation:
-\begin{equation}
-    -(\partial_t - k \partial_w) p(w,T|k,t) = -k(b+d) p(w,T|k,t) + k b p(w,T|k+1,t) + k d p(w,T| k-1, t)
-\end{equation}
-We will solve for the Laplace transform $\hat{p}(z,T|k,t) = \int_0^\infty dw\; e^{-wz} p(w,T|k,t) = \hat{p}^k(z,T|1,t)$.
-Using the following identity for the derivative of the Laplace transform
-\begin{equation}
-    \begin{split}
-        \int_0^\infty e^{-wz} \partial_w p  \; dw& = [e^{-wz} p]_0^\infty - \int_0^\infty  p \partial_w e^{-wz} \; dw = 0 + z \int_0^\infty  p e^{-wz} = z \hat{p}
-    \end{split}
-\end{equation}
-and setting $k=1$, we have
-\begin{equation}
-    -\partial_t \hat{p}(z,T|t) = -(b+d+z) \hat{p}(z,T|t) + b \hat{p}^2(z,T|t) + d
-\end{equation}
-This simplifies further to if we substitute $\phi(z,T|t) = 1 - \hat{p}(z,T|t)$.
-\begin{equation}
-    \begin{split}
-        \partial_t \phi(z,T|t) & = -(b+d+z) (1-\phi(z,T|t)) + b (1-\phi(z,T|t))^2 + d \\
-         & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
-    \end{split}
-\end{equation}
-where it is important to note that the derivative is with respect to the first time point and the interval $T-t$ is shrinking with increasing $t$.
-
-\subsection*{Constant birth and death rate}
-If the fitness effect of the mutation in question is detrimental and the overall population is constant (background $b_0=d_0$), all mutant lineages will eventually die out and we can consider large $T-t$ and the long time asymptotic $\partial_t \phi(z,T|t)=0$.
-Further define $b=b_0 - s$ and $d=d_0$ where $s$ is the fitness cost of the mutation (so larger values indicate a greater fitness cost).
-The steady state generating function is then
-\begin{equation}
-    0 = -z - (b-d-z) \phi(z) + b \phi(z)^2
-\end{equation}
-with solution
-\begin{equation}
-    \begin{split}
-        \phi(z) & = -\frac{s + z}{2(b_0 - s)} \pm \frac{\sqrt{(s+z)^2 + 4z(b_0-s)}}{2(b_0-s)} \\
-        &\approx -\frac{s + z}{2b_0} \pm \frac{\sqrt{(s+z)^2 + 4z b_0}}{2b_0} \\
-        & \approx \begin{cases}
-            \frac{z}{s+z} & (s+z)^2 \gg 4zb_0 \\
-            \sqrt{\frac{z}{b_0}}\left(1+\frac{(s+z)^2}{8zb_0}\right)-\frac{s + z}{2b_0}  & (s+z)^2 \ll 4zb_0 \\
-        \end{cases}
-    \end{split}
-\end{equation}
-Since $\phi(z) = 1 - \int e^{-wz}p(w) \; dw$, $\phi(\epsilon)$ is exactly the probability that a lineage is sampled when the entire population is sampled at rate $\epsilon$.
-We thus expect two regimes: if the square of the fitness cost exceeds the sampling intensity (typically at 1\% or less), the probability of sampling a lineage is essentially inversely proportional to the fitness cost.
-The sampling probability of lineages with smaller costs effects depends less strongly on $s$.
-Their sampling mostly comes down to stochasticity independent of the fitness cost.
-
-\subsection*{Growing populations}
-In many scenarios relevant for lineages that arise during a viral outbreak, the background population isn't constant but is undergoing a rapid exponential expansion.
-The background birth rate $b_0$ is bigger than $d_0$ in this case.
-Since the population is growing, deleterious mutations can increase in frequency deterministically and we can not send the $t$ to infinity as before.
-Instead, we need to integrate
-\begin{equation}
-    \begin{split}
-        \partial_t \phi(z,T|t) & = -z - (b-d-z) \phi(z,T|t) + b \phi(z,T|t)^2
-    \end{split}
-\end{equation}
-backwards in time starting from $\phi(z,T|T)=0$ at $t=T$.
-While $\phi(z,T|t)$ is small and the quadratic term can be neglected, this is approximately solved by
-\begin{equation}
-\begin{split}
-    \phi(z,T|t) & = z e^{\int_t^T (b-d-z)dt'} \int_t^T e^{-\int_\tau^T (b-d-z)dt'} d\tau \\
-    & = z e^{(b-d-z)(T-t)} \int_t^T e^{-(b-d-z)(T-\tau)} d\tau \\
-    & = z e^{(b-d-z)(T-t)}\left[1   - e^{-(b-d-z)(T-t)}\right]/(b-d-z) \\
-    & = \frac{z}{b-d-z}\left[e^{(b-d-z)(T-t)} - 1\right] = \frac{z}{\gamma_0 - s - z}\left[e^{\gamma_0 (T-t) - (s+z) (T-t)} - 1\right]
-\end{split}
-\end{equation}
-where $\gamma_0$ is the growth rate of the background population.
-
-At longer times when $z e^{\gamma_0 (T-t)}\sim 1$ and $\phi$ is no longer small, $\phi$ tends towards a constant value determined by the same quadratic equation as above.
-This limit is neither interesting or relevant for the present purpose, since there are very few lineages that emerged early enough to have saturated $\phi$.
-Instead, we need to average $\phi$ (the linear approximation) over all the time points when the lineage could have arisen.
-\begin{equation}
-    \begin{split}
-        \langle \phi \rangle &\sim \int_t^T dt' \; e^{-\gamma_0(T-t')} \frac{z (e^{\gamma_0(T-t') - (s+z)(T-t')}-1)}{(\gamma_0 - s - z)}\\
-        &= \int_t^T dt' \;  \frac{z (e^{-(s+z)(T-t')}-e^{-\gamma_0(T-t')})}{(\gamma_0 - s - z)}\\
-        &\approx \begin{cases}
-            \frac{z}{\gamma_0-s-z} [ \frac{1}{z+s} - \frac{1}{\gamma_0}] & s(T-t)\gg 1\\
-            \frac{z}{\gamma_0-s-z} [ (T-t) - \frac{(s+z)(T-t)^2}{2} - \frac{1}{\gamma_0}] & s(T-t)<1 \\
-        \end{cases}
-    \end{split}
-\end{equation}
-This derivation assumed that $\gamma_0(T-t)\gg1$, i.e.~that the overall population size has expanded substantially.
-The most relevant fitness effects will be those with $s(T-t)>1$, that is the fitness effect has strong effect on variant frequency, but $s<\gamma_0$ such that the variant is still spreading and can give rise to large lineages in an expanding variant.
-In this case, the above simplifies to
-\begin{equation}
-    \langle \phi \rangle \approx \frac{z}{\gamma_0(z+s)}
-\end{equation}
-
-In a variant that has been growing with rate $\gamma_0$ for a time $\tau=T-t$ and sampled with $z=\epsilon$, we thus expect that the number of times we observe separate mutant lineages depends on $s$ as
-\begin{equation}
-    \langle \phi \rangle \approx \frac{\epsilon}{\gamma_0(\epsilon+s)}
-\end{equation}
-This has a very similar behavior as the solution for constant population size, which suggests that the overall dependence on $s$ is robust and we can assume that the number of times a mutation is observed is inversely proportional to its effect on fitness.
-The same basic dependency is observed at steady state in a quasi-species \citep{zanini2017vivo}.
-In a constant population, this relationship breaks down for dense sampling $\epsilon > \sqrt{s}$.
-In growing population, the approximation fails if the product of fitness effect and the time over which the variant has grown, $s\tau$, is small, i.e., if the fitness cost does not affect variant frequency strongly.
-In these cases, there is still a dependence on $s$, but it is weaker.
-
-\clearpage
-
-\section{Supplementary figures}
-
-\begin{figure*}[h]
-\centering
-\includegraphics[width=0.4\linewidth]{figs/gisaid_corr.png}
-\caption{
-Correlation between amino-acid fitness estimates from the $\sim$7-million sequence set of all publicly available sequences as of May-11-2023, or the set of all $\sim$14-million sequences in GISAID as of March-29-2023.
-The orange text in the upper-left corner of the plot shows the Pearson correlation.
-The high correlation indicates that our estimates are not substantially limited by noise related to statistical sampling of mutation counts, since using another tree with twice as many sequences does not substantially alter the estimates.
-Note that the main text figures all use the smaller $\sim$7-million publicly available sequence set.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/mat_aa_fitness_correlations.html} for an interactive version of this plot that also shows the correlations for other sets of publicly available sequences.
-\label{fig:gisaid_corr}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=\linewidth]{figs/synonymous_constraint_figure.pdf}
-\caption{
-{\bf Non-coding constraint is concentrated in a few isolated regions in SARS-CoV-2 genome.}
-{\bf (A), (B)} The average fitness cost of synonymous mutations (do not change amino acids) in ORF1ab, S, E, M, N, and ORF3a, non-coding regions between ORFs, and all positions in ORF6, ORF7a, ORF7b, ORF8, ORF10 (as these ORFs tolerate stop codons, see Figure~\ref{fig:dist}B).
-Panel A shows the entire genome, panel B zooms into the 3' end of the genome.
-Fitness effects are averaged in sliding windows of size 25 and 50.
-With the exception of a small number of well defined regions, in most of the genome these averages fluctuate around zero.
-Regions with large fitness costs for non-coding mutations correspond to the ribosomal slippage site at the end of ORF1a, the transcription regulatory sites of S, ORF3a, E, M, and N, as well as two regions in the center of E and the end of M (see red bars in the top part of each panel) \citep{markov2023evolution}.
-Notably, TRS sites of ORF6, ORF7a and ORF8 don't show strong signal of conservation, consistent with their tolerance of stop codons.
-{\bf (C), (D)} The distribution of inferred fitness cost of four-fold degenerate sites at regions of known non-coding constraint within the protein coding genes (the ribosomal slippage and TRS sites) versus all other four-fold degenerate sites.
-In total, 25 four-fold degenerate sites are in these regions of known constraint, whereas 4844 four-fold degenerate sites are outside these regions.
-Panel C shows that a large fraction of mutations in these known regions of constraint are deleterious, while the distribution for the remaining four-fold degenerate sites is centered around 0 between -2 and +1.
-Furthermore, strongly deleterious mutations with fitness estimate below -3 predominantly fall in the 25 sites of known non-coding constraint.
-If the two additional conserved regions in E and M are included, all but one mutation outside these regions has a fitness cost estimate above -3.
-\label{fig:synonymous}
-}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figs/effects_dist_position_order.pdf}
-\caption{
-A version of Figure~\ref{fig:dist}B but with genes ordered by position in the genome rather than extent of constraint on nonsynonymous mutations.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/effects_dist_position_order.html} for an interactive version of this plot.
-\label{fig:effects_dist_position_order}
-}
-\end{figure*}
-
-\begin{figure*}[h]
-\includegraphics[width=\linewidth]{figs/ORF3a.pdf}
-\caption{
-Effects of stop-codon and amino-acid mutations across ORF3a.
-The black area plot shows the mean effect of all amino-acid mutations at each site, and the purple points show the effects of stop codon mutations.
-There is strong selection against stop codons (negative effects) for all but the C-terminus of ORF3a, but only a few positions show strong selection against amino-acid substitutions.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/ORF3a.html} for an interactive version of this plot along with zoomable heatmap of the effects of specific amino-acid substitutions.
-\label{fig:ORF3a}
-}
-\end{figure*}
-
-\begin{figure*}[h]
-\centering
-\includegraphics[width=\linewidth]{figs/dnds.png}
-\caption{
-The site-mean of the estimated fitness effects outperforms dN/dS ratios as an indicator of functional constraint as assessed by correlation with deep mutational scanning measurements.
-{\bf (A)}
-Pearson correlation of log dN/dS ratio at each site with the mean estimated fitness of all amino acid mutations at each site from the current study.
-For most genes, the log dN/dS ratio is correlated with the mean fitness effect, but the correlations are only modest (always substantially less than one).
-The exception is the accessory proteins which are mostly under minimal amino-acid level selection (Figure~\ref{fig:dist}B) and only show a weak correlation.
-{\bf (B)}
-Correlation of the log dN/dS ratio or the mean estimated fitness effect at each site with the mean impact of amino-acid mutations at that site as experimentally measured by deep mutational scanning.
-The numbers in the upper right of each panel give the Pearson correlation.
-In every case, the mean estimated fitness effect (top row) is much better correlated with the experimental measurements than the log dN/dS (bottom row).
-The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
-Note that the y-axis in each panel is extended so that it fully contains the trend line over the entire x-axis domain, explaining why the y-axis limits vary across the rows even though the range of the data are the same on the y-axis for each row.
-Both the fitness estimates and dN/dS values shown here are calculated using only sequences available as of Jan-31-2022 (see Methods).
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/public_2022-01-31/dnds_corr.html} for an interactive version of this plot, and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version that uses fitness estimates from all SARS-CoV-2 sequences including more recent ones.
-\label{fig:dnds}
-}
-\end{figure*}
-
-\begin{figure*}[h]
-\centering
-\includegraphics[width=0.75\linewidth]{figs/comparator_corr.png}
-\caption{
-The mutation effect estimates from the current study are substantially better correlated with deep mutational scanning measurements for spike than predicted mutation effects made using three other approaches.
-Each row shows the correlation of the fitness estimates from the current study and full-spike deep mutational scanning measurements versus values from another study.
-The Maher et al (2022) and Thadani et (2023) studies make predictions for individual amino-acids, so those correlations are with the mutation-level fitness estimates and deep mutational scanning measurements.
-The Rodriguez-Rivas et al (2022) study only reports site-level predictions, so those correlations are with the site mean fitness estimates and deep mutational scanning measurements.
-Each row only shows mutations or sites with a reported value from all three approaches.
-The numbers in the upper right give the Pearson correlation.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/comparator_corr.html} for an interactive version of this plot.
-\label{fig:comparator}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.7\linewidth]{figs/fixed_dist.pdf}
-\caption{
-Distribution of fitness effects of all amino-acid mutations relative to Wuhan-Hu-1, and just those mutations that fixed in at least one clade of SARS-CoV-2 (using the Nextstrain clade definitions).
-The vertical dashed line at zero indicates the effect of a neutral mutation.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts_hist.html} for an interactive version of this plot that allows adjustment of the minimum expected count threshold.
-\label{fig:fixed_dist}
-}
-\end{figure*}
-
-\begin{figure*}[b]
-\includegraphics[width=\linewidth]{figs/fixed.pdf}
-\caption{
-Effects of individual mutations that fixed in at least one clade of SARS-CoV-2, faceted by whether they are in spike or another protein.
-``Mutation polarity'' indicates if the point shows the effect of the mutation estimated using all viral clades (including those that have fixed the mutation), or just from direct forward occurrences of the mutation in clades in which it has not yet fixed.
-Some mutations are estimated to be more favorable when including clades in which they have fixed (blue circles) in addition to just clades in which it has not yet fixed (orange squares)---when this occurs, it suggests epistatic entrenchment of the mutations~\citep{shah2015contingency, starr2018pervasive}.
-Note that clades in which a mutation has already fixed contribute to estimates of its fitness via estimates of the effect of its reversion and via estimates of the effects of mutations to other amino acids at the same site.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/clade_fixed_muts.html} for an interactive version of this plot.
-\label{fig:fixed}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.75\linewidth]{figs/terminal.png}
-\caption{
-Relationship between fitness effects of mutations and two measures of the number of descendants.
-At top is shown the log ratio of counts of the mutation on non-terminal (internal) to terminal (tip) branches; larger values indicate mutations more likely to be found in viruses that leave descendants.
-At bottom is shown the mean log number of tip descendants that share all the mutations on each branch containing the mutation of interest; larger values again indicate mutations more likely to be found in viruses that leave more descendants.
-Each point is an amino-acid mutation, the orange line is a least-squares regression, and the orange text in the upper left give the number of mutations and the Pearson correlation coefficient.
-This plot shows only mutations with at least 10 expected counts and 5 actual counts.
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/fitness_vs_terminal.html} for an interactive version of this plot that allows filtering by the number of actual or expected counts, or by gene.
-The number of descendants is calculated using the ``leaves\_sharing\_mutations'' variable of the UShER mutation-annotated tree.
-\label{fig:terminal}
-}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\includegraphics[width=0.7\linewidth]{figs/nmut_counts.pdf}
-\caption{
-Number of mutations per branch on all branches of the SARS-CoV-2 tree with at least one mutation.
-The analysis used here only considers mutations on branches with four or fewer mutations and so excludes the small fraction of highly mutated branches.
-\label{fig:nmut_counts}
-}
-\end{figure*}
-
 \end{document}

### 2023-08-11 paper/paper_bioRxiv_v2.tex
manuscript formatted for final Virus Evolution submission (#42)
* made paragraph on last 5 kb of genome a bit more succinct

* add track changes manuscript

* change to author-date citations

* created separate PDF for supplemental material

* made version of `paper.tex` with refs in file and supp excluded for final submission

* make TIFF of fig1 and remove from final `paper.tex`

* make fig 2 a separate TIFF for final submission

* make fig 3 a separate EPS for final submission

* made fig 4 separate TIFF for final submission

* made fig 5 a separate EPS

* added paper with figs for early ac

--- 2023-08-08 paper/paper.tex
@@
 Most synonymous mutations are nearly neutral, most stop codons are highly deleterious, and amino-acid mutations range from slightly beneficial to highly deleterious (Figure~\ref{fig:dist}A).
-The handful of synonymous or noncoding mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
+The handful of synonymous mutations with highly deleterious effects are in either regions of known non-coding constraint (e.g., the ORF1ab ribosomal slippage site~\cite{bhatt2021structural}) or two regions in the center of E and the end of M (Figure~\ref{fig:synonymous}).
 
@@
 Observations such as these could help guide experimental studies to better understand protein function.
-
-The accessory proteins are encoded in the last 5kb of the SARS-CoV-2 genome, meaning that our results show that this portion of the genome (with the exception of the genes coding for nucleoprotein, E, and M) is under relatively weak purifying selection (Figure~\ref{fig:effects_dist_position_order}).
-Note that some accessory proteins also vary in presence/absence across coronaviruses~\citep{v2021coronavirus,llanes2020betacoronavirus}, and are mostly dispensable in cell culture~\citep{mcgrath2022sars,silvas2021contribution,liu2022live,liu2014accessory}.
-Decreased purifying selection in genes encoded near the end of the genome is also observed in more distantly related orders of large single-stranded RNA viruses such as closteroviruses~\citep{wang2011genetic, dawson2013virus}.
 
@@
 \end{itemize}
-The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors or viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
+The rationale for the first exclusion is that highly mutated branches are often indicative of sequencing errors of viral evolution in chronically infected humans, neither of which correspond to the pattern of typical SARS-CoV-2 transmission in acute infections.
 Because the virus's evolution is very densely sampled, only a small fraction of branches have more than four mutations (Figure~\ref{fig:nmut_counts}).
@@
 For the comparison to the dN/dS approaches shown in Figure~\ref{fig:dnds}, we used the dN/dS values available at \url{https://github.com/spond/SARS-CoV-2-variation}~\cite{martin2021emergence} for all SARS-CoV-2 sequences, which were calculated using the FEL approach~\cite{kosakovsky2005not}.
-Specifically, the file \url{https://raw.githubusercontent.com/spond/SARS-CoV-2-variation/master/windowed-sites-fel-all.csv} contains dN/dS estimates made using large-scale sequence sets from GISAID in three-month windows starting with the earliest sequences from December 2019 and continuing up until Jan-31-2022.
-We averaged the dN/dS values for each site over all three month windows, and analyzed those time-window averaged dN/dS values.
-In order to ensure comparability in the sequence sets used, for the comparisons in Figure~\ref{fig:dnds}, we used fitness estimates from our approach made using only sequences in the mutation-annotated tree as of Jan-31-2022.
-We restricted our analysis to this timeframe because the large-scale dN/dS analyses at \url{https://github.com/spond/SARS-CoV-2-variation} (which use state-of-the-art methods) were only available for that range of dates.
-
 The dN/dS ratios only provide a single number for each site, which cannot be directly compared to either the mutation-effect estimates or the deep mutational scanning, which estimate the effects of individual amino-acid mutations.
@@
 The dN/dS ratios are taken from \url{https://github.com/spond/SARS-CoV-2-variation}.
-Note that the y-axis in each panel is extended so that it fully contains the trend line over the entire x-axis domain, explaining why the y-axis limits vary across the rows even though the range of the data are the same on the y-axis for each row.
-Both the fitness estimates and dN/dS values shown here are calculated using only sequences available as of Jan-31-2022 (see Methods).
-See \url{https://jbloomlab.github.io/SARS2-mut-fitness/public_2022-01-31/dnds_corr.html} for an interactive version of this plot, and \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version that uses fitness estimates from all SARS-CoV-2 sequences including more recent ones.
+See \url{https://jbloomlab.github.io/SARS2-mut-fitness/dnds_corr.html} for an interactive version of this plot.
 \label{fig:dnds}
